# Algebra

```{r}
#| label: setup-algebra
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}
```

## Overview

This chapter develops six connected algebraic ideas:

1. the modulus function;
2. graphs involving modulus functions;
3. modulus inequalities;
4. polynomial division;
5. the factor theorem;
6. the remainder theorem.

The central aim is to connect symbolic algebra, graphical interpretation and numerical checking.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 1.1 The modulus function

### Key mathematical ideas

The modulus of a real number is its distance from zero on the number line.

$$
|x| =
\begin{cases}
x, & x \geq 0,\\
-x, & x < 0.
\end{cases}
$$

Therefore, $|x|$ is always non-negative.

For example,

$$
|5|=5,\qquad |-5|=5.
$$

A useful equation rule is:

$$
|A|=k,\quad k\geq 0
\quad \Longleftrightarrow \quad
A=k \ \text{or}\ A=-k.
$$

If $k<0$, then the equation $|A|=k$ has no real solution.

### Visualisation using R

```{r}
#| label: fig-modulus-basic
#| fig-cap: "The graph of y = |x|."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-5, 5, length.out = 500)
)

df$y <- abs(df$x)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_equal() +
  labs(
    title = "Graph of y = |x|",
    subtitle = "The rule changes at x = 0.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

The sharp corner at the origin occurs because the function changes from $y=-x$ to $y=x$.

### Worked example

Solve

$$
|3x-2|=10.
$$

Since $|A|=10$ means $A=10$ or $A=-10$, we solve

$$
3x-2=10
$$

or

$$
3x-2=-10.
$$

From the first equation,

$$
3x=12,
$$

so

$$
x=4.
$$

From the second equation,

$$
3x=-8,
$$

so

$$
x=-\dfrac{8}{3}.
$$

Therefore,

$$
x=4
\quad \text{or} \quad
x=-\dfrac{8}{3}.
$$

### Short investigation

Change the value of $a$ in the code below and observe the graph of $y=|x-a|$.

```{r}
#| label: fig-modulus-shift
#| fig-cap: "Exploring the family y = |x - a|."
#| fig-width: 6
#| fig-height: 4

a <- 2

df <- data.frame(
  x = seq(-5, 7, length.out = 600)
)

df$y <- abs(df$x - a)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = a, linetype = "dashed") +
  labs(
    title = paste0("Graph of y = |x - ", a, "|"),
    subtitle = "The turning point occurs at x = a.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Exercises 1.1

#### Bronze

1. Solve $|x|=7$.
2. Solve $|x|=0$.
3. Solve $|x|=-3$.
4. Solve $|2x|=8$.
5. Solve $|x+1|=5$.

#### Silver

1. Solve $|3x-1|=8$.
2. Solve $|5-2x|=9$.
3. Solve $|4x+7|=3$.
4. Solve $|6-3x|=12$.
5. Solve $|2x+5|=|x-1|$.

#### Gold

1. Solve $|3x-4|=2x+1$.
2. Solve $|x+6|=3x-2$.
3. Solve $|2x-5|=|x+4|$.
4. Find all real values of $x$ such that $|x-2|=x^2-4$.
5. Find all real values of $x$ such that $|2x+1|=x^2+x-2$.

## 1.2 Graphs involving modulus functions

### Key mathematical ideas

For a linear expression $ax+b$, the graph

$$
y=|ax+b|
$$

has a turning point where

$$
ax+b=0.
$$

Solving,

$$
x=-\dfrac{b}{a}.
$$

The graph is V-shaped because the output is reflected above the $x$-axis.

### Visualisation using R

```{r}
#| label: fig-modulus-linear
#| fig-cap: "The graph of y = |2x - 3|."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-2, 5, length.out = 600)
)

df$y <- abs(2 * df$x - 3)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 3/2, linetype = "dashed") +
  annotate("point", x = 3/2, y = 0, size = 3) +
  annotate("text", x = 3/2, y = 0.8, label = "turning point") +
  labs(
    title = "Graph of y = |2x - 3|",
    subtitle = "The turning point occurs where 2x - 3 = 0.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Sketch

$$
y=|3x+6|.
$$

The expression inside the modulus is zero when

$$
3x+6=0.
$$

Therefore,

$$
x=-2.
$$

So the turning point is

$$
(-2,0).
$$

For $x\geq -2$,

$$
|3x+6|=3x+6.
$$

For $x<-2$,

$$
|3x+6|=-(3x+6)=-3x-6.
$$

Hence the graph is made from two straight-line pieces meeting at $(-2,0)$.

### Short investigation

The following code plots several graphs of the form $y=|mx+c|$.

```{r}
#| label: fig-modulus-family
#| fig-cap: "A family of graphs y = |mx + c|."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = rep(seq(-5, 5, length.out = 600), 3),
  case = rep(c("m = 1, c = 0", "m = 2, c = -3", "m = -1, c = 2"), each = 600)
)

df$y <- ifelse(
  df$case == "m = 1, c = 0",
  abs(df$x),
  ifelse(
    df$case == "m = 2, c = -3",
    abs(2 * df$x - 3),
    abs(-df$x + 2)
  )
)

ggplot(df, aes(x = x, y = y, linetype = case)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(
    title = "Graphs of y = |mx + c|",
    subtitle = "Changing m and c changes the steepness and turning point.",
    x = "x",
    y = "y",
    linetype = "Function"
  ) +
  book_theme()
```

### Exercises 1.2

#### Bronze

1. State the turning point of $y=|x-4|$.
2. State the turning point of $y=|x+3|$.
3. State the turning point of $y=|2x-6|$.
4. State the turning point of $y=|5x+10|$.
5. State the turning point of $y=|-3x+9|$.

#### Silver

1. Sketch $y=|x-2|$.
2. Sketch $y=|2x+4|$.
3. Sketch $y=|3x-6|$.
4. Find the coordinates where $y=|x-1|$ meets $y=3$.
5. Find the coordinates where $y=|2x+1|$ meets $y=5$.

#### Gold

1. Solve graphically and algebraically: $|x-3|=2$.
2. Solve graphically and algebraically: $|2x+1|=3$.
3. Find the coordinates of intersection of $y=|x+2|$ and $y=x+4$.
4. Find the coordinates of intersection of $y=|2x-1|$ and $y=x+2$.
5. Determine the values of $k$ for which $|x-1|=k$ has exactly two distinct real solutions.

## 1.3 Modulus inequalities

### Key mathematical ideas

Modulus inequalities can often be interpreted using distance.

For example,

$$
|x-a|<r
$$

means that the distance from $x$ to $a$ is less than $r$.

So

$$
a-r<x<a+r.
$$

Similarly,

$$
|x-a|>r
$$

means that the distance from $x$ to $a$ is greater than $r$.

So

$$
x<a-r
\quad \text{or} \quad
x>a+r.
$$

### Visualisation using R

```{r}
#| label: fig-modulus-inequality
#| fig-cap: "The shaded solution region for |x - 2| < 3."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-4, 8, length.out = 600)
)

df$y <- abs(df$x - 2)
df$inside <- df$y < 3

ggplot(df, aes(x = x, y = y)) +
  geom_ribbon(
    data = subset(df, inside),
    aes(ymin = y, ymax = 3),
    alpha = 0.3
  ) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 3, linetype = "dashed") +
  geom_vline(xintercept = -1, linetype = "dashed") +
  geom_vline(xintercept = 5, linetype = "dashed") +
  annotate("text", x = 2, y = 1.4, label = "-1 < x < 5") +
  labs(
    title = "Graph of y = |x - 2| and y = 3",
    subtitle = "The shaded region shows where |x - 2| < 3.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Solve

$$
|x-5|\leq 4.
$$

This means the distance from $x$ to $5$ is at most $4$.

Therefore,

$$
-4\leq x-5\leq 4.
$$

Adding $5$ throughout gives

$$
1\leq x\leq 9.
$$

So the solution is

$$
x\in[1,9].
$$

### Short investigation

Change the value of $r$ below and observe how the solution interval changes.

```{r}
#| label: fig-modulus-inequality-family
#| fig-cap: "Exploring the shaded region for |x - 1| < r."
#| fig-width: 6
#| fig-height: 4

r <- 2.5

df <- data.frame(
  x = seq(-4, 6, length.out = 600)
)

df$y <- abs(df$x - 1)
df$inside <- df$y < r

ggplot(df, aes(x = x, y = y)) +
  geom_ribbon(
    data = subset(df, inside),
    aes(ymin = y, ymax = r),
    alpha = 0.3
  ) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = r, linetype = "dashed") +
  geom_vline(xintercept = 1 - r, linetype = "dashed") +
  geom_vline(xintercept = 1 + r, linetype = "dashed") +
  annotate(
    "text",
    x = 1,
    y = r / 2,
    label = paste0(1 - r, " < x < ", 1 + r)
  ) +
  labs(
    title = paste0("Graph of y = |x - 1| and y = ", r),
    subtitle = "The shaded region changes as r changes.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Exercises 1.3

#### Bronze

1. Solve $|x|<4$.
2. Solve $|x|>6$.
3. Solve $|x-2|<5$.
4. Solve $|x+1|\leq 3$.
5. Solve $|x-4|\geq 7$.

#### Silver

1. Solve $|2x|<10$.
2. Solve $|3x-6|\leq 9$.
3. Solve $|5-2x|<7$.
4. Solve $|4x+8|>12$.
5. Solve $|6-3x|\geq 15$.

#### Gold

1. Solve $|2x-1|<5$.
2. Solve $|3x+4|\leq 2$.
3. Solve $|x-2|<|x+4|$.
4. Solve $|2x+1|\geq |x-5|$.
5. Find all values of $k$ such that $|x-3|<k$ has no real solution.

## 1.4 Polynomial division

### Key mathematical ideas

Polynomial division is based on the division algorithm.

If $P(x)$ is divided by $D(x)$, then

$$
P(x)=D(x)Q(x)+R(x),
$$

where $Q(x)$ is the quotient and $R(x)$ is the remainder.

If the divisor is linear, then the remainder is constant.

If the divisor is quadratic, then the remainder is at most linear.

For examination work, the safest written method is **long division**. Synthetic division is a useful checking method for divisors of the form $x-a$, but it should not replace the full written method unless it is explicitly allowed.

### Worked example using long division

Divide

$$
x^3-2x^2-5x+6
$$

by

$$
x-1.
$$

We divide term by term.

First,

$$
\dfrac{x^3}{x}=x^2.
$$

Multiply $x-1$ by $x^2$:

$$
x^2(x-1)=x^3-x^2.
$$

Subtract this from the original polynomial:

$$
\begin{aligned}
x^3-2x^2-5x+6-(x^3-x^2)
&=x^3-2x^2-5x+6-x^3+x^2\\
&=-x^2-5x+6.
\end{aligned}
$$

Next,

$$
\dfrac{-x^2}{x}=-x.
$$

Multiply $x-1$ by $-x$:

$$
-x(x-1)=-x^2+x.
$$

Subtract:

$$
\begin{aligned}
-x^2-5x+6-(-x^2+x)
&=-x^2-5x+6+x^2-x\\
&=-6x+6.
\end{aligned}
$$

Finally,

$$
\dfrac{-6x}{x}=-6.
$$

Multiply $x-1$ by $-6$:

$$
-6(x-1)=-6x+6.
$$

Subtract:

$$
(-6x+6)-(-6x+6)=0.
$$

Therefore,

$$
x^3-2x^2-5x+6=(x-1)(x^2-x-6).
$$

So the quotient is

$$
x^2-x-6
$$

and the remainder is

$$
0.
$$

### Synthetic division check

Synthetic division gives a quick check when the divisor has the form $x-a$.

Here, the divisor is $x-1$, so $a=1$.

The coefficients of

$$
x^3-2x^2-5x+6
$$

are

$$
1,\ -2,\ -5,\ 6.
$$

The synthetic division check is

$$
\begin{array}{c|rrrr}
1 & 1 & -2 & -5 & 6\\
  &   & 1 & -1 & -6\\
\hline
  & 1 & -1 & -6 & 0
\end{array}
$$

This gives quotient coefficients

$$
1,\ -1,\ -6,
$$

so the quotient is

$$
x^2-x-6.
$$

The final number is $0$, so the remainder is $0$.

This agrees with the long division result.

::: {.callout-warning}
Synthetic division is included here as a checking method. For examination working, present the long division method unless the examination explicitly allows synthetic division.
:::

### Visualisation using R

```{r}
#| label: fig-polynomial-division-root
#| fig-cap: "The graph of f(x) = x^3 - 2x^2 - 5x + 6."
#| fig-width: 6
#| fig-height: 4

f <- function(x) x^3 - 2*x^2 - 5*x + 6

df <- data.frame(
  x = seq(-3.5, 4, length.out = 800)
)

df$y <- f(df$x)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("point", x = 1, y = 0, size = 3) +
  annotate("text", x = 1.2, y = 6, label = "x = 1") +
  labs(
    title = expression(f(x) == x^3 - 2*x^2 - 5*x + 6),
    subtitle = "Since f(1) = 0, division by x - 1 leaves remainder 0.",
    x = "x",
    y = "f(x)"
  ) +
  book_theme()
```

### Short investigation

Use R to check the identity

$$
P(x)=D(x)Q(x)+R(x).
$$

```{r}
P <- function(x) x^3 - 2*x^2 - 5*x + 6
D <- function(x) x - 1
Q <- function(x) x^2 - x - 6
R <- function(x) 0

test_values <- c(-2, 0, 1, 3)

data.frame(
  x = test_values,
  P_x = P(test_values),
  DQ_plus_R = D(test_values) * Q(test_values) + R(test_values)
)
```

### Synthetic division in R

The following code performs the synthetic division check.

```{r}
root <- 1
coefficients <- c(1, -2, -5, 6)

bottom_row <- numeric(length(coefficients))
bottom_row[1] <- coefficients[1]

for (i in 2:length(coefficients)) {
  bottom_row[i] <- coefficients[i] + root * bottom_row[i - 1]
}

data.frame(
  coefficient = coefficients,
  synthetic_bottom_row = bottom_row
)
```

The last entry is the remainder. The previous entries are the quotient coefficients.

### Exercises 1.4

#### Bronze

1. Divide $x^2+5x+6$ by $x+2$.
2. Divide $x^2-9$ by $x-3$.
3. Divide $x^2-4x+3$ by $x-1$.
4. Divide $x^3-1$ by $x-1$.
5. Divide $x^3+8$ by $x+2$.

#### Silver

1. Divide $x^3+2x^2-x-2$ by $x+2$.
2. Divide $x^3-4x^2+x+6$ by $x-2$.
3. Divide $2x^3+3x^2-8x+3$ by $x+3$.
4. Divide $x^4-1$ by $x-1$.
5. Divide $x^3+3x^2+3x+1$ by $x+1$.

#### Gold

1. Divide $x^4-3x^3+2x^2+x-1$ by $x-1$.
2. Divide $2x^4-x^3-7x^2+5x+3$ by $x+1$.
3. Divide $x^4+2x^3-3x^2-4x+4$ by $x^2-1$.
4. Divide $2x^4+3x^3-x^2+5x-2$ by $x^2+x-1$.
5. Given that $P(x)=x^3+ax^2+bx+6$ is divisible by $x-2$, find a relation between $a$ and $b$.

## 1.5 The factor theorem

### Key mathematical ideas

The factor theorem states:

If

$$
P(a)=0,
$$

then

$$
x-a
$$

is a factor of $P(x)$.

Conversely, if $x-a$ is a factor of $P(x)$, then

$$
P(a)=0.
$$

### Worked example

Show that $x-2$ is a factor of

$$
P(x)=x^3-5x^2+2x+8.
$$

By the factor theorem, $x-2$ is a factor if $P(2)=0$.

Now,

$$
P(2)=2^3-5(2)^2+2(2)+8.
$$

So

$$
P(2)=8-20+4+8=0.
$$

Therefore,

$$
x-2
$$

is a factor of $P(x)$.

To factorise fully, divide $P(x)$ by $x-2$.

Using long division,

$$
x^3-5x^2+2x+8=(x-2)(x^2-3x-4).
$$

Then

$$
x^2-3x-4=(x-4)(x+1).
$$

Therefore,

$$
P(x)=(x-2)(x-4)(x+1).
$$

### Synthetic division check

For the divisor $x-2$, we use $a=2$.

The coefficients of

$$
x^3-5x^2+2x+8
$$

are

$$
1,\ -5,\ 2,\ 8.
$$

The synthetic division check is

$$
\begin{array}{c|rrrr}
2 & 1 & -5 & 2 & 8\\
  &   & 2 & -6 & -8\\
\hline
  & 1 & -3 & -4 & 0
\end{array}
$$

This confirms that the quotient is

$$
x^2-3x-4
$$

and the remainder is $0$.

### Visualisation using R

```{r}
#| label: fig-factor-theorem
#| fig-cap: "The graph of P(x) = x^3 - 5x^2 + 2x + 8."
#| fig-width: 6
#| fig-height: 4

P <- function(x) x^3 - 5*x^2 + 2*x + 8

df <- data.frame(
  x = seq(-2, 5, length.out = 800)
)

df$y <- P(df$x)

roots <- data.frame(
  x = c(-1, 2, 4),
  y = c(0, 0, 0)
)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(data = roots, aes(x = x, y = y), size = 3) +
  labs(
    title = expression(P(x) == x^3 - 5*x^2 + 2*x + 8),
    subtitle = "The x-intercepts correspond to linear factors.",
    x = "x",
    y = "P(x)"
  ) +
  book_theme()
```

### Short investigation

Change the roots $a$, $b$ and $c$ below.

```{r}
#| label: fig-cubic-roots-family
#| fig-cap: "Exploring f(x) = (x - a)(x - b)(x - c)."
#| fig-width: 6
#| fig-height: 4

a <- -2
b <- 1
c <- 3

f <- function(x) (x - a) * (x - b) * (x - c)

df <- data.frame(
  x = seq(-5, 5, length.out = 800)
)

df$y <- f(df$x)

root_df <- data.frame(
  x = c(a, b, c),
  y = c(0, 0, 0)
)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(data = root_df, aes(x = x, y = y), size = 3) +
  labs(
    title = "Graph of f(x) = (x - a)(x - b)(x - c)",
    subtitle = "The chosen values of a, b and c become the roots.",
    x = "x",
    y = "f(x)"
  ) +
  book_theme()
```

### Exercises 1.5

#### Bronze

1. Show that $x-1$ is a factor of $x^2-3x+2$.
2. Show that $x+2$ is a factor of $x^2+x-2$.
3. Show that $x-3$ is a factor of $x^2-5x+6$.
4. Show that $x+1$ is a factor of $x^3+1$.
5. Show that $x-2$ is a factor of $x^3-8$.

#### Silver

1. Show that $x-1$ is a factor of $x^3-4x^2+x+2$.
2. Show that $x+2$ is a factor of $x^3+3x^2-4x-12$.
3. Factorise $x^3-6x^2+11x-6$ fully.
4. Factorise $x^3+x^2-4x-4$ fully.
5. Factorise $2x^3-3x^2-8x+12$ fully.

#### Gold

1. Find $k$ if $x-2$ is a factor of $x^3+kx^2-4x+4$.
2. Find $k$ if $x+1$ is a factor of $2x^3+kx^2-5x+6$.
3. Given that $x-3$ is a factor of $x^3+ax^2+bx-9$, find a relation between $a$ and $b$.
4. The polynomial $x^3+px^2+qx-6$ has roots $1$, $2$ and $r$. Find $p$, $q$ and $r$.
5. The polynomial $2x^3+ax^2+bx-6$ has $x-1$ and $x+2$ as factors. Find $a$ and $b$.

## 1.6 The remainder theorem

### Key mathematical ideas

The remainder theorem states:

When $P(x)$ is divided by $x-a$, the remainder is

$$
P(a).
$$

More generally, when $P(x)$ is divided by $ax-b$, the remainder is

$$
P\left(\dfrac{b}{a}\right).
$$

This is connected to polynomial division because if

$$
P(x)=(x-a)Q(x)+R,
$$

then substituting $x=a$ gives

$$
P(a)=(a-a)Q(a)+R.
$$

Therefore,

$$
P(a)=R.
$$

### Worked example

Find the remainder when

$$
P(x)=x^3-2x^2-5x+6
$$

is divided by $x-2$.

By the remainder theorem, the remainder is $P(2)$.

Now,

$$
P(2)=2^3-2(2)^2-5(2)+6.
$$

So

$$
P(2)=8-8-10+6=-4.
$$

Therefore, the remainder is

$$
-4.
$$

### Long division check

Dividing

$$
x^3-2x^2-5x+6
$$

by

$$
x-2
$$

using long division gives:

First,

$$
\dfrac{x^3}{x}=x^2.
$$

Subtract

$$
x^2(x-2)=x^3-2x^2.
$$

Then

$$
x^3-2x^2-5x+6-(x^3-2x^2)=-5x+6.
$$

Next,

$$
\dfrac{-5x}{x}=-5.
$$

Subtract

$$
-5(x-2)=-5x+10.
$$

Then

$$
(-5x+6)-(-5x+10)=-4.
$$

Therefore,

$$
x^3-2x^2-5x+6=(x-2)(x^2-5)-4.
$$

The remainder is again

$$
-4.
$$

### Synthetic division check

For the divisor $x-2$, we use $a=2$.

The coefficients are

$$
1,\ -2,\ -5,\ 6.
$$

The synthetic division check is

$$
\begin{array}{c|rrrr}
2 & 1 & -2 & -5 & 6\\
  &   & 2 & 0 & -10\\
\hline
  & 1 & 0 & -5 & -4
\end{array}
$$

This gives quotient

$$
x^2-5
$$

and remainder

$$
-4.
$$

### Visualisation using R

```{r}
#| label: fig-remainder-theorem
#| fig-cap: "The value P(2) gives the remainder when P(x) is divided by x - 2."
#| fig-width: 6
#| fig-height: 4

P <- function(x) x^3 - 2*x^2 - 5*x + 6

df <- data.frame(
  x = seq(-3.5, 4, length.out = 800)
)

df$y <- P(df$x)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 2, linetype = "dashed") +
  annotate("point", x = 2, y = P(2), size = 3) +
  annotate("text", x = 2.45, y = P(2), label = "P(2) = -4") +
  labs(
    title = expression(P(x) == x^3 - 2*x^2 - 5*x + 6),
    subtitle = "The y-value at x = 2 gives the remainder.",
    x = "x",
    y = "P(x)"
  ) +
  book_theme()
```

### Short investigation

Evaluate the same polynomial at different values of $a$.

```{r}
P <- function(x) x^3 - 2*x^2 - 5*x + 6

a_values <- c(-2, -1, 0, 1, 2, 3)

data.frame(
  divisor = paste0("x - (", a_values, ")"),
  a = a_values,
  remainder = P(a_values)
)
```

### Exercises 1.6

#### Bronze

1. Find the remainder when $x^2+3x+1$ is divided by $x-1$.
2. Find the remainder when $x^2-5x+6$ is divided by $x-2$.
3. Find the remainder when $x^3+1$ is divided by $x+1$.
4. Find the remainder when $x^3-8$ is divided by $x-2$.
5. Find the remainder when $2x^2+x-4$ is divided by $x+2$.

#### Silver

1. Find the remainder when $x^3-4x^2+x+6$ is divided by $x-3$.
2. Find the remainder when $2x^3+x^2-5x+1$ is divided by $x+1$.
3. Find the remainder when $x^4-3x^2+2$ is divided by $x-2$.
4. Find the remainder when $3x^3-2x+5$ is divided by $2x-1$.
5. Find the remainder when $x^3+2x^2-x+4$ is divided by $3x+1$.

#### Gold

1. Find $k$ if $x-2$ leaves remainder $5$ when it divides $x^3+kx^2-3x+1$.
2. Find $k$ if $x+1$ leaves remainder $4$ when it divides $2x^3+kx^2-x+5$.
3. The polynomial $P(x)=x^3+ax^2+bx+2$ leaves remainder $4$ when divided by $x-1$ and remainder $-10$ when divided by $x+2$. Find two equations in $a$ and $b$.
4. Find the remainder when $P(x)=x^{50}+x^{25}+1$ is divided by $x-1$.
5. Find the remainder when $P(x)=2x^3-5x^2+7x-3$ is divided by $2x+3$.

## STEP-style extension questions

These questions are designed to require more than routine application.

### STEP-style Question 1

Let

$$
f(x)=|x-a|+|x-b|,
$$

where $a<b$.

Show that the minimum value of $f(x)$ is $b-a$.

### STEP-style Question 2

Find all real numbers $x$ satisfying

$$
|x-1|+|x+2|=7.
$$

### STEP-style Question 3

Let

$$
P(x)=x^3+ax^2+bx+c.
$$

Suppose $P(1)=P(2)=P(3)=0$.

Find $P(x)$ in factorised form and express $a$, $b$ and $c$.

### STEP-style Question 4

Let

$$
P(x)=x^4+ax^3+bx^2+cx+d.
$$

Suppose that $P(x)$ is divisible by both $x-1$ and $x+1$.

Find two equations involving $a$, $b$, $c$ and $d$.

### STEP-style Question 5

Let

$$
P(x)=x^3+px+q.
$$

Suppose that the remainder when $P(x)$ is divided by $x-1$ is $4$, and the remainder when $P(x)$ is divided by $x+2$ is $-5$.

Find $p$ and $q$.

## Reflection questions

1. What is the difference between solving $|A|=k$ and solving $A=k$?
2. Why do modulus equations sometimes require checking?
3. How does the graph of $y=|ax+b|$ reveal the solution structure?
4. Why is long division the safest written method for polynomial division?
5. How can synthetic division be used as a checking method?
6. How is polynomial division connected to numerical substitution?
7. What is the difference between the factor theorem and the remainder theorem?
8. What can R show clearly that algebra alone may hide?
9. What can algebra prove exactly that a graph can only suggest?

## Answers

### Answers to Exercises 1.1

#### Bronze

1. $x=\pm 7$.
2. $x=0$.
3. No real solution.
4. $x=\pm 4$.
5. $x=4$ or $x=-6$.

#### Silver

1. $x=3$ or $x=-\dfrac{7}{3}$.
2. $x=7$ or $x=-2$.
3. $x=-1$ or $x=-\dfrac{5}{2}$.
4. $x=6$ or $x=-2$.
5. $x=-6$ or $x=-\dfrac{4}{3}$.

#### Gold

1. $x=5$ or $x=\dfrac{3}{5}$.
2. $x=4$.
3. $x=9$ or $x=\dfrac{1}{3}$.
4. $x=-3$ or $x=2$.
5. $x=\dfrac{1+\sqrt{13}}{2}$ or $x=\dfrac{-3-\sqrt{13}}{2}$.

### Answers to Exercises 1.2

#### Bronze

1. $(4,0)$.
2. $(-3,0)$.
3. $(3,0)$.
4. $(-2,0)$.
5. $(3,0)$.

#### Silver

1. Turning point $(2,0)$.
2. Turning point $(-2,0)$.
3. Turning point $(2,0)$.
4. $(-2,3)$ and $(4,3)$.
5. $(-3,5)$ and $(2,5)$.

#### Gold

1. $x=1$ or $x=5$.
2. $x=1$ or $x=-2$.
3. $(-3,1)$.
4. $\left(-\dfrac{1}{3},\dfrac{5}{3}\right)$ and $(3,5)$.
5. $k>0$.

### Answers to Exercises 1.3

#### Bronze

1. $-4<x<4$.
2. $x<-6$ or $x>6$.
3. $-3<x<7$.
4. $-4\leq x\leq 2$.
5. $x\leq -3$ or $x\geq 11$.

#### Silver

1. $-5<x<5$.
2. $-1\leq x\leq 5$.
3. $-1<x<6$.
4. $x<-5$ or $x>1$.
5. $x\leq -3$ or $x\geq 7$.

#### Gold

1. $-2<x<3$.
2. $-2\leq x\leq -\dfrac{2}{3}$.
3. $x>-1$.
4. $x\leq -6$ or $x\geq \dfrac{4}{3}$.
5. $k\leq 0$.

### Answers to Exercises 1.4

#### Bronze

1. Quotient $x+3$, remainder $0$.
2. Quotient $x+3$, remainder $0$.
3. Quotient $x-3$, remainder $0$.
4. Quotient $x^2+x+1$, remainder $0$.
5. Quotient $x^2-2x+4$, remainder $0$.

#### Silver

1. Quotient $x^2-1$, remainder $0$.
2. Quotient $x^2-2x-3$, remainder $0$.
3. Quotient $2x^2-3x+1$, remainder $0$.
4. Quotient $x^3+x^2+x+1$, remainder $0$.
5. Quotient $x^2+2x+1$, remainder $0$.

#### Gold

1. Quotient $x^3-2x^2+1$, remainder $0$.
2. Quotient $2x^3-3x^2-4x+9$, remainder $-6$.
3. Quotient $x^2+2x-2$, remainder $-2x+2$.
4. Quotient $2x^2+x$, remainder $6x-2$.
5. $2a+b+7=0$.

### Answers to Exercises 1.5

#### Bronze

1. $f(1)=0$, so $x-1$ is a factor.
2. $f(-2)=0$, so $x+2$ is a factor.
3. $f(3)=0$, so $x-3$ is a factor.
4. $f(-1)=0$, so $x+1$ is a factor.
5. $f(2)=0$, so $x-2$ is a factor.

#### Silver

1. $f(1)=0$, so $x-1$ is a factor.
2. $f(-2)=0$, so $x+2$ is a factor.
3. $(x-1)(x-2)(x-3)$.
4. $(x+1)(x-2)(x+2)$.
5. $(x-2)(x+2)(2x-3)$.

#### Gold

1. $k=-1$.
2. $k=-9$.
3. $3a+b+6=0$.
4. $r=3$, $p=-6$, $q=11$.
5. $a=5$, $b=-1$.

### Answers to Exercises 1.6

#### Bronze

1. $5$.
2. $0$.
3. $0$.
4. $0$.
5. $2$.

#### Silver

1. $0$.
2. $5$.
3. $6$.
4. $\dfrac{35}{8}$.
5. $\dfrac{122}{27}$.

#### Gold

1. $k=\dfrac{1}{2}$.
2. $k=0$.
3. $a+b=1$ and $4a-2b=-4$.
4. $3$.
5. $-\dfrac{63}{2}$.

### Answers to STEP-style extension questions

1. For $x\in[a,b]$, $|x-a|+|x-b|=(x-a)+(b-x)=b-a$. Outside this interval the sum is larger. Minimum value: $b-a$.
2. $x=-4$ or $x=3$.
3. $P(x)=(x-1)(x-2)(x-3)=x^3-6x^2+11x-6$, so $a=-6$, $b=11$, $c=-6$.
4. Since $P(1)=0$ and $P(-1)=0$, we get $1+a+b+c+d=0$ and $1-a+b-c+d=0$.
5. $p=0$ and $q=3$.