# Differentiation

```{r}
#| label: setup-differentiation
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

numeric_derivative <- function(f, x, h = 1e-5) {
  (f(x + h) - f(x - h)) / (2 * h)
}
```

## Overview

This chapter develops seven connected differentiation ideas:

1. the product rule;
2. the quotient rule;
3. derivatives of exponential functions;
4. derivatives of natural logarithmic functions;
5. derivatives of trigonometric functions;
6. implicit differentiation;
7. parametric differentiation.

The central aim is to extend differentiation beyond simple powers of $x$. We now differentiate products, quotients, exponential functions, logarithmic functions, trigonometric functions, implicit curves and parametric curves.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 4.1 The product rule

### Key mathematical ideas

Suppose

$$
y=uv,
$$

where $u$ and $v$ are both functions of $x$.

Then the product rule is

$$
\dfrac{dy}{dx}
=
u\dfrac{dv}{dx}
+
v\dfrac{du}{dx}.
$$

Equivalently,

$$
\dfrac{d}{dx}(uv)=u'v+uv'.
$$

The product rule is needed when a function is naturally written as the product of two differentiable functions.

### Visualisation using R

```{r}
#| label: fig-product-rule-gradient
#| fig-cap: "Comparing the numerical and exact derivative of y = x^2 e^x."
#| fig-width: 6
#| fig-height: 4

f <- function(x) x^2 * exp(x)

df <- data.frame(
  x = seq(-3, 2, length.out = 600)
)

df$y <- f(df$x)
df$exact_derivative <- exp(df$x) * (df$x^2 + 2 * df$x)
df$numeric_derivative <- numeric_derivative(f, df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = exact_derivative, linetype = "exact derivative"), linewidth = 1.1) +
  geom_line(aes(y = numeric_derivative, linetype = "numerical derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Derivative of y = x^2 e^x",
    subtitle = "The product rule agrees with numerical gradient estimates.",
    x = "x",
    y = "gradient",
    linetype = "Method"
  ) +
  book_theme()
```

### Worked example

Differentiate

$$
y=x^2e^x.
$$

Let

$$
u=x^2
$$

and

$$
v=e^x.
$$

Then

$$
\dfrac{du}{dx}=2x
$$

and

$$
\dfrac{dv}{dx}=e^x.
$$

Using the product rule,

$$
\dfrac{dy}{dx}
=
x^2e^x+e^x(2x).
$$

Therefore,

$$
\dfrac{dy}{dx}
=
e^x(x^2+2x).
$$

### Short investigation

Change the power of $x$ in the code below.

```{r}
#| label: fig-product-rule-family
#| fig-cap: "Exploring derivatives of y = x^n e^x."
#| fig-width: 6
#| fig-height: 4

n <- 3

df <- data.frame(
  x = seq(-3, 3, length.out = 600)
)

df$y <- df$x^n * exp(df$x)
df$dy <- exp(df$x) * (n * df$x^(n - 1) + df$x^n)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = paste0("Function and derivative for y = x^", n, " e^x"),
    subtitle = "The product rule creates two terms.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 4.1

#### Bronze

1. Differentiate $y=x(x+3)$.
2. Differentiate $y=x^2(x-1)$.
3. Differentiate $y=x^3(2x+1)$.
4. Differentiate $y=(x+1)(x^2+2)$.
5. Differentiate $y=(2x-1)(3x+4)$.

#### Silver

1. Differentiate $y=x^2e^x$.
2. Differentiate $y=(x+1)e^{2x}$.
3. Differentiate $y=x^3\ln x$, where $x>0$.
4. Differentiate $y=(2x-1)\sin x$.
5. Differentiate $y=x^2\cos3x$.

#### Gold

1. Find the stationary point of $y=xe^{-x}$.
2. Find the equation of the tangent to $y=x^2\ln x$ at $x=1$.
3. Show that $y=(x^2+1)e^{-2x}$ has no stationary points.
4. Show that if $y=x^ne^x$, where $n$ is a positive integer, then
   $$
   \dfrac{dy}{dx}=x^{n-1}e^x(x+n).
   $$
5. Find the $x$-coordinates of the stationary points of $y=(x^2-4)e^x$.

## 4.2 The quotient rule

### Key mathematical ideas

Suppose

$$
y=\dfrac{u}{v},
$$

where $u$ and $v$ are differentiable functions of $x$ and $v\neq0$.

Then the quotient rule is

$$
\dfrac{dy}{dx}
=
\dfrac{v\dfrac{du}{dx}-u\dfrac{dv}{dx}}{v^2}.
$$

Equivalently,

$$
\dfrac{d}{dx}\left(\dfrac{u}{v}\right)
=
\dfrac{u'v-uv'}{v^2}.
$$

A common error is to differentiate the numerator and denominator separately. In general,

$$
\dfrac{d}{dx}\left(\dfrac{u}{v}\right)
\neq
\dfrac{u'}{v'}.
$$

### Visualisation using R

```{r}
#| label: fig-quotient-rule-gradient
#| fig-cap: "The function y = x / (x^2 + 1) and its derivative."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-5, 5, length.out = 800)
)

df$y <- df$x / (df$x^2 + 1)
df$dy <- (1 - df$x^2) / (df$x^2 + 1)^2

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Function and derivative",
    subtitle = "For y = x / (x^2 + 1), the derivative is zero at x = -1 and x = 1.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Differentiate

$$
y=\dfrac{x^2+1}{x-1}.
$$

Let

$$
u=x^2+1
$$

and

$$
v=x-1.
$$

Then

$$
\dfrac{du}{dx}=2x
$$

and

$$
\dfrac{dv}{dx}=1.
$$

Using the quotient rule,

$$
\dfrac{dy}{dx}
=
\dfrac{(x-1)(2x)-(x^2+1)(1)}{(x-1)^2}.
$$

Simplifying the numerator,

$$
(x-1)(2x)-(x^2+1)
=
2x^2-2x-x^2-1
=
x^2-2x-1.
$$

Therefore,

$$
\dfrac{dy}{dx}
=
\dfrac{x^2-2x-1}{(x-1)^2}.
$$

### Short investigation

Change the numerator and denominator and observe how the derivative changes.

```{r}
#| label: fig-quotient-stationary
#| fig-cap: "Stationary points of y = (x^2 + 1)/(x - 1)."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-4, 6, length.out = 1200)
)

df <- subset(df, abs(x - 1) > 0.05)

df$y <- (df$x^2 + 1) / (df$x - 1)
df$dy <- (df$x^2 - 2 * df$x - 1) / (df$x - 1)^2

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  coord_cartesian(ylim = c(-10, 10)) +
  labs(
    title = "Quotient rule and stationary points",
    subtitle = "Stationary points occur where the derivative is zero.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 4.2

#### Bronze

1. Differentiate $y=\dfrac{x+1}{x-2}$.
2. Differentiate $y=\dfrac{2x-3}{x+4}$.
3. Differentiate $y=\dfrac{x^2}{x+1}$.
4. Differentiate $y=\dfrac{x^2+1}{x}$.
5. Differentiate $y=\dfrac{1}{x^2+1}$.

#### Silver

1. Differentiate $y=\dfrac{x^2+3x}{x-1}$.
2. Differentiate $y=\dfrac{e^x}{x}$.
3. Differentiate $y=\dfrac{\ln x}{x}$, where $x>0$.
4. Find the gradient of $y=\dfrac{x+2}{x^2+1}$ at $x=1$.
5. Find the $x$-coordinates of the stationary points of $y=\dfrac{x^2+1}{x-1}$.

#### Gold

1. Find the equation of the tangent to $y=\dfrac{x^2-4}{x+1}$ at $x=0$.
2. Find the equation of the normal to $y=\dfrac{2x+1}{x-3}$ at $x=1$.
3. Find and classify the stationary points of $y=\dfrac{x}{x^2+1}$.
4. Show that
   $$
   \dfrac{d}{dx}\left(\dfrac{ax+b}{cx+d}\right)
   =
   \dfrac{ad-bc}{(cx+d)^2}.
   $$
5. Given that $y=\dfrac{x^2+ax+1}{x+1}$ has a stationary point when $x=2$, find $a$.

## 4.3 Derivatives of exponential functions

### Key mathematical ideas

The natural exponential function has the special property

$$
\dfrac{d}{dx}\left(e^x\right)=e^x.
$$

Using the chain rule,

$$
\dfrac{d}{dx}\left(e^{f(x)}\right)=f'(x)e^{f(x)}.
$$

In particular,

$$
\dfrac{d}{dx}\left(e^{ax+b}\right)=ae^{ax+b}.
$$

### Visualisation using R

```{r}
#| label: fig-exp-derivative
#| fig-cap: "The function y = e^x is its own derivative."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-3, 3, length.out = 600)
)

df$y <- exp(df$x)
df$dy <- exp(df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  labs(
    title = "The derivative of e^x",
    subtitle = "The function and its derivative are identical.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Differentiate

$$
y=e^{3x-2}.
$$

Let

$$
u=3x-2.
$$

Then

$$
\dfrac{du}{dx}=3.
$$

Using the chain rule,

$$
\dfrac{dy}{dx}=3e^{3x-2}.
$$

### Short investigation

Change the value of $a$ in $y=e^{ax}$.

```{r}
#| label: fig-exp-chain-family
#| fig-cap: "Exploring y = e^(ax) and its derivative."
#| fig-width: 6
#| fig-height: 4

a <- 2

df <- data.frame(
  x = seq(-2, 2, length.out = 600)
)

df$y <- exp(a * df$x)
df$dy <- a * exp(a * df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  labs(
    title = paste0("Function and derivative for y = e^(", a, "x)"),
    subtitle = "The chain rule multiplies by the coefficient of x.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 4.3

#### Bronze

1. Differentiate $y=e^x$.
2. Differentiate $y=e^{3x}$.
3. Differentiate $y=5e^{2x}$.
4. Differentiate $y=e^{-x}$.
5. Differentiate $y=e^{1-4x}$.

#### Silver

1. Differentiate $y=xe^x$.
2. Differentiate $y=x^2e^{3x}$.
3. Solve $\dfrac{d}{dx}\left(e^{2x-1}\right)=10$.
4. Find the stationary point of $y=(x-2)e^x$.
5. Find the equation of the tangent to $y=e^{2x}$ at $x=0$.

#### Gold

1. Find the $x$-coordinates of the stationary points of $y=(x^2-3x)e^x$.
2. Find the stationary point of $y=\dfrac{e^x}{x+1}$.
3. Show that the tangent to $y=e^{-x}$ at $x=0$ meets the $x$-axis at $x=1$.
4. Solve $e^{2x}-3e^x+2=0$.
5. Find the minimum value of $y=e^x+e^{-x}$.

## 4.4 Derivatives of natural logarithmic functions

### Key mathematical ideas

The derivative of the natural logarithm is

$$
\dfrac{d}{dx}(\ln x)=\dfrac{1}{x},
\quad x>0.
$$

Using the chain rule,

$$
\dfrac{d}{dx}\left(\ln f(x)\right)
=
\dfrac{f'(x)}{f(x)}.
$$

In particular,

$$
\dfrac{d}{dx}\left(\ln(ax+b)\right)
=
\dfrac{a}{ax+b}.
$$

The argument of a logarithm must be positive.

### Visualisation using R

```{r}
#| label: fig-log-derivative
#| fig-cap: "The derivative of y = ln(x) is y = 1/x."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(0.1, 6, length.out = 600)
)

df$y <- log(df$x)
df$dy <- 1 / df$x

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "ln(x)"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "1/x"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Natural logarithm and its derivative",
    subtitle = "The derivative decreases as x increases.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Differentiate

$$
y=\ln(2x+5).
$$

Let

$$
f(x)=2x+5.
$$

Then

$$
f'(x)=2.
$$

Therefore,

$$
\dfrac{dy}{dx}
=
\dfrac{2}{2x+5}.
$$

### Short investigation

Change the linear expression inside the logarithm.

```{r}
#| label: fig-ln-linear-family
#| fig-cap: "Exploring derivatives of y = ln(ax + b)."
#| fig-width: 6
#| fig-height: 4

a <- 2
b <- 3

df <- data.frame(
  x = seq(0.1, 6, length.out = 600)
)

df$y <- log(a * df$x + b)
df$dy <- a / (a * df$x + b)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = paste0("Function and derivative for y = ln(", a, "x + ", b, ")"),
    subtitle = "The derivative has the form a / (ax + b).",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 4.4

#### Bronze

1. Differentiate $y=\ln x$.
2. Differentiate $y=\ln(3x)$.
3. Differentiate $y=\ln(x+2)$.
4. Differentiate $y=5\ln x$.
5. Differentiate $y=\ln(x^2)$.

#### Silver

1. Differentiate $y=\ln(2x-1)$.
2. Differentiate $y=x\ln x$, where $x>0$.
3. Differentiate $y=\ln(x^2+1)$.
4. Find the stationary point of $y=\dfrac{\ln x}{x}$, where $x>0$.
5. Find the equation of the tangent to $y=\ln(x+1)$ at $x=0$.

#### Gold

1. Find the stationary point of $y=x^2\ln x$, where $x>0$.
2. Find the stationary point of $y=\dfrac{\ln x}{x^2}$, where $x>0$.
3. Show that
   $$
   \dfrac{d}{dx}\left(\ln\sqrt{x^2+1}\right)
   =
   \dfrac{x}{x^2+1}.
   $$
4. Solve
   $$
   \dfrac{d}{dx}\left(\ln(x^2-4x+5)\right)=0.
   $$
5. Find the equation of the tangent to $y=\ln(2x+1)$ that is parallel to $y=2x+3$.

## 4.5 Derivatives of trigonometric functions

### Key mathematical ideas

For trigonometric differentiation, angles must be measured in radians.

The key derivatives are

$$
\dfrac{d}{dx}(\sin x)=\cos x,
$$

$$
\dfrac{d}{dx}(\cos x)=-\sin x,
$$

and

$$
\dfrac{d}{dx}(\tan x)=\sec^2x.
$$

Using the chain rule,

$$
\dfrac{d}{dx}\left(\sin(ax+b)\right)=a\cos(ax+b),
$$

$$
\dfrac{d}{dx}\left(\cos(ax+b)\right)=-a\sin(ax+b),
$$

and

$$
\dfrac{d}{dx}\left(\tan(ax+b)\right)=a\sec^2(ax+b).
$$

### Visualisation using R

```{r}
#| label: fig-sin-derivative
#| fig-cap: "The derivative of y = sin(x) is y = cos(x)."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-2 * pi, 2 * pi, length.out = 800)
)

df$y <- sin(df$x)
df$dy <- cos(df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "sin(x)"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "cos(x)"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "The derivative of sin(x)",
    subtitle = "Angles are measured in radians.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Differentiate

$$
y=\sin(3x+1).
$$

Using the chain rule,

$$
\dfrac{dy}{dx}=3\cos(3x+1).
$$

### Short investigation

Compare a trigonometric function with its derivative.

```{r}
#| label: fig-trig-derivative-family
#| fig-cap: "Function and derivative for y = cos(2x)."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-2 * pi, 2 * pi, length.out = 800)
)

df$y <- cos(2 * df$x)
df$dy <- -2 * sin(2 * df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y, linetype = "cos(2x)"), linewidth = 1.1) +
  geom_line(aes(y = dy, linetype = "derivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Derivative of y = cos(2x)",
    subtitle = "The chain rule changes the amplitude of the derivative.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 4.5

#### Bronze

1. Differentiate $y=\sin x$.
2. Differentiate $y=\cos x$.
3. Differentiate $y=\tan x$.
4. Differentiate $y=\sin2x$.
5. Differentiate $y=\cos(3x+1)$.

#### Silver

1. Differentiate $y=x\sin x$.
2. Differentiate $y=e^x\sin x$.
3. Differentiate $y=\sin^2x$.
4. Differentiate $y=\tan2x$.
5. Find the equation of the tangent to $y=\sin x$ at $x=\dfrac{\pi}{6}$.

#### Gold

1. Find and classify the stationary points of $y=\sin x+\cos x$ for $0\leq x<2\pi$.
2. Find the stationary points of $y=\sin x\cos x$ for $0\leq x\leq\pi$.
3. Solve $\dfrac{d}{dx}(x+2\cos x)=0$ for $0\leq x\leq2\pi$.
4. Prove that
   $$
   \dfrac{d}{dx}(\sec x)=\sec x\tan x.
   $$
5. Find the equation of the tangent to $y=\tan x$ at $x=\dfrac{\pi}{4}$.

## 4.6 Implicit differentiation

### Key mathematical ideas

Implicit differentiation is used when a curve is not written directly in the form

$$
y=f(x).
$$

For example,

$$
x^2+y^2=25
$$

defines a circle. We differentiate both sides with respect to $x$, remembering that $y$ is a function of $x$.

Since

$$
\dfrac{d}{dx}(y^2)=2y\dfrac{dy}{dx},
$$

we get

$$
2x+2y\dfrac{dy}{dx}=0.
$$

Therefore,

$$
\dfrac{dy}{dx}=-\dfrac{x}{y}.
$$

### Visualisation using R

```{r}
#| label: fig-implicit-circle
#| fig-cap: "The circle x^2 + y^2 = 25 and the tangent at (3, 4)."
#| fig-width: 6
#| fig-height: 5

theta <- seq(0, 2 * pi, length.out = 800)

circle <- data.frame(
  x = 5 * cos(theta),
  y = 5 * sin(theta)
)

tangent <- data.frame(
  x = seq(-1, 6, length.out = 200)
)

tangent$y <- 4 - (3 / 4) * (tangent$x - 3)

ggplot() +
  geom_path(data = circle, aes(x = x, y = y), linewidth = 1.1) +
  geom_line(data = tangent, aes(x = x, y = y), linetype = "dashed", linewidth = 1.1) +
  annotate("point", x = 3, y = 4, size = 3) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_equal() +
  labs(
    title = "Implicit differentiation on a circle",
    subtitle = "At (3, 4), the gradient is -3/4.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Find $\dfrac{dy}{dx}$ for

$$
x^2+xy+y^2=7.
$$

Differentiate each term with respect to $x$.

The derivative of $x^2$ is $2x$.

For $xy$, use the product rule:

$$
\dfrac{d}{dx}(xy)=x\dfrac{dy}{dx}+y.
$$

The derivative of $y^2$ is

$$
2y\dfrac{dy}{dx}.
$$

So

$$
2x+x\dfrac{dy}{dx}+y+2y\dfrac{dy}{dx}=0.
$$

Collecting the $\dfrac{dy}{dx}$ terms gives

$$
(x+2y)\dfrac{dy}{dx}=-(2x+y).
$$

Therefore,

$$
\dfrac{dy}{dx}
=
-\dfrac{2x+y}{x+2y}.
$$

### Short investigation

Change the point on the circle and observe the tangent.

```{r}
#| label: fig-implicit-circle-changing-point
#| fig-cap: "Changing the point of tangency on x^2 + y^2 = 25."
#| fig-width: 6
#| fig-height: 5

x0 <- 4
y0 <- 3

theta <- seq(0, 2 * pi, length.out = 800)

circle <- data.frame(
  x = 5 * cos(theta),
  y = 5 * sin(theta)
)

m <- -x0 / y0

tangent <- data.frame(
  x = seq(-1, 6, length.out = 200)
)

tangent$y <- y0 + m * (tangent$x - x0)

ggplot() +
  geom_path(data = circle, aes(x = x, y = y), linewidth = 1.1) +
  geom_line(data = tangent, aes(x = x, y = y), linetype = "dashed", linewidth = 1.1) +
  annotate("point", x = x0, y = y0, size = 3) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_equal() +
  labs(
    title = "Tangent to x^2 + y^2 = 25",
    subtitle = paste0("At (", x0, ", ", y0, "), the gradient is ", round(m, 3), "."),
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Exercises 4.6

#### Bronze

1. Find $\dfrac{dy}{dx}$ if $x^2+y^2=25$.
2. Find $\dfrac{dy}{dx}$ if $xy=6$.
3. Find $\dfrac{dy}{dx}$ if $y^2=4x$.
4. Find $\dfrac{dy}{dx}$ if $x^2+xy=10$.
5. Find $\dfrac{dy}{dx}$ if $x+y^3=7$.

#### Silver

1. Find the gradient of $x^2+y^2=25$ at $(3,4)$.
2. Find the equation of the tangent to $xy=12$ at $(3,4)$.
3. Find $\dfrac{dy}{dx}$ if $x^2+xy+y^2=7$.
4. Find the gradient of $y^2+xy=6$ at $(2,2)$.
5. Find the equation of the normal to $x^2+y^2=10$ at $(1,3)$.

#### Gold

1. For the circle $x^2+y^2=r^2$, prove that the tangent is perpendicular to the radius at any point on the circle.
2. Find $\dfrac{dy}{dx}$ for $x^3+y^3=6xy$.
3. Find the equation of the tangent to $x^2+xy+y^2=3$ at $(1,1)$.
4. Find the points on $x^2+y^2=25$ where the tangent has gradient $\dfrac{3}{4}$.
5. If $\ln(xy)=x-y$, find $\dfrac{dy}{dx}$ in terms of $x$ and $y$.

## 4.7 Parametric differentiation

### Key mathematical ideas

A curve may be defined parametrically by

$$
x=f(t),
\qquad
y=g(t).
$$

If both $x$ and $y$ depend on $t$, then

$$
\dfrac{dy}{dx}
=
\dfrac{\dfrac{dy}{dt}}{\dfrac{dx}{dt}},
$$

provided

$$
\dfrac{dx}{dt}\neq0.
$$

This is the chain rule in the form

$$
\dfrac{dy}{dt}
=
\dfrac{dy}{dx}\dfrac{dx}{dt}.
$$

### Visualisation using R

```{r}
#| label: fig-parametric-circle
#| fig-cap: "The parametric curve x = cos(t), y = sin(t)."
#| fig-width: 6
#| fig-height: 5

t <- seq(0, 2 * pi, length.out = 800)

df <- data.frame(
  t = t,
  x = cos(t),
  y = sin(t)
)

ggplot(df, aes(x = x, y = y)) +
  geom_path(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  coord_equal() +
  labs(
    title = "Parametric circle",
    subtitle = "x = cos(t), y = sin(t).",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

A curve is defined by

$$
x=t^2,
\qquad
y=t^3.
$$

Find $\dfrac{dy}{dx}$.

First,

$$
\dfrac{dx}{dt}=2t
$$

and

$$
\dfrac{dy}{dt}=3t^2.
$$

Therefore,

$$
\dfrac{dy}{dx}
=
\dfrac{\dfrac{dy}{dt}}{\dfrac{dx}{dt}}
=
\dfrac{3t^2}{2t}.
$$

For $t\neq0$,

$$
\dfrac{dy}{dx}=\dfrac{3t}{2}.
$$

### Short investigation

Change the parametric equations and observe the gradient.

```{r}
#| label: fig-parametric-cubic
#| fig-cap: "The parametric curve x = t^2, y = t^3 - 3t."
#| fig-width: 6
#| fig-height: 5

t <- seq(-3, 3, length.out = 800)

df <- data.frame(
  t = t,
  x = t^2,
  y = t^3 - 3 * t
)

ggplot(df, aes(x = x, y = y)) +
  geom_path(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(
    title = "Parametric curve",
    subtitle = "x = t^2, y = t^3 - 3t.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Exercises 4.7

#### Bronze

1. Find $\dfrac{dy}{dx}$ if $x=t^2$ and $y=t^3$.
2. Find $\dfrac{dy}{dx}$ if $x=2t+1$ and $y=t^2$.
3. Find $\dfrac{dy}{dx}$ if $x=\sin t$ and $y=\cos t$.
4. Find $\dfrac{dy}{dx}$ if $x=e^t$ and $y=e^{2t}$.
5. Find $\dfrac{dy}{dx}$ if $x=t+\dfrac{1}{t}$ and $y=t-\dfrac{1}{t}$.

#### Silver

1. Find the equation of the tangent to $x=t^2$, $y=t^3$ at $t=2$.
2. For $x=t^2+1$ and $y=t^3-t$, find the values of $t$ where the tangent is parallel to the $x$-axis.
3. Find $\dfrac{dy}{dx}$ if $x=2\cos t$ and $y=3\sin t$.
4. Find $\dfrac{dy}{dx}$ if $x=t^2$ and $y=\ln t$, where $t>0$.
5. Find the equation of the tangent to $x=t+1$, $y=t^2+2t$ at $t=1$.

#### Gold

1. For $x=t^2$ and $y=t^3-3t$, find the values of $t$ where the tangent is horizontal or vertical.
2. For $x=e^t$ and $y=te^t$, find the equation of the tangent at $t=0$.
3. Find $\dfrac{dy}{dx}$ if $x=\cos t$ and $y=\sin2t$.
4. For $x=t^2+1$ and $y=t^3-3t$, find the values of $t$ where the tangent is parallel to $y=3x$.
5. For $x=t^2$ and $y=t^3$, where $t>0$, show that
   $$
   \dfrac{dy}{dx}=\dfrac{3}{2}\sqrt{x}.
   $$

## STEP-style extension questions

These questions are designed to require more than routine application.

### STEP-style Question 1

Let $n$ be a positive integer and define

$$
f(x)=x^ne^{-x},
\qquad x>0.
$$

Show that $f$ has exactly one stationary point, and find its maximum value.

### STEP-style Question 2

Let

$$
f(x)=x+\dfrac{1}{x},
\qquad x>0.
$$

Use differentiation to prove that

$$
x+\dfrac{1}{x}\geq2.
$$

### STEP-style Question 3

The curve

$$
x^2+xy+y^2=3
$$

has tangents parallel to the coordinate axes at four points.

Find these four points.

### STEP-style Question 4

The parametric curve

$$
x=t^2+1,
\qquad
y=t^3-3t
$$

has tangents parallel to the line $y=3x$.

Find the corresponding values of $t$ and the coordinates of the points.

### STEP-style Question 5

Find constants $a$ and $b$ such that

$$
f(x)=e^x(a\sin x+b\cos x)
$$

satisfies

$$
f'(x)=e^x\sin x.
$$

## Reflection questions

1. Why is the product rule not simply the product of the derivatives?
2. Why is the quotient rule more delicate than differentiating numerator and denominator separately?
3. What makes $e^x$ special among exponential functions?
4. Why must logarithmic differentiation respect the domain of the logarithm?
5. Why must trigonometric derivatives use radians?
6. What is the difference between explicit and implicit differentiation?
7. How does parametric differentiation depend on the chain rule?
8. What can R show clearly about gradients that algebra alone may hide?
9. What can exact differentiation prove that numerical graphing can only suggest?

## Answers

### Answers to Exercises 4.1

#### Bronze

1. $\dfrac{dy}{dx}=2x+3$.
2. $\dfrac{dy}{dx}=3x^2-2x$.
3. $\dfrac{dy}{dx}=8x^3+3x^2$.
4. $\dfrac{dy}{dx}=3x^2+2x+2$.
5. $\dfrac{dy}{dx}=12x+5$.

#### Silver

1. $\dfrac{dy}{dx}=e^x(x^2+2x)$.
2. $\dfrac{dy}{dx}=e^{2x}(2x+3)$.
3. $\dfrac{dy}{dx}=x^2(3\ln x+1)$.
4. $\dfrac{dy}{dx}=2\sin x+(2x-1)\cos x$.
5. $\dfrac{dy}{dx}=2x\cos3x-3x^2\sin3x$.

#### Gold

1. The stationary point is $\left(1,\dfrac{1}{e}\right)$.
2. The tangent is $y=x-1$.
3. $\dfrac{dy}{dx}=-2e^{-2x}(x^2-x+1)$. Since $x^2-x+1>0$ for all real $x$, there are no stationary points.
4. Using the product rule,
   $$
   \dfrac{dy}{dx}=nx^{n-1}e^x+x^ne^x=x^{n-1}e^x(x+n).
   $$
5. $x=-1+\sqrt5$ or $x=-1-\sqrt5$.

### Answers to Exercises 4.2

#### Bronze

1. $\dfrac{dy}{dx}=-\dfrac{3}{(x-2)^2}$.
2. $\dfrac{dy}{dx}=\dfrac{11}{(x+4)^2}$.
3. $\dfrac{dy}{dx}=\dfrac{x^2+2x}{(x+1)^2}$.
4. $\dfrac{dy}{dx}=\dfrac{x^2-1}{x^2}$.
5. $\dfrac{dy}{dx}=-\dfrac{2x}{(x^2+1)^2}$.

#### Silver

1. $\dfrac{dy}{dx}=\dfrac{x^2-2x-3}{(x-1)^2}$.
2. $\dfrac{dy}{dx}=\dfrac{e^x(x-1)}{x^2}$.
3. $\dfrac{dy}{dx}=\dfrac{1-\ln x}{x^2}$.
4. The gradient is $-1$.
5. $x=1+\sqrt2$ or $x=1-\sqrt2$.

#### Gold

1. The tangent is $y=4x-4$.
2. The normal is
   $$
   y+\dfrac{3}{2}=\dfrac{4}{7}(x-1).
   $$
3. The stationary points are $\left(-1,-\dfrac{1}{2}\right)$ and $\left(1,\dfrac{1}{2}\right)$. The first is a minimum and the second is a maximum.
4. Using the quotient rule,
   $$
   \dfrac{d}{dx}\left(\dfrac{ax+b}{cx+d}\right)
   =
   \dfrac{a(cx+d)-c(ax+b)}{(cx+d)^2}
   =
   \dfrac{ad-bc}{(cx+d)^2}.
   $$
5. $a=-7$.

### Answers to Exercises 4.3

#### Bronze

1. $\dfrac{dy}{dx}=e^x$.
2. $\dfrac{dy}{dx}=3e^{3x}$.
3. $\dfrac{dy}{dx}=10e^{2x}$.
4. $\dfrac{dy}{dx}=-e^{-x}$.
5. $\dfrac{dy}{dx}=-4e^{1-4x}$.

#### Silver

1. $\dfrac{dy}{dx}=e^x(x+1)$.
2. $\dfrac{dy}{dx}=e^{3x}(3x^2+2x)$.
3. $x=\dfrac{1+\ln5}{2}$.
4. The stationary point is $(1,-e)$.
5. The tangent is $y=2x+1$.

#### Gold

1. $x=\dfrac{3+\sqrt{21}}{2}$ or $x=\dfrac{3-\sqrt{21}}{2}$.
2. The stationary point is $(0,1)$.
3. At $x=0$, the tangent is $y=1-x$, which meets the $x$-axis at $x=1$.
4. $x=0$ or $x=\ln2$.
5. The minimum value is $2$.

### Answers to Exercises 4.4

#### Bronze

1. $\dfrac{dy}{dx}=\dfrac{1}{x}$.
2. $\dfrac{dy}{dx}=\dfrac{1}{x}$.
3. $\dfrac{dy}{dx}=\dfrac{1}{x+2}$.
4. $\dfrac{dy}{dx}=\dfrac{5}{x}$.
5. $\dfrac{dy}{dx}=\dfrac{2}{x}$.

#### Silver

1. $\dfrac{dy}{dx}=\dfrac{2}{2x-1}$.
2. $\dfrac{dy}{dx}=\ln x+1$.
3. $\dfrac{dy}{dx}=\dfrac{2x}{x^2+1}$.
4. The stationary point occurs when $x=e$.
5. The tangent is $y=x$.

#### Gold

1. The stationary point occurs when $x=e^{-1/2}$.
2. The stationary point occurs when $x=\sqrt e$.
3. Since $\ln\sqrt{x^2+1}=\dfrac{1}{2}\ln(x^2+1)$,
   $$
   \dfrac{d}{dx}\left(\ln\sqrt{x^2+1}\right)
   =
   \dfrac{1}{2}\cdot\dfrac{2x}{x^2+1}
   =
   \dfrac{x}{x^2+1}.
   $$
4. $x=2$.
5. The tangent is $y=2x$.

### Answers to Exercises 4.5

#### Bronze

1. $\dfrac{dy}{dx}=\cos x$.
2. $\dfrac{dy}{dx}=-\sin x$.
3. $\dfrac{dy}{dx}=\sec^2x$.
4. $\dfrac{dy}{dx}=2\cos2x$.
5. $\dfrac{dy}{dx}=-3\sin(3x+1)$.

#### Silver

1. $\dfrac{dy}{dx}=\sin x+x\cos x$.
2. $\dfrac{dy}{dx}=e^x(\sin x+\cos x)$.
3. $\dfrac{dy}{dx}=2\sin x\cos x$.
4. $\dfrac{dy}{dx}=2\sec^2(2x)$.
5. The tangent is
   $$
   y-\dfrac{1}{2}
   =
   \dfrac{\sqrt3}{2}\left(x-\dfrac{\pi}{6}\right).
   $$

#### Gold

1. The stationary points are $x=\dfrac{\pi}{4}$ and $x=\dfrac{5\pi}{4}$. The first is a maximum with value $\sqrt2$, and the second is a minimum with value $-\sqrt2$.
2. $x=\dfrac{\pi}{4}$ and $x=\dfrac{3\pi}{4}$.
3. $x=\dfrac{\pi}{6}$ or $x=\dfrac{5\pi}{6}$.
4. Since $\sec x=\dfrac{1}{\cos x}$,
   $$
   \dfrac{d}{dx}(\sec x)
   =
   \dfrac{\sin x}{\cos^2x}
   =
   \sec x\tan x.
   $$
5. The tangent is
   $$
   y-1=2\left(x-\dfrac{\pi}{4}\right).
   $$

### Answers to Exercises 4.6

#### Bronze

1. $\dfrac{dy}{dx}=-\dfrac{x}{y}$.
2. $\dfrac{dy}{dx}=-\dfrac{y}{x}$.
3. $\dfrac{dy}{dx}=\dfrac{2}{y}$.
4. $\dfrac{dy}{dx}=-\dfrac{2x+y}{x}$.
5. $\dfrac{dy}{dx}=-\dfrac{1}{3y^2}$.

#### Silver

1. The gradient is $-\dfrac{3}{4}$.
2. The tangent is
   $$
   y-4=-\dfrac{4}{3}(x-3).
   $$
3. $\dfrac{dy}{dx}=-\dfrac{2x+y}{x+2y}$.
4. The gradient is $-\dfrac{1}{3}$.
5. The normal is
   $$
   y-3=3(x-1).
   $$

#### Gold

1. For $x^2+y^2=r^2$,
   $$
   \dfrac{dy}{dx}=-\dfrac{x}{y}.
   $$
   The radius from the origin to $(x,y)$ has gradient $\dfrac{y}{x}$. The product of the two gradients is $-1$, so the tangent is perpendicular to the radius.
2. $\dfrac{dy}{dx}=\dfrac{2y-x^2}{y^2-2x}$.
3. The tangent is $y-1=-(x-1)$.
4. The points are $(-3,4)$ and $(3,-4)$.
5. $\dfrac{dy}{dx}=\dfrac{y(x-1)}{x(y+1)}$.

### Answers to Exercises 4.7

#### Bronze

1. $\dfrac{dy}{dx}=\dfrac{3t}{2}$.
2. $\dfrac{dy}{dx}=t$.
3. $\dfrac{dy}{dx}=-\tan t$.
4. $\dfrac{dy}{dx}=2e^t$.
5. $\dfrac{dy}{dx}=\dfrac{t^2+1}{t^2-1}$.

#### Silver

1. The tangent is $y-8=3(x-4)$.
2. $t=\dfrac{1}{\sqrt3}$ or $t=-\dfrac{1}{\sqrt3}$.
3. $\dfrac{dy}{dx}=-\dfrac{3\cos t}{2\sin t}$.
4. $\dfrac{dy}{dx}=\dfrac{1}{2t^2}$.
5. The tangent is $y-3=4(x-2)$.

#### Gold

1. Horizontal tangents occur at $t=1$ and $t=-1$. A vertical tangent occurs at $t=0$.
2. The tangent is $y=x-1$.
3. $\dfrac{dy}{dx}=-\dfrac{2\cos2t}{\sin t}$.
4. $t=1+\sqrt2$ or $t=1-\sqrt2$.
5. Since $t>0$, $t=\sqrt x$. Therefore,
   $$
   \dfrac{dy}{dx}=\dfrac{3t}{2}=\dfrac{3}{2}\sqrt x.
   $$

### Answers to STEP-style extension questions

1. We have
   $$
   f'(x)=nx^{n-1}e^{-x}-x^ne^{-x}
   =
   x^{n-1}e^{-x}(n-x).
   $$
   Since $x>0$, the factors $x^{n-1}$ and $e^{-x}$ are positive. Therefore $f'(x)=0$ only when $x=n$. The derivative changes from positive to negative at $x=n$, so this is a maximum. The maximum value is
   $$
   f(n)=n^ne^{-n}.
   $$

2. For $x>0$,
   $$
   f'(x)=1-\dfrac{1}{x^2}.
   $$
   Thus $f'(x)=0$ when $x=1$. The derivative is negative for $0<x<1$ and positive for $x>1$, so $x=1$ gives a minimum. Therefore,
   $$
   f(x)\geq f(1)=2.
   $$
   Hence
   $$
   x+\dfrac{1}{x}\geq2.
   $$

3. For
   $$
   x^2+xy+y^2=3,
   $$
   implicit differentiation gives
   $$
   \dfrac{dy}{dx}=-\dfrac{2x+y}{x+2y}.
   $$
   Horizontal tangents occur when $2x+y=0$, so $y=-2x$. Substitution gives $x=\pm1$, producing $(1,-2)$ and $(-1,2)$. Vertical tangents occur when $x+2y=0$, so $x=-2y$. Substitution gives $y=\pm1$, producing $(-2,1)$ and $(2,-1)$.

4. For
   $$
   x=t^2+1,
   \qquad
   y=t^3-3t,
   $$
   we have
   $$
   \dfrac{dy}{dx}=\dfrac{3t^2-3}{2t}.
   $$
   Setting this equal to $3$ gives
   $$
   \dfrac{3t^2-3}{2t}=3.
   $$
   Hence
   $$
   t^2-2t-1=0.
   $$
   Therefore,
   $$
   t=1+\sqrt2
   \quad \text{or} \quad
   t=1-\sqrt2.
   $$
   The corresponding points are
   $$
   (4+2\sqrt2,\ 4+2\sqrt2)
   $$
   and
   $$
   (4-2\sqrt2,\ 4-2\sqrt2).
   $$

5. Let
   $$
   f(x)=e^x(a\sin x+b\cos x).
   $$
   Then
   $$
   f'(x)
   =
   e^x(a\sin x+b\cos x)
   +
   e^x(a\cos x-b\sin x).
   $$
   So
   $$
   f'(x)
   =
   e^x\left((a-b)\sin x+(a+b)\cos x\right).
   $$
   We require
   $$
   (a-b)\sin x+(a+b)\cos x=\sin x.
   $$
   Therefore,
   $$
   a-b=1
   $$
   and
   $$
   a+b=0.
   $$
   Solving gives
   $$
   a=\dfrac{1}{2},
   \qquad
   b=-\dfrac{1}{2}.
   $$