# Integration

```{r}
#| label: setup-integration
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}
```

## Overview

This chapter develops five connected integration ideas:

1. integration of exponential functions;
2. integration of expressions of the form $\dfrac{1}{ax+b}$;
3. integration of $\sin(ax+b)$, $\cos(ax+b)$ and $\sec^2(ax+b)$;
4. further integration of trigonometric functions;
5. the trapezium rule.

The central aim is to treat integration as reverse differentiation, while also developing numerical and graphical ways to estimate areas.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 5.1 Integration of exponential functions

### Key mathematical ideas

Since

$$
\dfrac{d}{dx}(e^x)=e^x,
$$

we have

$$
\int e^x\,dx=e^x+C.
$$

More generally,

$$
\int e^{ax+b}\,dx=\dfrac{1}{a}e^{ax+b}+C,
\quad a\neq0.
$$

This is because differentiating $\dfrac{1}{a}e^{ax+b}$ gives

$$
\dfrac{1}{a}\cdot ae^{ax+b}=e^{ax+b}.
$$

### Visualisation using R

```{r}
#| label: fig-exp-integration-area
#| fig-cap: "The area under y = e^x from x = 0 to x = 1."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(0, 1, length.out = 500)
)

df$y <- exp(df$x)

ggplot(df, aes(x = x, y = y)) +
  geom_area(alpha = 0.3) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Area under y = e^x",
    subtitle = "The exact area from 0 to 1 is e - 1.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

The exact area is

$$
\int_0^1 e^x\,dx
=
\left[e^x\right]_0^1
=
e-1.
$$

### Worked example

Find

$$
\int e^{3x-2}\,dx.
$$

Since the derivative of $3x-2$ is $3$, we divide by $3$:

$$
\int e^{3x-2}\,dx
=
\dfrac{1}{3}e^{3x-2}+C.
$$

Check by differentiating:

$$
\dfrac{d}{dx}\left(\dfrac{1}{3}e^{3x-2}\right)
=
\dfrac{1}{3}\cdot3e^{3x-2}
=
e^{3x-2}.
$$

### Short investigation

Change the value of $a$ in the code below and compare $e^{ax}$ with its antiderivative.

```{r}
#| label: fig-exp-antiderivative-family
#| fig-cap: "The function y = e^(ax) and an antiderivative."
#| fig-width: 6
#| fig-height: 4

a <- 2

df <- data.frame(
  x = seq(-2, 2, length.out = 600)
)

df$y_function <- exp(a * df$x)
df$antiderivative <- (1 / a) * exp(a * df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y_function, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = antiderivative, linetype = "antiderivative"), linewidth = 1.1) +
  labs(
    title = paste0("Function and antiderivative for y = e^(", a, "x)"),
    subtitle = "The antiderivative is divided by the coefficient of x.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 5.1

#### Bronze

1. Find $\int e^x\,dx$.
2. Find $\int e^{3x}\,dx$.
3. Find $\int 5e^{2x}\,dx$.
4. Find $\int e^{-x}\,dx$.
5. Find $\int 4e^{1-2x}\,dx$.

#### Silver

1. Find $\int \left(e^x+2e^{3x}\right)\,dx$.
2. Find $\int 3e^{2x-1}\,dx$.
3. Evaluate $\int_0^1 e^{2x}\,dx$.
4. Evaluate $\int_0^{\ln2} e^x\,dx$.
5. Find $a$ if $\int e^{ax}\,dx=\dfrac{1}{5}e^{5x}+C$.

#### Gold

1. Evaluate $\int_0^1 4e^{4x}\,dx$.
2. Find $\int e^{3x+2}\,dx$.
3. Evaluate $\int_0^{\ln3} 2e^{2x}\,dx$.
4. Find $k$ if $\dfrac{d}{dx}\left(ke^{-2x}\right)=e^{-2x}$.
5. The area under $y=3e^x$ from $x=0$ to $x=a$ is $6$. Find $a$.

## 5.2 Integration of $\dfrac{1}{ax+b}$

### Key mathematical ideas

Since

$$
\dfrac{d}{dx}\left(\ln|x|\right)=\dfrac{1}{x},
$$

we have

$$
\int \dfrac{1}{x}\,dx=\ln|x|+C.
$$

More generally,

$$
\int \dfrac{1}{ax+b}\,dx
=
\dfrac{1}{a}\ln|ax+b|+C,
\quad a\neq0.
$$

This follows because differentiating $\ln|ax+b|$ gives

$$
\dfrac{a}{ax+b}.
$$

Therefore, if the numerator is exactly the derivative of the denominator, the integral becomes a logarithm.

### Visualisation using R

```{r}
#| label: fig-log-integration-area
#| fig-cap: "The area under y = 1/(x + 1) from x = 0 to x = 2."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(0, 2, length.out = 500)
)

df$y <- 1 / (df$x + 1)

ggplot(df, aes(x = x, y = y)) +
  geom_area(alpha = 0.3) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Area under y = 1/(x + 1)",
    subtitle = "The exact area from 0 to 2 is ln 3.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

The exact area is

$$
\int_0^2 \dfrac{1}{x+1}\,dx
=
\left[\ln|x+1|\right]_0^2
=
\ln3.
$$

### Worked example

Find

$$
\int \dfrac{5}{2x-3}\,dx.
$$

Since

$$
\int \dfrac{1}{2x-3}\,dx
=
\dfrac{1}{2}\ln|2x-3|+C,
$$

we have

$$
\int \dfrac{5}{2x-3}\,dx
=
\dfrac{5}{2}\ln|2x-3|+C.
$$

### Short investigation

Change the values of $a$ and $b$ below.

```{r}
#| label: fig-log-antiderivative-family
#| fig-cap: "The function y = 1/(ax + b) and an antiderivative."
#| fig-width: 6
#| fig-height: 4

a <- 2
b <- 1

df <- data.frame(
  x = seq(0.1, 5, length.out = 600)
)

df$y_function <- 1 / (a * df$x + b)
df$antiderivative <- (1 / a) * log(abs(a * df$x + b))

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y_function, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = antiderivative, linetype = "antiderivative"), linewidth = 1.1) +
  labs(
    title = paste0("Function and antiderivative for y = 1/(", a, "x + ", b, ")"),
    subtitle = "The antiderivative is logarithmic.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 5.2

#### Bronze

1. Find $\int \dfrac{1}{x}\,dx$.
2. Find $\int \dfrac{1}{x+2}\,dx$.
3. Find $\int \dfrac{1}{2x+1}\,dx$.
4. Find $\int \dfrac{3}{3x-4}\,dx$.
5. Find $\int \dfrac{5}{x-1}\,dx$.

#### Silver

1. Find $\int \dfrac{4}{2x+3}\,dx$.
2. Find $\int \dfrac{7}{5x-2}\,dx$.
3. Evaluate $\int_0^1 \dfrac{1}{x+1}\,dx$.
4. Evaluate $\int_1^3 \dfrac{2}{2x+1}\,dx$.
5. Find $a$ and $b$ if
   $$
   \int \dfrac{1}{ax+b}\,dx=\dfrac{1}{4}\ln|4x-3|+C.
   $$

#### Gold

1. Evaluate $\int_0^2 \dfrac{3}{3x+1}\,dx$.
2. Evaluate $\int_0^2 \dfrac{5}{5x+2}\,dx$.
3. Solve
   $$
   \int_0^a \dfrac{1}{x+1}\,dx=\ln5.
   $$
4. Find
   $$
   \int \left(\dfrac{1}{x+1}+\dfrac{1}{x+2}\right)\,dx.
   $$
5. Find $k$ if
   $$
   \int_0^1 \dfrac{k}{2x+1}\,dx=\ln3.
   $$

## 5.3 Integration of $\sin(ax+b)$, $\cos(ax+b)$ and $\sec^2(ax+b)$

### Key mathematical ideas

Angles are measured in radians.

The standard trigonometric integrals are

$$
\int \sin x\,dx=-\cos x+C,
$$

$$
\int \cos x\,dx=\sin x+C,
$$

and

$$
\int \sec^2x\,dx=\tan x+C.
$$

Using the chain rule in reverse,

$$
\int \sin(ax+b)\,dx
=
-\dfrac{1}{a}\cos(ax+b)+C,
$$

$$
\int \cos(ax+b)\,dx
=
\dfrac{1}{a}\sin(ax+b)+C,
$$

and

$$
\int \sec^2(ax+b)\,dx
=
\dfrac{1}{a}\tan(ax+b)+C.
$$

### Visualisation using R

```{r}
#| label: fig-trig-integration
#| fig-cap: "The function y = cos(2x) and an antiderivative."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-pi, pi, length.out = 800)
)

df$y_function <- cos(2 * df$x)
df$antiderivative <- 0.5 * sin(2 * df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y_function, linetype = "function"), linewidth = 1.1) +
  geom_line(aes(y = antiderivative, linetype = "antiderivative"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Function and antiderivative for y = cos(2x)",
    subtitle = "The antiderivative is 0.5 sin(2x).",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Find

$$
\int 6\sin(3x-1)\,dx.
$$

Since

$$
\int \sin(3x-1)\,dx
=
-\dfrac{1}{3}\cos(3x-1)+C,
$$

we have

$$
\int 6\sin(3x-1)\,dx
=
-2\cos(3x-1)+C.
$$

Check by differentiating:

$$
\dfrac{d}{dx}\left(-2\cos(3x-1)\right)
=
6\sin(3x-1).
$$

### Short investigation

Compare $\sec^2x$ and its antiderivative $\tan x$.

```{r}
#| label: fig-sec2-tan
#| fig-cap: "The function y = sec^2(x) and its antiderivative y = tan(x)."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-1.2, 1.2, length.out = 600)
)

df$y_function <- 1 / cos(df$x)^2
df$antiderivative <- tan(df$x)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = y_function, linetype = "sec^2(x)"), linewidth = 1.1) +
  geom_line(aes(y = antiderivative, linetype = "tan(x)"), linewidth = 1.1) +
  labs(
    title = "Function and antiderivative",
    subtitle = "The derivative of tan(x) is sec^2(x).",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 5.3

#### Bronze

1. Find $\int \cos x\,dx$.
2. Find $\int \sin x\,dx$.
3. Find $\int \sec^2x\,dx$.
4. Find $\int \cos2x\,dx$.
5. Find $\int \sin3x\,dx$.

#### Silver

1. Find $\int 4\cos(2x+1)\,dx$.
2. Find $\int 6\sin(3x-2)\,dx$.
3. Find $\int 5\sec^2(5x)\,dx$.
4. Evaluate $\int_0^{\dfrac{\pi}{2}}\cos x\,dx$.
5. Evaluate $\int_0^{\dfrac{\pi}{4}}\sec^2x\,dx$.

#### Gold

1. Evaluate $\int_0^{\dfrac{\pi}{6}}3\cos3x\,dx$.
2. Evaluate $\int_0^{\dfrac{\pi}{2}}2\sin x\,dx$.
3. Evaluate $\int_0^{\dfrac{\pi}{8}}4\sec^2(2x)\,dx$.
4. Find $a$ if
   $$
   \int_0^{\dfrac{\pi}{6}}a\cos2x\,dx=3.
   $$
5. Solve
   $$
   \int_0^x \cos2t\,dt=\dfrac{1}{2}
   $$
   for $0\leq x\leq\dfrac{\pi}{2}$.

## 5.4 Further integration of trigonometric functions

### Key mathematical ideas

Some trigonometric integrals require identities before they can be integrated.

The most useful identities in this section are

$$
\sin^2x=\dfrac{1-\cos2x}{2},
$$

$$
\cos^2x=\dfrac{1+\cos2x}{2},
$$

and

$$
2\sin x\cos x=\sin2x.
$$

These identities turn powers or products of trigonometric functions into expressions that can be integrated directly.

### Visualisation using R

```{r}
#| label: fig-sin-squared-identity
#| fig-cap: "Checking sin^2(x) = (1 - cos(2x))/2."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(0, 2 * pi, length.out = 800)
)

df$left_side <- sin(df$x)^2
df$right_side <- (1 - cos(2 * df$x)) / 2

ggplot(df, aes(x = x)) +
  geom_line(aes(y = left_side, linetype = "sin^2(x)"), linewidth = 1.1) +
  geom_line(aes(y = right_side, linetype = "(1 - cos(2x))/2"), linewidth = 1.1) +
  labs(
    title = "Identity for sin^2(x)",
    subtitle = "The two graphs overlap.",
    x = "x",
    y = "value",
    linetype = "Expression"
  ) +
  book_theme()
```

### Worked example

Find

$$
\int \sin^2x\,dx.
$$

Using the identity

$$
\sin^2x=\dfrac{1-\cos2x}{2},
$$

we get

$$
\int \sin^2x\,dx
=
\int \dfrac{1-\cos2x}{2}\,dx.
$$

So

$$
\int \sin^2x\,dx
=
\dfrac{x}{2}-\dfrac{1}{4}\sin2x+C.
$$

### Short investigation

Compare exact area under $\sin^2x$ with numerical area.

```{r}
#| label: fig-sin-squared-area
#| fig-cap: "The area under y = sin^2(x) from 0 to pi."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(0, pi, length.out = 800)
)

df$y <- sin(df$x)^2

ggplot(df, aes(x = x, y = y)) +
  geom_area(alpha = 0.3) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Area under y = sin^2(x)",
    subtitle = "The exact area from 0 to pi is pi/2.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Exercises 5.4

#### Bronze

1. Find $\int \sin^2x\,dx$.
2. Find $\int \cos^2x\,dx$.
3. Find $\int \sin x\cos x\,dx$.
4. Find $\int (1+\cos2x)\,dx$.
5. Find $\int (1-\cos2x)\,dx$.

#### Silver

1. Evaluate $\int_0^\pi \sin^2x\,dx$.
2. Evaluate $\int_0^\pi \cos^2x\,dx$.
3. Evaluate $\int_0^{\dfrac{\pi}{2}}\sin x\cos x\,dx$.
4. Find $\int 4\cos^2x\,dx$.
5. Find $\int 6\sin^2(2x)\,dx$.

#### Gold

1. Evaluate $\int_0^{\dfrac{\pi}{4}}\sin^2x\,dx$.
2. Evaluate $\int_0^{\dfrac{\pi}{4}}\cos^2x\,dx$.
3. Evaluate $\int_0^{\dfrac{\pi}{2}}(\sin x+\cos x)^2\,dx$.
4. Find $k$ if
   $$
   \int_0^\pi k\sin^2x\,dx=5\pi.
   $$
5. Evaluate $\int_0^{\dfrac{\pi}{2}}\sin^2x\cos^2x\,dx$.

## 5.5 The trapezium rule

### Key mathematical ideas

The trapezium rule estimates the area under a curve by splitting the interval into equal strips and replacing each curved strip with a trapezium.

Suppose we want to estimate

$$
\int_a^b f(x)\,dx
$$

using $n$ equal strips.

The strip width is

$$
h=\dfrac{b-a}{n}.
$$

The $x$-values are

$$
x_0=a,\quad x_1=a+h,\quad x_2=a+2h,\quad \ldots,\quad x_n=b.
$$

Let

$$
y_i=f(x_i).
$$

Then the trapezium rule gives

$$
\int_a^b f(x)\,dx
\approx
\dfrac{h}{2}
\left[
y_0+2y_1+2y_2+\cdots+2y_{n-1}+y_n
\right].
$$

### Visualisation using R

```{r}
#| label: fig-trapezium-rule
#| fig-cap: "Trapezium rule estimate for y = x^2 from x = 0 to x = 2."
#| fig-width: 6
#| fig-height: 4

f <- function(x) x^2

curve_df <- data.frame(
  x = seq(0, 2, length.out = 800)
)

curve_df$y <- f(curve_df$x)

x_points <- seq(0, 2, length.out = 5)
y_points <- f(x_points)

trapezia <- do.call(
  rbind,
  lapply(1:(length(x_points) - 1), function(i) {
    data.frame(
      strip = i,
      x = c(x_points[i], x_points[i], x_points[i + 1], x_points[i + 1]),
      y = c(0, y_points[i], y_points[i + 1], 0)
    )
  })
)

ggplot() +
  geom_polygon(
    data = trapezia,
    aes(x = x, y = y, group = strip),
    alpha = 0.3
  ) +
  geom_line(
    data = curve_df,
    aes(x = x, y = y),
    linewidth = 1.1
  ) +
  geom_point(
    data = data.frame(x = x_points, y = y_points),
    aes(x = x, y = y),
    size = 2
  ) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Trapezium rule",
    subtitle = "The curve y = x^2 is approximated by straight-line segments.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Use the trapezium rule with $4$ strips to estimate

$$
\int_0^2 x^2\,dx.
$$

Here,

$$
a=0,\quad b=2,\quad n=4.
$$

Therefore,

$$
h=\dfrac{2-0}{4}=\dfrac{1}{2}.
$$

The $x$-values are

$$
0,\quad \dfrac{1}{2},\quad 1,\quad \dfrac{3}{2},\quad 2.
$$

The corresponding $y$-values for $y=x^2$ are

$$
0,\quad \dfrac{1}{4},\quad 1,\quad \dfrac{9}{4},\quad 4.
$$

Using the trapezium rule,

$$
T
=
\dfrac{h}{2}
\left[
y_0+2y_1+2y_2+2y_3+y_4
\right].
$$

So

$$
T
=
\dfrac{1}{4}
\left[
0+2\left(\dfrac{1}{4}\right)+2(1)+2\left(\dfrac{9}{4}\right)+4
\right].
$$

Therefore,

$$
T=\dfrac{11}{4}.
$$

### Short investigation

Change the number of strips and observe the estimate.

```{r}
#| label: fig-trapezium-estimate-changing-n
#| fig-cap: "Trapezium rule estimates for y = x^2 on [0, 2]."
#| fig-width: 6
#| fig-height: 4

trapezium_estimate <- function(f, a, b, n) {
  x <- seq(a, b, length.out = n + 1)
  y <- f(x)
  h <- (b - a) / n
  (h / 2) * (y[1] + 2 * sum(y[2:n]) + y[n + 1])
}

n_values <- 1:20

df <- data.frame(
  n = n_values,
  estimate = sapply(
    n_values,
    function(n) trapezium_estimate(function(x) x^2, 0, 2, n)
  )
)

df$exact <- 8 / 3

ggplot(df, aes(x = n)) +
  geom_line(aes(y = estimate, linetype = "trapezium estimate"), linewidth = 1.1) +
  geom_line(aes(y = exact, linetype = "exact value"), linewidth = 1.1) +
  labs(
    title = "Trapezium rule estimates",
    subtitle = "As the number of strips increases, the estimate approaches the exact value.",
    x = "number of strips",
    y = "area estimate",
    linetype = "Quantity"
  ) +
  book_theme()
```

### Exercises 5.5

#### Bronze

1. Use the trapezium rule with $2$ strips to estimate $\int_0^2 x^2\,dx$.
2. Use the trapezium rule with $4$ strips to estimate $\int_0^4 (x+1)\,dx$.
3. Use the trapezium rule with $2$ strips to estimate $\int_1^3 \dfrac{1}{x}\,dx$.
4. Use the trapezium rule with $1$ strip to estimate $\int_0^1 e^x\,dx$.
5. Use the trapezium rule with $2$ strips to estimate $\int_0^{\dfrac{\pi}{2}}\cos x\,dx$.

#### Silver

1. Use the trapezium rule with $4$ strips to estimate $\int_0^2 x^2\,dx$.
2. Use the trapezium rule with $2$ strips to estimate $\int_0^2 \dfrac{1}{x+1}\,dx$.
3. Use the trapezium rule with $2$ strips to estimate $\int_0^1 e^x\,dx$.
4. Use the trapezium rule with $4$ strips to estimate $\int_0^\pi \sin x\,dx$.
5. Write the trapezium rule estimate using $3$ strips for $\int_a^b f(x)\,dx$.

#### Gold

1. Use the trapezium rule with $4$ strips to estimate $\int_0^2 x^2\,dx$, and compare it with the exact value.
2. For $f(x)=x^2+1$, use $2$ strips on $[0,a]$. Find the positive integer $a$ such that the trapezium rule estimate is $5$.
3. Prove that the trapezium rule is exact for any linear function $f(x)=mx+c$.
4. Use the trapezium rule with $4$ strips to estimate $\int_1^2 \dfrac{1}{x}\,dx$.
5. Use the trapezium rule with $3$ strips to estimate $\int_0^{\dfrac{\pi}{2}}\sin x\,dx$.

## STEP-style extension questions

These questions are designed to require proof, structure or exact reasoning.

### STEP-style Question 1

Prove that

$$
\int_0^\pi \sin^2x\,dx=\dfrac{\pi}{2}.
$$

### STEP-style Question 2

Let $k>0$ and $a>0$. Suppose

$$
e^{ka}=4
$$

and

$$
\int_0^a e^{kx}\,dx=3.
$$

Find $k$ and $a$.

### STEP-style Question 3

Use calculus to prove that

$$
\ln(1+a)<a
$$

for all $a>0$.

Then interpret the inequality in terms of the area under

$$
y=\dfrac{1}{1+x}.
$$

### STEP-style Question 4

Find constants $A$ and $B$ such that

$$
F(x)=Ae^{2x}+B\sin3x
$$

is an antiderivative of

$$
6e^{2x}+12\cos3x.
$$

### STEP-style Question 5

For $f(x)=x^2$ on the interval $[0,a]$, use $n$ equal strips.

Prove that the trapezium rule estimate is

$$
\dfrac{a^3}{3}+\dfrac{a^3}{6n^2}.
$$

Hence explain why the trapezium rule overestimates

$$
\int_0^a x^2\,dx.
$$

## Reflection questions

1. Why is integration called reverse differentiation?
2. Why does $\int e^{ax+b}\,dx$ require division by $a$?
3. Why does $\int \dfrac{1}{ax+b}\,dx$ produce a logarithm?
4. Why must trigonometric integrals use radians?
5. Why are identities useful before integrating powers of sine and cosine?
6. What does the trapezium rule estimate geometrically?
7. Why does increasing the number of strips usually improve the trapezium rule estimate?
8. What can R show clearly about area that algebra alone may hide?
9. What can exact integration prove that numerical approximation can only suggest?

## Answers

### Answers to Exercises 5.1

#### Bronze

1. $e^x+C$.
2. $\dfrac{1}{3}e^{3x}+C$.
3. $\dfrac{5}{2}e^{2x}+C$.
4. $-e^{-x}+C$.
5. $-2e^{1-2x}+C$.

#### Silver

1. $e^x+\dfrac{2}{3}e^{3x}+C$.
2. $\dfrac{3}{2}e^{2x-1}+C$.
3. $\dfrac{e^2-1}{2}$.
4. $1$.
5. $a=5$.

#### Gold

1. $e^4-1$.
2. $\dfrac{1}{3}e^{3x+2}+C$.
3. $8$.
4. $k=-\dfrac{1}{2}$.
5. $a=\ln3$.

### Answers to Exercises 5.2

#### Bronze

1. $\ln|x|+C$.
2. $\ln|x+2|+C$.
3. $\dfrac{1}{2}\ln|2x+1|+C$.
4. $\ln|3x-4|+C$.
5. $5\ln|x-1|+C$.

#### Silver

1. $2\ln|2x+3|+C$.
2. $\dfrac{7}{5}\ln|5x-2|+C$.
3. $\ln2$.
4. $\ln\left(\dfrac{7}{3}\right)$.
5. $a=4$ and $b=-3$.

#### Gold

1. $\ln7$.
2. $\ln6$.
3. $a=4$.
4. $\ln|x+1|+\ln|x+2|+C$.
5. $k=2$.

### Answers to Exercises 5.3

#### Bronze

1. $\sin x+C$.
2. $-\cos x+C$.
3. $\tan x+C$.
4. $\dfrac{1}{2}\sin2x+C$.
5. $-\dfrac{1}{3}\cos3x+C$.

#### Silver

1. $2\sin(2x+1)+C$.
2. $-2\cos(3x-2)+C$.
3. $\tan5x+C$.
4. $1$.
5. $1$.

#### Gold

1. $1$.
2. $2$.
3. $2$.
4. $a=4\sqrt3$.
5. $x=\dfrac{\pi}{4}$.

### Answers to Exercises 5.4

#### Bronze

1. $\dfrac{x}{2}-\dfrac{1}{4}\sin2x+C$.
2. $\dfrac{x}{2}+\dfrac{1}{4}\sin2x+C$.
3. $\dfrac{1}{2}\sin^2x+C$.
4. $x+\dfrac{1}{2}\sin2x+C$.
5. $x-\dfrac{1}{2}\sin2x+C$.

#### Silver

1. $\dfrac{\pi}{2}$.
2. $\dfrac{\pi}{2}$.
3. $\dfrac{1}{2}$.
4. $2x+\sin2x+C$.
5. $3x-\dfrac{3}{4}\sin4x+C$.

#### Gold

1. $\dfrac{\pi}{8}-\dfrac{1}{4}$.
2. $\dfrac{\pi}{8}+\dfrac{1}{4}$.
3. $\dfrac{\pi}{2}+1$.
4. $k=10$.
5. $\dfrac{\pi}{16}$.

### Answers to Exercises 5.5

#### Bronze

1. $3$.
2. $12$.
3. $\dfrac{7}{6}$.
4. $\dfrac{1+e}{2}$.
5. $\dfrac{\pi}{8}(1+\sqrt2)$.

#### Silver

1. $\dfrac{11}{4}$.
2. $\dfrac{7}{6}$.
3. $\dfrac{1+2\sqrt e+e}{4}$.
4. $\dfrac{\pi}{4}(1+\sqrt2)$.
5. If $h=\dfrac{b-a}{3}$, then
   $$
   T=\dfrac{h}{2}\left[y_0+2y_1+2y_2+y_3\right].
   $$

#### Gold

1. The trapezium estimate is $\dfrac{11}{4}$. The exact value is $\dfrac{8}{3}$. The estimate is too large by $\dfrac{1}{12}$.
2. $a=2$.
3. On each strip, a linear function is already a straight line, so the trapezium exactly matches the area under the graph. Hence the trapezium rule is exact for linear functions.
4. $\dfrac{1171}{1680}$.
5. $\dfrac{\pi}{12}(2+\sqrt3)$.

### Answers to STEP-style extension questions

1. Since
   $$
   \sin^2x=\dfrac{1-\cos2x}{2},
   $$
   we have
   $$
   \int_0^\pi \sin^2x\,dx
   =
   \int_0^\pi \dfrac{1-\cos2x}{2}\,dx.
   $$
   Therefore,
   $$
   \int_0^\pi \sin^2x\,dx
   =
   \left[\dfrac{x}{2}-\dfrac{1}{4}\sin2x\right]_0^\pi
   =
   \dfrac{\pi}{2}.
   $$

2. Since $e^{ka}=4$,
   $$
   \int_0^a e^{kx}\,dx
   =
   \left[\dfrac{1}{k}e^{kx}\right]_0^a
   =
   \dfrac{4-1}{k}.
   $$
   This is equal to $3$, so $k=1$. Hence $e^a=4$, giving $a=\ln4$.

3. Define
   $$
   g(a)=a-\ln(1+a).
   $$
   Then
   $$
   g'(a)=1-\dfrac{1}{1+a}=\dfrac{a}{1+a}.
   $$
   For $a>0$, $g'(a)>0$, and $g(0)=0$. Therefore $g(a)>0$ for $a>0$, so
   $$
   \ln(1+a)<a.
   $$
   Since
   $$
   \ln(1+a)=\int_0^a \dfrac{1}{1+x}\,dx,
   $$
   this says that the area under $y=\dfrac{1}{1+x}$ from $0$ to $a$ is less than the area of the rectangle of width $a$ and height $1$.

4. We require
   $$
   F'(x)=6e^{2x}+12\cos3x.
   $$
   Since
   $$
   F'(x)=2Ae^{2x}+3B\cos3x,
   $$
   we get
   $$
   2A=6
   $$
   and
   $$
   3B=12.
   $$
   Therefore,
   $$
   A=3
   \quad \text{and} \quad
   B=4.
   $$

5. The strip width is
   $$
   h=\dfrac{a}{n}.
   $$
   The ordinates are
   $$
   y_i=\left(\dfrac{ia}{n}\right)^2.
   $$
   The trapezium estimate is
   $$
   T=
   \dfrac{h}{2}
   \left[
   y_0+2y_1+\cdots+2y_{n-1}+y_n
   \right].
   $$
   Since $y_0=0$ and $y_n=a^2$,
   $$
   T
   =
   \dfrac{a}{2n}
   \left[
   2\sum_{i=1}^{n-1}\left(\dfrac{ia}{n}\right)^2+a^2
   \right].
   $$
   Hence
   $$
   T=
   \dfrac{a^3}{2n}
   \left[
   \dfrac{2}{n^2}\sum_{i=1}^{n-1}i^2+1
   \right].
   $$
   Using
   $$
   \sum_{i=1}^{n-1}i^2=\dfrac{(n-1)n(2n-1)}{6},
   $$
   we obtain
   $$
   T=\dfrac{a^3}{3}+\dfrac{a^3}{6n^2}.
   $$
   But
   $$
   \int_0^a x^2\,dx=\dfrac{a^3}{3}.
   $$
   Therefore the trapezium rule overestimates the exact value by
   $$
   \dfrac{a^3}{6n^2}.
   $$