# Numerical solutions of equations

```{r}
#| label: setup-numerical-methods
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

fixed_point_steps <- function(g, x_start, n_steps) {
  values <- numeric(n_steps + 1)
  values[1] <- x_start
  
  for (i in seq_len(n_steps)) {
    values[i + 1] <- g(values[i])
  }
  
  data.frame(
    iteration = 0:n_steps,
    x_value = values
  )
}

make_cobweb <- function(g, x_start, n_steps) {
  segments <- data.frame(
    step = integer(),
    x = numeric(),
    y = numeric(),
    xend = numeric(),
    yend = numeric()
  )
  
  x_current <- x_start
  
  for (i in seq_len(n_steps)) {
    y_next <- g(x_current)
    
    vertical_segment <- data.frame(
      step = i,
      x = x_current,
      y = x_current,
      xend = x_current,
      yend = y_next
    )
    
    horizontal_segment <- data.frame(
      step = i,
      x = x_current,
      y = y_next,
      xend = y_next,
      yend = y_next
    )
    
    segments <- rbind(segments, vertical_segment, horizontal_segment)
    x_current <- y_next
  }
  
  segments
}
```

## Overview

This chapter develops three connected numerical methods ideas:

1. finding a starting point for a root;
2. improving a solution using iteration;
3. using iterative processes to solve problems involving other areas of mathematics.

The central aim is to understand how equations can be solved approximately when exact algebraic methods are difficult or unavailable.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 6.1 Finding a starting point

### Key mathematical ideas

A root of the equation

$$
f(x)=0
$$

is a value of $x$ where the graph of $y=f(x)$ crosses or touches the $x$-axis.

A common way to locate a starting interval is the **change of sign test**.

If $f$ is continuous on $[a,b]$ and

$$
f(a)f(b)<0,
$$

then there is at least one root of

$$
f(x)=0
$$

between $a$ and $b$.

This works because a continuous graph cannot move from one side of the $x$-axis to the other without crossing it.

However, the test must be used carefully:

- a change of sign guarantees at least one root only when the function is continuous;
- no change of sign does not always mean there is no root;
- a root may touch the axis without crossing it.

### Visualisation using R

```{r}
#| label: fig-sign-change-root
#| fig-cap: "The root of f(x) = x^3 - x - 4 lies between 1 and 2."
#| fig-width: 6
#| fig-height: 4

f <- function(x) x^3 - x - 4

df <- data.frame(
  x = seq(0, 3, length.out = 800)
)

df$y <- f(df$x)

ggplot(df, aes(x = x, y = y)) +
  annotate(
    "rect",
    xmin = 1,
    xmax = 2,
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.15
  ) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  annotate("point", x = 1, y = f(1), size = 3) +
  annotate("point", x = 2, y = f(2), size = 3) +
  annotate("text", x = 1, y = f(1) - 1, label = "f(1) < 0") +
  annotate("text", x = 2, y = f(2) + 1, label = "f(2) > 0") +
  labs(
    title = "Change of sign",
    subtitle = "The shaded interval contains a root because the function changes sign.",
    x = "x",
    y = "f(x)"
  ) +
  book_theme()
```

### Worked example

Show that the equation

$$
x^3-x-4=0
$$

has a root between $1$ and $2$.

Let

$$
f(x)=x^3-x-4.
$$

Then

$$
f(1)=1^3-1-4=-4.
$$

Also,

$$
f(2)=2^3-2-4=2.
$$

Since

$$
f(1)<0
$$

and

$$
f(2)>0,
$$

there is a change of sign between $1$ and $2$.

Therefore, since $f$ is continuous, the equation

$$
x^3-x-4=0
$$

has at least one root between $1$ and $2$.

### Short investigation

Use R to test smaller intervals for the root of

$$
x^3-x-4=0.
$$

```{r}
#| label: tbl-sign-change-table
#| echo: true

f <- function(x) x^3 - x - 4

test_points <- seq(1.6, 1.9, by = 0.05)

data.frame(
  x = test_points,
  f_x = round(f(test_points), 6)
)
```

This table helps us identify a smaller interval containing the root.

### Exercises 6.1

#### Bronze

1. Show that $x^2-3=0$ has a root between $1$ and $2$.
2. Show that $x^3-2x-5=0$ has a root between $2$ and $3$.
3. Show that $e^x-3=0$ has a root between $1$ and $2$.
4. Show that $\ln x+x-2=0$ has a root between $1$ and $2$.
5. Show that $\cos x-x=0$ has a root between $0$ and $1$.

#### Silver

1. Show that $x^3-x-4=0$ has a root between $1.7$ and $1.8$.
2. Show that $x^3-2x-2=0$ has a root between $1.7$ and $1.8$.
3. Show that $e^{-x}-x=0$ has a root between $0.5$ and $0.6$.
4. Show that $\ln x+x-3=0$ has a root between $2$ and $3$.
5. Show that $x^3-6x+2=0$ has roots in each of the intervals $(-3,-2)$, $(0,1)$ and $(2,3)$.

#### Gold

1. Show that $\ln x+x-2=0$ has exactly one root for $x>0$.
2. Show that $x^3-3x+1=0$ has three roots, one in each of the intervals $(-2,-1)$, $(0,1)$ and $(1,2)$.
3. Explain why a change of sign does not guarantee a root if the function is not continuous.
4. Explain why no change of sign does not prove that there is no root.
5. Show that the root of $x^3-x-4=0$ is between $1.7955$ and $1.7965$.

## 6.2 Improving your solution

### Key mathematical ideas

Once we have a starting point or a starting interval, we can improve the approximation using an **iterative formula**.

An iterative formula has the form

$$
x_{n+1}=g(x_n).
$$

This means:

- start with an initial value $x_1$;
- substitute it into the formula to get $x_2$;
- substitute $x_2$ into the formula to get $x_3$;
- continue until the values appear stable to the required accuracy.

If the sequence converges to a value $\alpha$, then

$$
\alpha=g(\alpha).
$$

So $\alpha$ is a fixed point of $g$.

An iteration may:

- converge quickly;
- converge slowly;
- oscillate while converging;
- enter a periodic loop;
- diverge.

A starting point and a rearrangement must therefore be chosen carefully.

### Visualisation using R: convergence

```{r}
#| label: fig-cobweb-convergence
#| fig-cap: "Cobweb diagram for x_{n+1} = (x_n + 4)^(1/3), starting from x_1 = 1.5."
#| fig-width: 6
#| fig-height: 5

g <- function(x) (x + 4)^(1 / 3)

curve_df <- data.frame(
  x = seq(1.2, 2.2, length.out = 800)
)

curve_df$g_x <- g(curve_df$x)

line_df <- data.frame(
  x = seq(1.2, 2.2, length.out = 800)
)

line_df$y <- line_df$x

segments_df <- make_cobweb(g, 1.5, 6)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = g_x), linewidth = 1.1) +
  geom_line(data = line_df, aes(x = x, y = y), linetype = "dashed", linewidth = 1.1) +
  geom_segment(
    data = segments_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.8
  ) +
  labs(
    title = "Convergent fixed-point iteration",
    subtitle = "The cobweb moves towards the fixed point.",
    x = "x_n",
    y = "x_{n+1}"
  ) +
  coord_equal() +
  book_theme()
```

### Worked example

The equation

$$
x^3-x-4=0
$$

can be rearranged as

$$
x=\sqrt[3]{x+4}.
$$

This gives the iterative formula

$$
x_{n+1}=\sqrt[3]{x_n+4}.
$$

Starting with

$$
x_1=1.5,
$$

we obtain

$$
x_2=\sqrt[3]{1.5+4}=1.765174\ldots
$$

Then

$$
x_3=\sqrt[3]{1.765174\ldots+4}=1.793098\ldots
$$

and

$$
x_4=1.795988\ldots
$$

Continuing,

$$
x_5=1.796287\ldots
$$

and

$$
x_6=1.796318\ldots
$$

The values are settling down to

$$
x\approx1.796.
$$

Therefore, the root of

$$
x^3-x-4=0
$$

is approximately

$$
1.796
$$

to three decimal places.

### Why starting point and rearrangement matter

The same equation

$$
x^3-x-4=0
$$

can also be rearranged as

$$
x=x^3-4.
$$

This gives

$$
x_{n+1}=x_n^3-4.
$$

Starting from $x_1=1.5$ gives

$$
x_2=-0.625,
$$

$$
x_3=-4.244140625,
$$

and then the values move rapidly away from the root.

So the rearrangement

$$
x_{n+1}=x_n^3-4
$$

is a poor iteration for this root.

### Visualisation using R: divergence

```{r}
#| label: fig-divergent-iteration
#| fig-cap: "Divergence for x_{n+1} = x_n^3 - 4, starting from x_1 = 1.5."
#| fig-width: 6
#| fig-height: 4

g_diverge <- function(x) x^3 - 4

iteration_df <- fixed_point_steps(g_diverge, 1.5, 4)
iteration_df$x_clipped <- pmax(pmin(iteration_df$x_value, 10), -10)

ggplot(iteration_df, aes(x = iteration, y = x_clipped)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0) +
  labs(
    title = "A divergent iteration",
    subtitle = "Values have been clipped to the interval [-10, 10] for readability.",
    x = "iteration",
    y = "x_n"
  ) +
  book_theme()
```

### Visualisation using R: periodic loop

The formula

$$
x_{n+1}=1-x_n
$$

has fixed point

$$
x=\dfrac{1}{2}.
$$

But if we start from

$$
x_1=0.2,
$$

then

$$
x_2=0.8,
$$

$$
x_3=0.2,
$$

and the sequence loops forever.

```{r}
#| label: fig-periodic-loop
#| fig-cap: "A period-two loop for x_{n+1} = 1 - x_n."
#| fig-width: 6
#| fig-height: 5

g_period <- function(x) 1 - x

curve_df <- data.frame(
  x = seq(0, 1, length.out = 800)
)

curve_df$g_x <- g_period(curve_df$x)

line_df <- data.frame(
  x = seq(0, 1, length.out = 800)
)

line_df$y <- line_df$x

segments_df <- make_cobweb(g_period, 0.2, 6)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = g_x), linewidth = 1.1) +
  geom_line(data = line_df, aes(x = x, y = y), linetype = "dashed", linewidth = 1.1) +
  geom_segment(
    data = segments_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.8
  ) +
  labs(
    title = "Periodic loop",
    subtitle = "The sequence alternates between 0.2 and 0.8.",
    x = "x_n",
    y = "x_{n+1}"
  ) +
  coord_equal() +
  book_theme()
```

### A useful convergence test

Suppose the iteration is

$$
x_{n+1}=g(x_n)
$$

and $\alpha$ is a fixed point.

If

$$
|g'(\alpha)|<1,
$$

then the iteration is usually locally convergent near $\alpha$.

If

$$
|g'(\alpha)|>1,
$$

then the iteration is usually locally divergent near $\alpha$.

For

$$
g(x)=\sqrt[3]{x+4},
$$

we have

$$
g'(x)=\dfrac{1}{3(x+4)^{2/3}}.
$$

Near the root $\alpha\approx1.796$,

$$
|g'(\alpha)|\approx0.103<1,
$$

so the iteration converges.

For

$$
g(x)=x^3-4,
$$

we have

$$
g'(x)=3x^2.
$$

Near $\alpha\approx1.796$,

$$
|g'(\alpha)|\approx9.68>1,
$$

so the iteration diverges.

### Exercises 6.2

#### Bronze

1. Use $x_{n+1}=\sqrt{2+x_n}$ with $x_1=1$ to find $x_2$, $x_3$ and $x_4$.
2. Use $x_{n+1}=\sqrt[3]{x_n+4}$ with $x_1=1.5$ to find $x_2$, $x_3$ and $x_4$.
3. Use $x_{n+1}=\cos x_n$ with $x_1=0.5$ to find $x_2$, $x_3$ and $x_4$.
4. Use $x_{n+1}=e^{-x_n}$ with $x_1=0.5$ to find $x_2$, $x_3$ and $x_4$.
5. Use $x_{n+1}=1-x_n$ with $x_1=0.2$ to find $x_2$, $x_3$, $x_4$ and $x_5$.

#### Silver

1. Use $x_{n+1}=\sqrt{2+x_n}$ with $x_1=1$ to estimate the fixed point to three decimal places.
2. Use $x_{n+1}=\sqrt[3]{x_n+4}$ with $x_1=1.5$ to estimate the root of $x^3-x-4=0$ to three decimal places.
3. Use $x_{n+1}=e^{-x_n}$ with $x_1=0.5$ to estimate the solution of $e^{-x}=x$ to three decimal places.
4. Show that $x_{n+1}=x_n^3-4$ diverges when $x_1=1.5$.
5. Show that $x_{n+1}=\dfrac{2}{x_n}$ enters a periodic loop when $x_1=1$.

#### Gold

1. For the equation $x^3-x-4=0$, compare the two iterations
   $$
   x_{n+1}=\sqrt[3]{x_n+4}
   $$
   and
   $$
   x_{n+1}=x_n^3-4.
   $$
   Explain why one is suitable and the other is not.
2. Prove that the iteration $x_{n+1}=1-x_n$ enters a period-two loop unless $x_1=\dfrac{1}{2}$.
3. Explain why $x_{n+1}=e^{-x_n}$ converges with oscillation near its fixed point.
4. Suppose an iteration $x_{n+1}=g(x_n)$ converges to a value $L$. Prove that $L=g(L)$, assuming $g$ is continuous.
5. Use the iteration
   $$
   x_{n+1}=\dfrac{1}{2}\left(x_n+\dfrac{3}{x_n}\right)
   $$
   with $x_1=2$ to estimate $\sqrt3$ to five decimal places.

## 6.3 Using iterative processes to solve problems involving other areas of mathematics

### Key mathematical ideas

Iteration is not restricted to polynomial equations. It can be used in many areas of mathematics, including:

- exponential equations;
- logarithmic equations;
- trigonometric equations;
- geometry problems leading to transcendental equations;
- equations where exact algebraic methods are unavailable.

For example, the equation

$$
\cos x=x
$$

cannot be solved exactly using elementary algebra.

But it can be written as

$$
x=\cos x,
$$

which gives the iteration

$$
x_{n+1}=\cos x_n.
$$

The solution is the fixed point where the graphs of

$$
y=\cos x
$$

and

$$
y=x
$$

intersect.

### Visualisation using R

```{r}
#| label: fig-cos-fixed-point
#| fig-cap: "Solving cos(x) = x using fixed-point iteration."
#| fig-width: 6
#| fig-height: 5

g_cos <- function(x) cos(x)

curve_df <- data.frame(
  x = seq(0, 1.2, length.out = 800)
)

curve_df$g_x <- g_cos(curve_df$x)

line_df <- data.frame(
  x = seq(0, 1.2, length.out = 800)
)

line_df$y <- line_df$x

segments_df <- make_cobweb(g_cos, 0.5, 8)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = g_x), linewidth = 1.1) +
  geom_line(data = line_df, aes(x = x, y = y), linetype = "dashed", linewidth = 1.1) +
  geom_segment(
    data = segments_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.8
  ) +
  labs(
    title = "Fixed point of cos(x)",
    subtitle = "The iteration x_{n+1} = cos(x_n) converges to approximately 0.739.",
    x = "x_n",
    y = "x_{n+1}"
  ) +
  coord_equal() +
  book_theme()
```

### Worked example

Use iteration to solve

$$
\cos x=x
$$

to three decimal places.

Use

$$
x_{n+1}=\cos x_n
$$

with

$$
x_1=0.5.
$$

Then

$$
x_2=\cos(0.5)=0.877582\ldots
$$

$$
x_3=\cos(0.877582\ldots)=0.639012\ldots
$$

$$
x_4=0.802685\ldots
$$

The values oscillate but slowly converge.

Continuing gives

$$
x\approx0.739.
$$

Therefore, the solution of

$$
\cos x=x
$$

is

$$
x=0.739
$$

to three decimal places.

### A trigonometric example

The equation

$$
\theta=2\sin\theta
$$

has the fixed-point iteration

$$
\theta_{n+1}=2\sin\theta_n.
$$

This type of equation can occur when a geometric or circular problem produces both a linear angle term and a trigonometric term.

```{r}
#| label: fig-theta-two-sin-theta
#| fig-cap: "Solving theta = 2 sin(theta) using iteration."
#| fig-width: 6
#| fig-height: 5

g_theta <- function(theta) 2 * sin(theta)

curve_df <- data.frame(
  theta = seq(0, 2.4, length.out = 800)
)

curve_df$g_theta <- g_theta(curve_df$theta)

line_df <- data.frame(
  theta = seq(0, 2.4, length.out = 800)
)

line_df$y <- line_df$theta

segments_df <- make_cobweb(g_theta, 1.85, 10)

ggplot() +
  geom_line(data = curve_df, aes(x = theta, y = g_theta), linewidth = 1.1) +
  geom_line(data = line_df, aes(x = theta, y = y), linetype = "dashed", linewidth = 1.1) +
  geom_segment(
    data = segments_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    linewidth = 0.8
  ) +
  labs(
    title = "Iteration for theta = 2 sin(theta)",
    subtitle = "The non-zero solution is approximately 1.895.",
    x = "theta_n",
    y = "theta_{n+1}"
  ) +
  coord_equal() +
  book_theme()
```

### Worked example

Use the iteration

$$
\theta_{n+1}=2\sin\theta_n
$$

with

$$
\theta_1=1.85
$$

to solve

$$
\theta=2\sin\theta
$$

to three decimal places.

The first few iterations are

$$
\theta_2=1.922550\ldots,
$$

$$
\theta_3=1.877539\ldots,
$$

$$
\theta_4=1.906644\ldots,
$$

$$
\theta_5=1.888262\ldots.
$$

The sequence oscillates while converging.

Continuing gives

$$
\theta\approx1.895.
$$

To confirm the rounding, define

$$
f(\theta)=\theta-2\sin\theta.
$$

Then

$$
f(1.8945)<0
$$

and

$$
f(1.8955)>0.
$$

Therefore, the non-zero solution is

$$
\theta=1.895
$$

to three decimal places.

### Short investigation

Compare four different iterative behaviours.

```{r}
#| label: tbl-iteration-behaviours
#| echo: true

iterations <- data.frame(
  behaviour = c(
    "converges quickly",
    "oscillates but converges",
    "periodic loop",
    "diverges"
  ),
  iteration = c(
    "x_{n+1} = (x_n + 4)^(1/3)",
    "x_{n+1} = cos(x_n)",
    "x_{n+1} = 1 - x_n",
    "x_{n+1} = x_n^3 - 4"
  ),
  starting_value = c(1.5, 0.5, 0.2, 1.5)
)

iterations
```

### Exercises 6.3

#### Bronze

1. Use $x_{n+1}=\cos x_n$ with $x_1=0.5$ to solve $\cos x=x$ to three decimal places.
2. Use $x_{n+1}=e^{-x_n}$ with $x_1=0.5$ to solve $e^{-x}=x$ to three decimal places.
3. Use $\theta_{n+1}=2\sin\theta_n$ with $\theta_1=1.85$ to estimate the non-zero solution of $\theta=2\sin\theta$.
4. Use $x_{n+1}=\ln(x_n+2)$ with $x_1=1$ to solve $\ln(x+2)=x$ to three decimal places.
5. The iteration $x_{n+1}=\sqrt{x_n+6}$ has a positive fixed point. Find the equation satisfied by this fixed point and hence find its value.

#### Silver

1. Use a suitable iteration to solve $x^3-x-4=0$ to three decimal places.
2. Use $\theta_{n+1}=2\sin\theta_n$ to solve $\theta=2\sin\theta$ to three decimal places.
3. Explain why the iteration $x_{n+1}=\dfrac{2}{x_n}$ is not suitable for finding $\sqrt2$ when $x_1=1$.
4. Explain why the iteration $x_{n+1}=2x_n-1$ is not suitable for finding its fixed point unless the starting value is already correct.
5. Use $x_{n+1}=\ln(x_n+2)$ with $x_1=1$ to solve $\ln(x+2)=x$ to three decimal places.

#### Gold

1. Prove that $\cos x=x$ has exactly one solution in the interval $(0,1)$.
2. Show that the iteration $x_{n+1}=e^{-x_n}$ converges with oscillation near its fixed point.
3. Prove that $x_{n+1}=1-x_n$ enters a period-two loop unless $x_1=\dfrac{1}{2}$.
4. Use a change of sign test to confirm that the root of $x^3-x-4=0$ is $1.796$ to three decimal places.
5. Use a change of sign test to confirm that the non-zero solution of $\theta=2\sin\theta$ is $1.895$ to three decimal places.

## STEP-style extension questions

These questions are designed to require proof, structure or exact reasoning.

### STEP-style Question 1

Let $f$ be continuous on $[a,b]$.

Suppose

$$
f(a)<0
$$

and

$$
f(b)>0.
$$

Explain why there must be at least one value $\alpha\in(a,b)$ such that

$$
f(\alpha)=0.
$$

Then give an example showing why continuity is needed.

### STEP-style Question 2

Suppose $g$ maps the interval $[a,b]$ into itself and there is a constant $k$ such that

$$
0<k<1
$$

and

$$
|g'(x)|\leq k
$$

for every $x\in[a,b]$.

Let $\alpha$ be a fixed point of $g$ in $[a,b]$.

Show that the iteration

$$
x_{n+1}=g(x_n)
$$

satisfies

$$
|x_{n+1}-\alpha|\leq k|x_n-\alpha|.
$$

Hence explain why the iteration converges to $\alpha$.

### STEP-style Question 3

The equation

$$
x^2=2
$$

can be written as

$$
x=\dfrac{2}{x}.
$$

This gives the iteration

$$
x_{n+1}=\dfrac{2}{x_n}.
$$

Show that if $x_1\neq\sqrt2$ and $x_1\neq-\sqrt2$, then the sequence does not converge but enters a period-two loop.

### STEP-style Question 4

The equation

$$
x^3-x-4=0
$$

can be rearranged in two ways:

$$
x=\sqrt[3]{x+4}
$$

and

$$
x=x^3-4.
$$

Let $\alpha$ be the root between $1$ and $2$.

Use derivatives of the two iteration functions to explain why

$$
x_{n+1}=\sqrt[3]{x_n+4}
$$

converges near $\alpha$, while

$$
x_{n+1}=x_n^3-4
$$

diverges near $\alpha$.

### STEP-style Question 5

Let

$$
f(x)=\cos x-x.
$$

Use calculus and a change of sign argument to prove that the equation

$$
\cos x=x
$$

has exactly one solution in $[0,1]$.

Then explain why the iteration

$$
x_{n+1}=\cos x_n
$$

converges with oscillation when started near the solution.

## Reflection questions

1. Why is a starting point needed before using an iterative method?
2. What does a change of sign tell us about a continuous function?
3. Why does no change of sign not always rule out a root?
4. What is a fixed point of an iteration?
5. Why can two rearrangements of the same equation behave differently?
6. What does a cobweb diagram show about convergence?
7. How can a starting point cause a periodic loop?
8. Why might an iteration diverge even though it has the correct fixed point?
9. What can R show clearly about iteration that numerical tables may hide?
10. What can exact reasoning prove that numerical iteration can only suggest?

## Answers

### Answers to Exercises 6.1

#### Bronze

1. Let $f(x)=x^2-3$. Then $f(1)=-2$ and $f(2)=1$, so there is a root between $1$ and $2$.
2. Let $f(x)=x^3-2x-5$. Then $f(2)=-1$ and $f(3)=16$, so there is a root between $2$ and $3$.
3. Let $f(x)=e^x-3$. Then $f(1)=e-3<0$ and $f(2)=e^2-3>0$, so there is a root between $1$ and $2$.
4. Let $f(x)=\ln x+x-2$. Then $f(1)=-1$ and $f(2)=\ln2>0$, so there is a root between $1$ and $2$.
5. Let $f(x)=\cos x-x$. Then $f(0)=1$ and $f(1)=\cos1-1<0$, so there is a root between $0$ and $1$.

#### Silver

1. For $f(x)=x^3-x-4$, $f(1.7)=-0.787$ and $f(1.8)=0.032$, so there is a root between $1.7$ and $1.8$.
2. For $f(x)=x^3-2x-2$, $f(1.7)=-0.487$ and $f(1.8)=0.232$, so there is a root between $1.7$ and $1.8$.
3. For $f(x)=e^{-x}-x$, $f(0.5)=0.1065\ldots$ and $f(0.6)=-0.0512\ldots$, so there is a root between $0.5$ and $0.6$.
4. For $f(x)=\ln x+x-3$, $f(2)=\ln2-1<0$ and $f(3)=\ln3>0$, so there is a root between $2$ and $3$.
5. For $f(x)=x^3-6x+2$, there are sign changes on $(-3,-2)$, $(0,1)$ and $(2,3)$, so each interval contains a root.

#### Gold

1. Let $f(x)=\ln x+x-2$. Since
   $$
   f'(x)=\dfrac{1}{x}+1>0
   $$
   for $x>0$, $f$ is strictly increasing. Since $f(1)<0$ and $f(2)>0$, there is exactly one root for $x>0$.
2. Let $f(x)=x^3-3x+1$. We have sign changes on $(-2,-1)$, $(0,1)$ and $(1,2)$, so the equation has roots in all three intervals.
3. For example,
   $$
   f(x)=\dfrac{1}{x-1}
   $$
   has $f(0)<0$ and $f(2)>0$, but it has no root between $0$ and $2$. The sign change occurs because of a discontinuity.
4. For example,
   $$
   f(x)=(x-1)^2
   $$
   has no sign change between $0$ and $2$, but it has a root at $x=1$.
5. For $f(x)=x^3-x-4$, $f(1.7955)<0$ and $f(1.7965)>0$, so the root lies between $1.7955$ and $1.7965$. Therefore the root is $1.796$ to three decimal places.

### Answers to Exercises 6.2

#### Bronze

1. $x_2=1.7321$, $x_3=1.9319$, $x_4=1.9829$.
2. $x_2=1.7652$, $x_3=1.7931$, $x_4=1.7960$.
3. $x_2=0.8776$, $x_3=0.6390$, $x_4=0.8027$.
4. $x_2=0.6065$, $x_3=0.5452$, $x_4=0.5797$.
5. $x_2=0.8$, $x_3=0.2$, $x_4=0.8$, $x_5=0.2$.

#### Silver

1. The iteration converges to $x=2.000$.
2. The root is $x=1.796$ to three decimal places.
3. The solution is $x=0.567$ to three decimal places.
4. Starting with $x_1=1.5$, the iteration gives $x_2=-0.625$, $x_3=-4.2441\ldots$ and $x_4=-80.4486\ldots$, so it diverges.
5. Starting with $x_1=1$, the iteration gives $x_2=2$, $x_3=1$, $x_4=2$, and so on. It enters a period-two loop.

#### Gold

1. For $g_1(x)=\sqrt[3]{x+4}$,
   $$
   |g_1'(\alpha)|\approx0.103<1,
   $$
   so the iteration is locally convergent. For $g_2(x)=x^3-4$,
   $$
   |g_2'(\alpha)|\approx9.68>1,
   $$
   so the iteration is locally divergent.
2. If $x_{n+1}=1-x_n$, then
   $$
   x_{n+2}=1-x_{n+1}=1-(1-x_n)=x_n.
   $$
   Therefore the sequence repeats every two terms unless $x_n=\dfrac{1}{2}$.
3. At the fixed point $\alpha$, $e^{-\alpha}=\alpha$. For $g(x)=e^{-x}$,
   $$
   g'(\alpha)=-e^{-\alpha}=-\alpha\approx-0.567.
   $$
   Since $|g'(\alpha)|<1$, the iteration converges. Since $g'(\alpha)<0$, it oscillates while converging.
4. If $x_n\to L$ and $x_{n+1}=g(x_n)$, then taking limits gives
   $$
   L=g(L),
   $$
   provided $g$ is continuous.
5. The iteration gives
   $$
   x_2=1.75,
   $$
   $$
   x_3=1.732142857\ldots,
   $$
   $$
   x_4=1.73205081\ldots.
   $$
   Hence
   $$
   \sqrt3\approx1.73205.
   $$

### Answers to Exercises 6.3

#### Bronze

1. $x=0.739$ to three decimal places.
2. $x=0.567$ to three decimal places.
3. The non-zero solution is $\theta=1.895$ to three decimal places.
4. $x=1.146$ to three decimal places.
5. A fixed point satisfies
   $$
   x=\sqrt{x+6}.
   $$
   Squaring gives
   $$
   x^2=x+6,
   $$
   so
   $$
   x^2-x-6=0.
   $$
   The positive fixed point is $x=3$.

#### Silver

1. A suitable iteration is
   $$
   x_{n+1}=\sqrt[3]{x_n+4}.
   $$
   Starting near the root gives $x=1.796$ to three decimal places.
2. The non-zero solution is $\theta=1.895$ to three decimal places.
3. Starting with $x_1=1$, the iteration gives $1,2,1,2,\ldots$, so it does not converge to $\sqrt2$.
4. The fixed point satisfies $x=2x-1$, so $x=1$. But if $x_1\neq1$, then the distance from $1$ doubles each time, so the iteration diverges.
5. The solution is $x=1.146$ to three decimal places.

#### Gold

1. Let $f(x)=\cos x-x$. Then
   $$
   f'(x)=-\sin x-1<0
   $$
   on $(0,1)$. So $f$ is strictly decreasing. Since $f(0)>0$ and $f(1)<0$, there is exactly one root in $(0,1)$.
2. For $g(x)=e^{-x}$, at the fixed point $\alpha$,
   $$
   g'(\alpha)=-e^{-\alpha}=-\alpha.
   $$
   Since $\alpha\approx0.567$, we have $|g'(\alpha)|<1$, so the iteration converges. The negative derivative explains the oscillation.
3. Since
   $$
   x_{n+2}=1-x_{n+1}=1-(1-x_n)=x_n,
   $$
   the sequence repeats every two terms. It is constant only if $x_n=1-x_n$, giving $x_n=\dfrac{1}{2}$.
4. For $f(x)=x^3-x-4$, $f(1.7955)<0$ and $f(1.7965)>0$, so the root lies between $1.7955$ and $1.7965$. Therefore the root is $1.796$ to three decimal places.
5. For $f(\theta)=\theta-2\sin\theta$, $f(1.8945)<0$ and $f(1.8955)>0$, so the non-zero solution lies between $1.8945$ and $1.8955$. Therefore
   $$
   \theta=1.895
   $$
   to three decimal places.

### Answers to STEP-style extension questions

1. Since $f$ is continuous, its graph cannot move from a negative value at $x=a$ to a positive value at $x=b$ without crossing the $x$-axis. Therefore there must be some $\alpha\in(a,b)$ such that
   $$
   f(\alpha)=0.
   $$
   Continuity is needed. For example,
   $$
   f(x)=\dfrac{1}{x-1}
   $$
   changes sign between $0$ and $2$, but has no root in that interval because it is discontinuous at $x=1$.

2. By the mean value theorem, for some point $c$ between $x_n$ and $\alpha$,
   $$
   g(x_n)-g(\alpha)=g'(c)(x_n-\alpha).
   $$
   Since $g(\alpha)=\alpha$,
   $$
   x_{n+1}-\alpha=g'(c)(x_n-\alpha).
   $$
   Taking absolute values gives
   $$
   |x_{n+1}-\alpha|
   =
   |g'(c)||x_n-\alpha|
   \leq
   k|x_n-\alpha|.
   $$
   Since $0<k<1$, the error shrinks at each step, so the iteration converges to $\alpha$.

3. The iteration is
   $$
   x_{n+1}=\dfrac{2}{x_n}.
   $$
   Then
   $$
   x_{n+2}=\dfrac{2}{x_{n+1}}
   =
   \dfrac{2}{\dfrac{2}{x_n}}
   =
   x_n.
   $$
   Hence the sequence repeats every two terms. It is constant only when
   $$
   x_n=\dfrac{2}{x_n},
   $$
   giving
   $$
   x_n^2=2.
   $$
   Therefore the sequence converges only if the starting value is already $\sqrt2$ or $-\sqrt2$.

4. Let
   $$
   g_1(x)=\sqrt[3]{x+4}.
   $$
   Then
   $$
   g_1'(x)=\dfrac{1}{3(x+4)^{2/3}}.
   $$
   At the root $\alpha\approx1.796$,
   $$
   |g_1'(\alpha)|\approx0.103<1.
   $$
   So the iteration
   $$
   x_{n+1}=\sqrt[3]{x_n+4}
   $$
   converges near $\alpha$.

   Now let
   $$
   g_2(x)=x^3-4.
   $$
   Then
   $$
   g_2'(x)=3x^2.
   $$
   At $\alpha\approx1.796$,
   $$
   |g_2'(\alpha)|\approx9.68>1.
   $$
   Therefore the iteration
   $$
   x_{n+1}=x_n^3-4
   $$
   diverges near $\alpha$.

5. Let
   $$
   f(x)=\cos x-x.
   $$
   Then
   $$
   f(0)=1>0
   $$
   and
   $$
   f(1)=\cos1-1<0.
   $$
   So there is at least one root in $(0,1)$.

   Also,
   $$
   f'(x)=-\sin x-1<0
   $$
   for $0\leq x\leq1$.
   Therefore $f$ is strictly decreasing and can cross the $x$-axis only once. Hence the solution is unique.

   The iteration
   $$
   x_{n+1}=\cos x_n
   $$
   has iteration function $g(x)=\cos x$. At the fixed point $\alpha$,
   $$
   g'(\alpha)=-\sin\alpha.
   $$
   Since $0<\sin\alpha<1$, we have
   $$
   |g'(\alpha)|<1.
   $$
   Therefore the iteration converges near $\alpha$. Since $g'(\alpha)<0$, the convergence is oscillatory.