# Further algebra

```{r}
#| label: setup-further-algebra
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

general_binom <- function(power, k_value) {
  if (k_value == 0) {
    return(1)
  }
  prod(power - 0:(k_value - 1)) / factorial(k_value)
}

binomial_approx <- function(x_value, power, highest_power) {
  total <- rep(0, length(x_value))
  for (k_value in 0:highest_power) {
    total <- total + general_binom(power, k_value) * x_value^k_value
  }
  total
}
```

## Overview

This chapter develops five connected algebraic ideas:

1. improper algebraic fractions;
2. partial fractions;
3. binomial expansion of $(1+x)^n$ for values of $n$ that are not positive integers;
4. binomial expansion of $(a+x)^n$ for values of $n$ that are not positive integers;
5. using partial fractions together with binomial expansions.

The central aim is to rewrite algebraic expressions into forms that are easier to expand, approximate, integrate or sum.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 7.1 Improper algebraic fractions

### Key mathematical ideas

An algebraic fraction is **proper** when

$$
\deg(\text{numerator})<\deg(\text{denominator}).
$$

It is **improper** when

$$
\deg(\text{numerator})\geq\deg(\text{denominator}).
$$

If an algebraic fraction is improper, do long division first.

The division algorithm says that

$$
P(x)=D(x)Q(x)+R(x),
$$

where $Q(x)$ is the quotient and $R(x)$ is the remainder.

Therefore,

$$
\dfrac{P(x)}{D(x)}
=
Q(x)+\dfrac{R(x)}{D(x)}.
$$

The fraction

$$
\dfrac{R(x)}{D(x)}
$$

is then proper.

::: {.callout-warning}
Before using partial fractions, always check the degrees. If the power of the top is greater than or equal to the power of the bottom, do long division first.
:::

### Visualisation using R

```{r}
#| label: fig-improper-fraction-decomposition
#| fig-cap: "An improper algebraic fraction written as a polynomial plus a proper fraction."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-4, 4, length.out = 1000)
)

df <- subset(df, abs(x^2 + 1) > 0.05)

df$original <- (2 * df$x^3 + 3 * df$x^2 - df$x + 5) / (df$x^2 + 1)
df$decomposed <- 2 * df$x + 3 + (-3 * df$x + 2) / (df$x^2 + 1)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = original, linetype = "original fraction"), linewidth = 1.1) +
  geom_line(aes(y = decomposed, linetype = "polynomial plus proper fraction"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  labs(
    title = "Checking an improper algebraic fraction",
    subtitle = "The two graphs overlap because the two expressions are identical.",
    x = "x",
    y = "value",
    linetype = "Expression"
  ) +
  book_theme()
```

### Worked example

Express

$$
\dfrac{2x^3+3x^2-x+5}{x^2+1}
$$

as the sum of a polynomial and a proper fraction.

Since the degree of the numerator is greater than the degree of the denominator, we divide first.

First,

$$
\dfrac{2x^3}{x^2}=2x.
$$

Multiply:

$$
2x(x^2+1)=2x^3+2x.
$$

Subtract:

$$
2x^3+3x^2-x+5-(2x^3+2x)=3x^2-3x+5.
$$

Next,

$$
\dfrac{3x^2}{x^2}=3.
$$

Multiply:

$$
3(x^2+1)=3x^2+3.
$$

Subtract:

$$
3x^2-3x+5-(3x^2+3)=-3x+2.
$$

Therefore,

$$
2x^3+3x^2-x+5
=
(x^2+1)(2x+3)+(-3x+2).
$$

Hence,

$$
\dfrac{2x^3+3x^2-x+5}{x^2+1}
=
2x+3+\dfrac{-3x+2}{x^2+1}.
$$

### Short investigation

Change the numerator below and compare the original fraction with its decomposed form.

```{r}
#| label: fig-improper-fraction-family
#| fig-cap: "Checking a long-division decomposition graphically."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-5, 5, length.out = 1000)
)

df <- subset(df, abs(x - 2) > 0.05)

df$original <- (df$x^3 - 4 * df$x + 1) / (df$x - 2)
df$decomposed <- df$x^2 + 2 * df$x + 1 / (df$x - 2)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = original, linetype = "original"), linewidth = 1.1) +
  geom_line(aes(y = decomposed, linetype = "decomposed"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 2, linetype = "dashed") +
  coord_cartesian(ylim = c(-20, 20)) +
  labs(
    title = "Long division check",
    subtitle = "The vertical line marks the value excluded from the domain.",
    x = "x",
    y = "value",
    linetype = "Expression"
  ) +
  book_theme()
```

### Exercises 7.1

#### Bronze

1. Express $\dfrac{x^2+5x+6}{x+2}$ as a polynomial plus a proper fraction.
2. Express $\dfrac{x^2-1}{x-1}$ as a polynomial plus a proper fraction.
3. Express $\dfrac{x^3+2x^2+3x+4}{x+1}$ as a polynomial plus a proper fraction.
4. Express $\dfrac{2x^3+x^2-5}{x+2}$ as a polynomial plus a proper fraction.
5. Express $\dfrac{x^4-1}{x^2+1}$ as a polynomial plus a proper fraction.

#### Silver

1. Express $\dfrac{x^3-4x+1}{x-2}$ as a polynomial plus a proper fraction.
2. Express $\dfrac{3x^3-2x^2+x-7}{x+1}$ as a polynomial plus a proper fraction.
3. Express $\dfrac{x^4+2x^2+1}{x^2+2x+1}$ as a polynomial plus a proper fraction.
4. Express $\dfrac{2x^4-x^3+3x^2-x+5}{x^2-1}$ as a polynomial plus a proper fraction.
5. Express $\dfrac{x^5+x+1}{x^2+x+1}$ as a polynomial plus a proper fraction.

#### Gold

1. Express $\dfrac{x^3+1}{x^2+x+1}$ as a polynomial plus a proper fraction.
2. Express $\dfrac{3x^4+2x^3-x+7}{x^2+x-1}$ as a polynomial plus a proper fraction.
3. Explain why $\dfrac{x^3+2x+1}{x^2+1}$ must be divided before using partial fractions.
4. Given that
   $$
   \dfrac{x^3+ax^2+bx+5}{x+1}
   =
   x^2+2x-1+\dfrac{6}{x+1},
   $$
   find $a$ and $b$.
5. Show that
   $$
   \dfrac{x^4-1}{x^2+1}=x^2-1.
   $$

## 7.2 Partial fractions

### Key mathematical ideas

Partial fractions reverse the process of adding algebraic fractions.

This chapter uses three main proper-fraction types.

### Type 1: distinct linear factors

If the denominator has distinct linear factors, use one constant numerator over each factor.

For example,

$$
\dfrac{5x+1}{(x-1)(x+2)}
=
\dfrac{A}{x-1}+\dfrac{B}{x+2}.
$$

### Type 2: repeated linear factors

If the denominator has a repeated linear factor, include each power of that factor.

For example,

$$
\dfrac{3x+5}{(x-1)^2}
=
\dfrac{A}{x-1}+\dfrac{B}{(x-1)^2}.
$$

For a denominator such as $(x-1)^3$, we would use

$$
\dfrac{A}{x-1}+\dfrac{B}{(x-1)^2}+\dfrac{C}{(x-1)^3}.
$$

### Type 3: an irreducible quadratic factor

If the denominator contains a quadratic factor that cannot be factorised over the real numbers, use a linear numerator over that quadratic.

For example,

$$
\dfrac{7x-1}{(x-1)(x^2+2)}
=
\dfrac{A}{x-1}+\dfrac{Bx+C}{x^2+2}.
$$

### Worked example

Express

$$
\dfrac{7x-1}{(x-1)(x^2+2)}
$$

in partial fractions.

Write

$$
\dfrac{7x-1}{(x-1)(x^2+2)}
=
\dfrac{A}{x-1}+\dfrac{Bx+C}{x^2+2}.
$$

Multiplying throughout by $(x-1)(x^2+2)$ gives

$$
7x-1=A(x^2+2)+(Bx+C)(x-1).
$$

Expanding,

$$
7x-1
=
A x^2+2A+Bx^2-Bx+Cx-C.
$$

Collecting terms,

$$
7x-1
=
(A+B)x^2+(C-B)x+(2A-C).
$$

Equating coefficients gives

$$
A+B=0,
$$

$$
C-B=7,
$$

and

$$
2A-C=-1.
$$

Solving gives

$$
A=2,\quad B=-2,\quad C=5.
$$

Therefore,

$$
\dfrac{7x-1}{(x-1)(x^2+2)}
=
\dfrac{2}{x-1}+\dfrac{-2x+5}{x^2+2}.
$$

### Visualisation using R

```{r}
#| label: fig-partial-fractions-check
#| fig-cap: "Checking a partial-fraction decomposition."
#| fig-width: 6
#| fig-height: 4

df <- data.frame(
  x = seq(-4, 4, length.out = 1200)
)

df <- subset(df, abs(x - 1) > 0.05)

df$original <- (7 * df$x - 1) / ((df$x - 1) * (df$x^2 + 2))
df$decomposed <- 2 / (df$x - 1) + (-2 * df$x + 5) / (df$x^2 + 2)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = original, linetype = "original"), linewidth = 1.1) +
  geom_line(aes(y = decomposed, linetype = "partial fractions"), linewidth = 1.1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  coord_cartesian(ylim = c(-12, 12)) +
  labs(
    title = "Partial fractions",
    subtitle = "The two graphs overlap away from the excluded value x = 1.",
    x = "x",
    y = "value",
    linetype = "Expression"
  ) +
  book_theme()
```

### Short investigation

Compare the three partial-fraction templates.

```{r}
#| label: tbl-partial-fraction-types
#| echo: true

data.frame(
  type = c(
    "distinct linear factors",
    "repeated linear factor",
    "irreducible quadratic factor"
  ),
  example_denominator = c(
    "(x - 1)(x + 2)",
    "(x - 1)^2",
    "(x - 1)(x^2 + 2)"
  ),
  template = c(
    "A/(x - 1) + B/(x + 2)",
    "A/(x - 1) + B/(x - 1)^2",
    "A/(x - 1) + (Bx + C)/(x^2 + 2)"
  )
)
```

### Exercises 7.2

#### Bronze

1. Express $\dfrac{5x+1}{(x-1)(x+2)}$ in partial fractions.
2. Express $\dfrac{7x+1}{(x+1)(x+3)}$ in partial fractions.
3. Express $\dfrac{3x+5}{(x-1)^2}$ in partial fractions.
4. Express $\dfrac{4x-1}{(x+2)^2}$ in partial fractions.
5. Express $\dfrac{7x-1}{(x-1)(x^2+2)}$ in partial fractions.

#### Silver

1. Express $\dfrac{2x+5}{(x-2)(x+1)}$ in partial fractions.
2. Express $\dfrac{2x+1}{(x+2)^2}$ in partial fractions.
3. Express $\dfrac{3x^2+x+3}{(x+1)(x^2+4)}$ in partial fractions.
4. Express $\dfrac{1}{x(x+1)}$ in partial fractions.
5. Express $\dfrac{x^2+5x+7}{(x+1)(x+2)}$ in partial fractions.

#### Gold

1. Express $\dfrac{4x+8}{(x+1)^2(x+3)}$ in partial fractions.
2. Express $\dfrac{2x^2+4x-1}{(x-2)(x^2+1)}$ in partial fractions.
3. Explain why $\dfrac{x^2+5x+7}{(x+1)(x+2)}$ needs long division first.
4. Show that
   $$
   \dfrac{1}{r(r+1)}=\dfrac{1}{r}-\dfrac{1}{r+1}.
   $$
   Hence find
   $$
   \sum_{r=1}^{n}\dfrac{1}{r(r+1)}.
   $$
5. Suppose
   $$
   \dfrac{kx+5}{(x-1)(x+2)}
   =
   \dfrac{A}{x-1}+\dfrac{B}{x+2}
   $$
   and $A=2B$. Find $k$.

## 7.3 Binomial expansion of $(1+x)^n$ for values of $n$ that are not positive integers

### Key mathematical ideas

For any rational number $n$,

$$
(1+x)^n
=
1+nx+\dfrac{n(n-1)}{2!}x^2
+
\dfrac{n(n-1)(n-2)}{3!}x^3
+\cdots
$$

provided

$$
|x|<1.
$$

When $n$ is not a non-negative integer, the expansion does not usually terminate. The more terms we include, the better the approximation tends to be inside the interval of validity.

### Visualisation using R

```{r}
#| label: fig-binomial-approximation-one-plus-x
#| fig-cap: "Approximating (1 + x)^n using more binomial terms."
#| fig-width: 6
#| fig-height: 4

power <- -0.5

x_values <- seq(-0.9, 0.9, length.out = 800)

exact_df <- data.frame(
  x = x_values,
  y_exact = (1 + x_values)^power
)

approx_df <- expand.grid(
  x = x_values,
  highest_power = c(1, 2, 3, 4)
)

approx_df$terms_label <- paste0("up to x^", approx_df$highest_power)
approx_df$y_value <- mapply(
  function(x_input, highest_power_input) {
    binomial_approx(x_input, power, highest_power_input)
  },
  approx_df$x,
  approx_df$highest_power
)

ggplot() +
  geom_line(
    data = exact_df,
    aes(x = x, y = y_exact, linetype = "exact"),
    linewidth = 1.2
  ) +
  geom_line(
    data = approx_df,
    aes(x = x, y = y_value, linetype = terms_label),
    linewidth = 0.9
  ) +
  coord_cartesian(ylim = c(0, 4)) +
  labs(
    title = "Binomial approximation for (1 + x)^n",
    subtitle = "Here n = -0.5. Extra terms improve the approximation inside -1 < x < 1.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Expand

$$
(1+x)^{-2}
$$

up to and including the term in $x^3$.

Here,

$$
n=-2.
$$

So

$$
(1+x)^{-2}
=
1+(-2)x
+
\dfrac{(-2)(-3)}{2!}x^2
+
\dfrac{(-2)(-3)(-4)}{3!}x^3
+\cdots
$$

Therefore,

$$
(1+x)^{-2}
=
1-2x+3x^2-4x^3+\cdots.
$$

The expansion is valid for

$$
|x|<1.
$$

### Short investigation

Use R to compare approximations for several powers.

```{r}
#| label: fig-binomial-different-powers
#| fig-cap: "Different rational powers give different binomial approximation behaviour."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-0.8, 0.8, length.out = 600)

df <- expand.grid(
  x = x_values,
  power_label = c("n = 1/2", "n = -1", "n = -2")
)

df$power <- ifelse(df$power_label == "n = 1/2", 0.5, ifelse(df$power_label == "n = -1", -1, -2))

df$exact <- (1 + df$x)^df$power
df$approximation <- mapply(
  function(x_input, power_input) {
    binomial_approx(x_input, power_input, 3)
  },
  df$x,
  df$power
)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = exact, linetype = "exact"), linewidth = 1.1) +
  geom_line(aes(y = approximation, linetype = "up to x^3"), linewidth = 1.1) +
  facet_wrap(~ power_label) +
  labs(
    title = "Binomial approximations up to x^3",
    subtitle = "The approximation is most reliable near x = 0.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Exercises 7.3

#### Bronze

1. Expand $(1+x)^{\frac{1}{2}}$ up to and including the term in $x^2$.
2. Expand $(1+x)^{-1}$ up to and including the term in $x^3$.
3. Expand $(1+x)^{-2}$ up to and including the term in $x^3$.
4. Expand $(1+x)^{\frac{3}{2}}$ up to and including the term in $x^2$.
5. State the range of validity for the expansion of $(1+x)^n$ when $n$ is not a non-negative integer.

#### Silver

1. Expand $(1-2x)^{-1}$ up to and including the term in $x^3$. State the range of validity.
2. Expand $(1+3x)^{\frac{1}{2}}$ up to and including the term in $x^3$.
3. Expand $(1-4x)^{-\frac{1}{2}}$ up to and including the term in $x^2$.
4. Use the binomial expansion of $(1+x)^{\frac{1}{2}}$ to estimate $\sqrt{1.04}$.
5. Expand $(1+x)^{-\frac{1}{2}}$ up to and including the term in $x^3$.

#### Gold

1. Show that the coefficient of $x^3$ in $(1+x)^n$ is
   $$
   \dfrac{n(n-1)(n-2)}{6}.
   $$
2. Find the coefficient of $x^3$ in $(1-2x)^{\frac{1}{2}}$.
3. Expand $(1+x)^{\frac{1}{3}}(1-x)^{-1}$ up to and including the term in $x^3$.
4. Find $k$ if the coefficient of $x^2$ in $(1+kx)^{-\frac{1}{2}}$ is $6$.
5. Find the range of validity for the expansion of $(1-2x)^{\frac{1}{2}}$ and explain why substituting $x=0.6$ is invalid.

## 7.4 Binomial expansion of $(a+x)^n$ for values of $n$ that are not positive integers

### Key mathematical ideas

To expand

$$
(a+x)^n,
$$

factor out $a$ first:

$$
(a+x)^n
=
a^n\left(1+\dfrac{x}{a}\right)^n.
$$

Then apply the binomial expansion to

$$
\left(1+\dfrac{x}{a}\right)^n.
$$

The expansion is valid when

$$
\left|\dfrac{x}{a}\right|<1.
$$

Equivalently,

$$
|x|<|a|.
$$

### Visualisation using R

```{r}
#| label: fig-binomial-approximation-a-plus-x
#| fig-cap: "Approximating sqrt(4 + x) using binomial expansions."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-3.5, 3.5, length.out = 800)

exact_df <- data.frame(
  x = x_values,
  y_exact = sqrt(4 + x_values)
)

approx_df <- expand.grid(
  x = x_values,
  highest_power = c(1, 2, 3)
)

approx_df$terms_label <- paste0("up to x^", approx_df$highest_power)
approx_df$y_value <- 2 * mapply(
  function(x_input, highest_power_input) {
    binomial_approx(x_input / 4, 0.5, highest_power_input)
  },
  approx_df$x,
  approx_df$highest_power
)

ggplot() +
  geom_line(
    data = exact_df,
    aes(x = x, y = y_exact, linetype = "exact"),
    linewidth = 1.2
  ) +
  geom_line(
    data = approx_df,
    aes(x = x, y = y_value, linetype = terms_label),
    linewidth = 0.9
  ) +
  labs(
    title = "Binomial approximation for sqrt(4 + x)",
    subtitle = "The expansion is valid for |x| < 4.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Expand

$$
(4+x)^{\frac{1}{2}}
$$

up to and including the term in $x^3$.

First write

$$
(4+x)^{\frac{1}{2}}
=
4^{\frac{1}{2}}\left(1+\dfrac{x}{4}\right)^{\frac{1}{2}}.
$$

So

$$
(4+x)^{\frac{1}{2}}
=
2\left(1+\dfrac{x}{4}\right)^{\frac{1}{2}}.
$$

Now

$$
(1+u)^{\frac{1}{2}}
=
1+\dfrac{1}{2}u-\dfrac{1}{8}u^2+\dfrac{1}{16}u^3+\cdots.
$$

Using

$$
u=\dfrac{x}{4},
$$

we get

$$
(4+x)^{\frac{1}{2}}
=
2\left[
1+\dfrac{1}{2}\left(\dfrac{x}{4}\right)
-\dfrac{1}{8}\left(\dfrac{x}{4}\right)^2
+\dfrac{1}{16}\left(\dfrac{x}{4}\right)^3
+\cdots
\right].
$$

Therefore,

$$
(4+x)^{\frac{1}{2}}
=
2+\dfrac{x}{4}-\dfrac{x^2}{64}+\dfrac{x^3}{512}+\cdots.
$$

The expansion is valid when

$$
|x|<4.
$$

### Short investigation

Use the approximation to estimate square roots near $4$.

```{r}
#| label: tbl-sqrt-four-plus-x-approximations
#| echo: true

x_values <- c(-0.5, -0.2, 0.1, 0.5)

data.frame(
  x = x_values,
  exact = round(sqrt(4 + x_values), 8),
  approximation_up_to_x_squared = round(2 + x_values / 4 - x_values^2 / 64, 8)
)
```

### Exercises 7.4

#### Bronze

1. Expand $(2+x)^{-1}$ up to and including the term in $x^2$.
2. Expand $(4+x)^{\frac{1}{2}}$ up to and including the term in $x^2$.
3. Expand $(3+x)^{-1}$ up to and including the term in $x^2$.
4. State the range of validity for the expansion of $(5+x)^{-\frac{1}{2}}$.
5. Use the expansion of $(4+x)^{\frac{1}{2}}$ up to and including the term in $x^2$ to estimate $\sqrt{4.1}$.

#### Silver

1. Expand $(2-3x)^{-1}$ up to and including the term in $x^3$.
2. Expand $(9+x)^{\frac{1}{2}}$ up to and including the term in $x^2$.
3. Expand $(4-2x)^{-\frac{1}{2}}$ up to and including the term in $x^2$.
4. Find the range of validity for the expansion of $(3-2x)^{-1}$.
5. Find the coefficient of $x^3$ in the expansion of $(2+x)^{-3}$.

#### Gold

1. Show that
   $$
   (a+x)^n
   =
   a^n\left(1+\dfrac{x}{a}\right)^n.
   $$
   Hence explain why the range of validity is $|x|<|a|$.
2. Find $k$ if the coefficient of $x^2$ in $(4+kx)^{\frac{1}{2}}$ is $-\dfrac{1}{64}$.
3. Expand
   $$
   \dfrac{\sqrt{4+x}}{2-x}
   $$
   up to and including the term in $x^2$.
4. Find the range of validity for the expansion of
   $$
   \dfrac{\sqrt{4+x}}{2-x}.
   $$
5. Use the expansion of $(9+x)^{\frac{1}{2}}$ up to and including the term in $x^2$ to estimate $\sqrt{9.2}$.

## 7.5 Partial fractions and binomial expansions

### Key mathematical ideas

Partial fractions are often used before binomial expansion.

For example,

$$
\dfrac{1+5x}{(1-x)(1+2x)}
$$

can be split into simpler fractions:

$$
\dfrac{1+5x}{(1-x)(1+2x)}
=
\dfrac{2}{1-x}-\dfrac{1}{1+2x}.
$$

Each term can then be expanded separately.

The expansion

$$
\dfrac{1}{1-x}
=
1+x+x^2+x^3+\cdots
$$

is valid for

$$
|x|<1.
$$

The expansion

$$
\dfrac{1}{1+2x}
=
1-2x+4x^2-8x^3+\cdots
$$

is valid for

$$
|2x|<1.
$$

So the combined expansion is valid for

$$
|x|<\dfrac{1}{2}.
$$

### Visualisation using R

```{r}
#| label: fig-partial-fractions-binomial-expansion
#| fig-cap: "Exact function and binomial approximation after partial fractions."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-0.45, 0.45, length.out = 800)

df <- data.frame(
  x = x_values
)

df$exact <- (1 + 5 * df$x) / ((1 - df$x) * (1 + 2 * df$x))
df$approximation <- 1 + 4 * df$x - 2 * df$x^2 + 10 * df$x^3

ggplot(df, aes(x = x)) +
  geom_line(aes(y = exact, linetype = "exact"), linewidth = 1.1) +
  geom_line(aes(y = approximation, linetype = "up to x^3"), linewidth = 1.1) +
  labs(
    title = "Partial fractions before binomial expansion",
    subtitle = "The approximation is accurate near x = 0 and valid for |x| < 1/2.",
    x = "x",
    y = "value",
    linetype = "Curve"
  ) +
  book_theme()
```

### Worked example

Obtain the expansion of

$$
\dfrac{1+5x}{(1-x)(1+2x)}
$$

in ascending powers of $x$, up to and including the term in $x^3$.

First write

$$
\dfrac{1+5x}{(1-x)(1+2x)}
=
\dfrac{A}{1-x}+\dfrac{B}{1+2x}.
$$

Multiplying through by $(1-x)(1+2x)$ gives

$$
1+5x=A(1+2x)+B(1-x).
$$

Equating coefficients gives

$$
A+B=1
$$

and

$$
2A-B=5.
$$

Solving gives

$$
A=2,\quad B=-1.
$$

Therefore,

$$
\dfrac{1+5x}{(1-x)(1+2x)}
=
\dfrac{2}{1-x}-\dfrac{1}{1+2x}.
$$

Now,

$$
\dfrac{2}{1-x}
=
2(1+x+x^2+x^3+\cdots)
$$

and

$$
-\dfrac{1}{1+2x}
=
-(1-2x+4x^2-8x^3+\cdots).
$$

Therefore,

$$
\dfrac{1+5x}{(1-x)(1+2x)}
=
1+4x-2x^2+10x^3+\cdots.
$$

The expansion is valid when

$$
|x|<\dfrac{1}{2}.
$$

### Short investigation

Compare exact and approximate values.

```{r}
#| label: tbl-partial-fractions-binomial-values
#| echo: true

x_values <- c(-0.2, -0.1, 0.1, 0.2)

data.frame(
  x = x_values,
  exact = round((1 + 5 * x_values) / ((1 - x_values) * (1 + 2 * x_values)), 6),
  approximation = round(1 + 4 * x_values - 2 * x_values^2 + 10 * x_values^3, 6)
)
```

### Exercises 7.5

#### Bronze

1. Obtain the expansion of $\dfrac{3+x}{(1-x)(1+x)}$ up to and including the term in $x^3$.
2. Obtain the expansion of $\dfrac{1}{(1-x)(1+2x)}$ up to and including the term in $x^3$.
3. Obtain the expansion of $\dfrac{1+5x}{(1-x)(1+2x)}$ up to and including the term in $x^3$.
4. State the range of validity for the expansion of $\dfrac{1+5x}{(1-x)(1+2x)}$.
5. Obtain the expansion of $\dfrac{2+7x}{(1-x)(1+3x)}$ up to and including the term in $x^2$.

#### Silver

1. Obtain the expansion of $\dfrac{1+4x+x^2}{(1-x)(1+2x)}$ up to and including the term in $x^3$.
2. Obtain the expansion of $\dfrac{5+2x}{(1-x)(1+x)^2}$ up to and including the term in $x^3$.
3. Obtain the expansion of $\dfrac{2-5x}{(1+x)(1-3x)}$ up to and including the term in $x^3$. State the range of validity.
4. Use the expansion of $\dfrac{1+5x}{(1-x)(1+2x)}$ to estimate its value when $x=0.1$.
5. Find the coefficient of $x^3$ in the expansion of $\dfrac{1+5x}{(1-x)(1+2x)}$.

#### Gold

1. Show that
   $$
   \dfrac{5+2x}{(1-x)(1+x)^2}
   =
   \dfrac{7}{4(1-x)}
   +\dfrac{7}{4(1+x)}
   +\dfrac{3}{2(1+x)^2}.
   $$
2. Explain why long division is needed before expanding
   $$
   \dfrac{1+4x+x^2}{(1-x)(1+2x)}.
   $$
3. Find the range of validity for the expansion of
   $$
   \dfrac{2-5x}{(1+x)(1-3x)}.
   $$
4. Find $k$ if the coefficient of $x$ in the expansion of
   $$
   \dfrac{1+kx}{(1-x)(1+2x)}
   $$
   is $5$.
5. Find the coefficient of $x^3$ in the expansion of
   $$
   \dfrac{1+ax}{(1-x)(1+2x)}
   $$
   in terms of $a$.

## STEP-style extension questions

These questions are designed to require proof, structure or exact reasoning.

### STEP-style Question 1

Let $n$ be rational. Prove that the coefficient of $x^r$ in the binomial expansion of $(1+x)^n$ is

$$
\dfrac{n(n-1)(n-2)\cdots(n-r+1)}{r!}.
$$

Use your result to find the coefficient of $x^4$ in

$$
(1-2x)^{-\frac{1}{2}}.
$$

### STEP-style Question 2

Use partial fractions to prove that

$$
\sum_{r=1}^{n}\dfrac{1}{r(r+1)}
=
\dfrac{n}{n+1}.
$$

### STEP-style Question 3

Show that

$$
\dfrac{1+x}{(1-2x)(1+x)}
=
\dfrac{1}{1-2x},
$$

for $x\neq -1$. Then explain why this simplification affects the apparent range of validity of a binomial expansion.

### STEP-style Question 4

Express

$$
\dfrac{x^3+3x^2+5x+1}{(x+1)(x^2+1)}
$$

in the form

$$
A+\dfrac{B}{x+1}+\dfrac{Cx+D}{x^2+1}.
$$

### STEP-style Question 5

Let

$$
F(x)=\dfrac{1+ax}{(1-x)(1+2x)}.
$$

Find the value of $a$ such that the coefficient of $x^3$ in the expansion of $F(x)$ is zero.

## Reflection questions

1. Why must an improper algebraic fraction be divided before partial fractions?
2. What are the three main partial-fraction types in this chapter?
3. Why does an irreducible quadratic factor require a linear numerator?
4. Why does the binomial expansion of $(1+x)^n$ have a validity condition when $n$ is not a non-negative integer?
5. Why do extra terms improve approximation near $x=0$?
6. Why should $(a+x)^n$ be rewritten as $a^n(1+\frac{x}{a})^n$ before expanding?
7. Why does the final range of validity in a combined expansion come from the most restrictive condition?
8. What can R show clearly about approximation error that algebra alone may hide?
9. What can exact algebra prove that a graph can only suggest?

## Answers

### Answers to Exercises 7.1

#### Bronze

1. $x+3$.
2. $x+1$.
3. $x^2+x+2+\dfrac{2}{x+1}$.
4. $2x^2-3x+6-\dfrac{17}{x+2}$.
5. $x^2-1$.

#### Silver

1. $x^2+2x+\dfrac{1}{x-2}$.
2. $3x^2-5x+6-\dfrac{13}{x+1}$.
3. $x^2-2x+5+\dfrac{-8x-4}{x^2+2x+1}$.
4. $2x^2-x+5+\dfrac{-2x+10}{x^2-1}$.
5. $x^3-x^2+1$.

#### Gold

1. $x-1+\dfrac{2}{x^2+x+1}$.
2. $3x^2-x+4+\dfrac{-6x+11}{x^2+x-1}$.
3. The numerator has degree $3$ and the denominator has degree $2$, so the fraction is improper.
4. $a=3$ and $b=1$.
5. Since $x^4-1=(x^2+1)(x^2-1)$, the quotient is $x^2-1$ and the remainder is $0$.

### Answers to Exercises 7.2

#### Bronze

1. $\dfrac{2}{x-1}+\dfrac{3}{x+2}$.
2. $-\dfrac{3}{x+1}+\dfrac{10}{x+3}$.
3. $\dfrac{3}{x-1}+\dfrac{8}{(x-1)^2}$.
4. $\dfrac{4}{x+2}-\dfrac{9}{(x+2)^2}$.
5. $\dfrac{2}{x-1}+\dfrac{-2x+5}{x^2+2}$.

#### Silver

1. $\dfrac{3}{x-2}-\dfrac{1}{x+1}$.
2. $\dfrac{2}{x+2}-\dfrac{3}{(x+2)^2}$.
3. $\dfrac{1}{x+1}+\dfrac{2x-1}{x^2+4}$.
4. $\dfrac{1}{x}-\dfrac{1}{x+1}$.
5. $1+\dfrac{3}{x+1}-\dfrac{1}{x+2}$.

#### Gold

1. $\dfrac{1}{x+1}+\dfrac{2}{(x+1)^2}-\dfrac{1}{x+3}$.
2. $\dfrac{3}{x-2}+\dfrac{-x+2}{x^2+1}$.
3. The numerator and denominator both have degree $2$, so the fraction is improper and long division must be done first.
4. The identity is
   $$
   \dfrac{1}{r(r+1)}=\dfrac{1}{r}-\dfrac{1}{r+1}.
   $$
   Therefore,
   $$
   \sum_{r=1}^{n}\dfrac{1}{r(r+1)}
   =
   1-\dfrac{1}{n+1}
   =
   \dfrac{n}{n+1}.
   $$
5. $k=5$.

### Answers to Exercises 7.3

#### Bronze

1. $1+\dfrac{1}{2}x-\dfrac{1}{8}x^2+\cdots$.
2. $1-x+x^2-x^3+\cdots$.
3. $1-2x+3x^2-4x^3+\cdots$.
4. $1+\dfrac{3}{2}x+\dfrac{3}{8}x^2+\cdots$.
5. $|x|<1$.

#### Silver

1. $1+2x+4x^2+8x^3+\cdots$, valid for $|x|<\dfrac{1}{2}$.
2. $1+\dfrac{3}{2}x-\dfrac{9}{8}x^2+\dfrac{27}{16}x^3+\cdots$.
3. $1+2x+6x^2+\cdots$.
4. $\sqrt{1.04}\approx1.0198$.
5. $1-\dfrac{1}{2}x+\dfrac{3}{8}x^2-\dfrac{5}{16}x^3+\cdots$.

#### Gold

1. The coefficient of $x^3$ is the third binomial coefficient:
   $$
   \dfrac{n(n-1)(n-2)}{3!}
   =
   \dfrac{n(n-1)(n-2)}{6}.
   $$
2. $-\dfrac{1}{2}$.
3. $1+\dfrac{4}{3}x+\dfrac{11}{9}x^2+\dfrac{104}{81}x^3+\cdots$.
4. $k=\pm4$.
5. The range is $|x|<\dfrac{1}{2}$. Substituting $x=0.6$ is invalid because $|2x|>1$.

### Answers to Exercises 7.4

#### Bronze

1. $\dfrac{1}{2}-\dfrac{x}{4}+\dfrac{x^2}{8}+\cdots$.
2. $2+\dfrac{x}{4}-\dfrac{x^2}{64}+\cdots$.
3. $\dfrac{1}{3}-\dfrac{x}{9}+\dfrac{x^2}{27}+\cdots$.
4. $|x|<5$.
5. $\sqrt{4.1}\approx2.02484$.

#### Silver

1. $\dfrac{1}{2}+\dfrac{3}{4}x+\dfrac{9}{8}x^2+\dfrac{27}{16}x^3+\cdots$.
2. $3+\dfrac{x}{6}-\dfrac{x^2}{216}+\cdots$.
3. $\dfrac{1}{2}+\dfrac{x}{8}+\dfrac{3x^2}{64}+\cdots$.
4. $|x|<\dfrac{3}{2}$.
5. $-\dfrac{5}{32}$.

#### Gold

1. Factorising gives
   $$
   (a+x)^n=a^n\left(1+\dfrac{x}{a}\right)^n.
   $$
   The binomial expansion requires
   $$
   \left|\dfrac{x}{a}\right|<1,
   $$
   so $|x|<|a|$.
2. $k=\pm1$.
3. $1+\dfrac{5}{8}x+\dfrac{39}{128}x^2+\cdots$.
4. $|x|<2$.
5. $\sqrt{9.2}\approx3.03315$.

### Answers to Exercises 7.5

#### Bronze

1. $3+x+3x^2+x^3+\cdots$.
2. $1-x+3x^2-5x^3+\cdots$.
3. $1+4x-2x^2+10x^3+\cdots$.
4. $|x|<\dfrac{1}{2}$.
5. $2+3x+\cdots$.

#### Silver

1. $1+3x+6x^3+\cdots$.
2. $5-3x+8x^2-6x^3+\cdots$.
3. $2-x+4x^2+5x^3+\cdots$, valid for $|x|<\dfrac{1}{3}$.
4. The estimate is $1.390$.
5. $10$.

#### Gold

1. Multiplying the right-hand side by $(1-x)(1+x)^2$ gives $5+2x$, so the identity holds.
2. The numerator and denominator both have degree $2$, so the fraction is improper.
3. The conditions are $|x|<1$ and $|3x|<1$, so the range is $|x|<\dfrac{1}{3}$.
4. $k=6$.
5. $3a-5$.

### Answers to STEP-style extension questions

1. The coefficient of $x^r$ is obtained by choosing the first $r$ descending factors from $n$ and dividing by $r!$:
   $$
   \dfrac{n(n-1)(n-2)\cdots(n-r+1)}{r!}.
   $$
   For $(1-2x)^{-\frac{1}{2}}$, the coefficient of $x^4$ is
   $$
   \dfrac{35}{128}\cdot16=\dfrac{35}{8}.
   $$

2. Since
   $$
   \dfrac{1}{r(r+1)}
   =
   \dfrac{1}{r}-\dfrac{1}{r+1},
   $$
   the sum telescopes:
   $$
   \sum_{r=1}^{n}\dfrac{1}{r(r+1)}
   =
   \left(1-\dfrac{1}{2}\right)+\left(\dfrac{1}{2}-\dfrac{1}{3}\right)+\cdots+\left(\dfrac{1}{n}-\dfrac{1}{n+1}\right).
   $$
   Therefore,
   $$
   \sum_{r=1}^{n}\dfrac{1}{r(r+1)}
   =
   1-\dfrac{1}{n+1}
   =
   \dfrac{n}{n+1}.
   $$

3. For $x\neq-1$,
   $$
   \dfrac{1+x}{(1-2x)(1+x)}
   =
   \dfrac{1}{1-2x}.
   $$
   If the expression is simplified first, the only binomial expansion needed is
   $$
   \dfrac{1}{1-2x}=1+2x+4x^2+\cdots,
   $$
   which is valid for $|x|<\dfrac{1}{2}$. Without simplification, the factor $1+x$ may misleadingly suggest another condition, but it cancels from the function.

4. Long division gives quotient $1$ and remainder $2x^2+4x$. Therefore,
   $$
   \dfrac{x^3+3x^2+5x+1}{(x+1)(x^2+1)}
   =
   1+\dfrac{2x^2+4x}{(x+1)(x^2+1)}.
   $$
   Now
   $$
   \dfrac{2x^2+4x}{(x+1)(x^2+1)}
   =
   -\dfrac{1}{x+1}+\dfrac{3x+1}{x^2+1}.
   $$
   Hence,
   $$
   \dfrac{x^3+3x^2+5x+1}{(x+1)(x^2+1)}
   =
   1-\dfrac{1}{x+1}+\dfrac{3x+1}{x^2+1}.
   $$

5. Expanding
   $$
   \dfrac{1+ax}{(1-x)(1+2x)}
   $$
   gives coefficient of $x^3$ equal to
   $$
   3a-5.
   $$
   Setting this equal to zero gives
   $$
   a=\dfrac{5}{3}.
   $$