# Vectors

```{r}
#| label: setup-vectors
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

axis_layers <- function() {
  list(
    geom_hline(yintercept = 0),
    geom_vline(xintercept = 0)
  )
}

arrow_style <- function() {
  grid::arrow(length = grid::unit(0.18, "cm"))
}

make_line_2d <- function(point, direction, t_min = -3, t_max = 3, n = 250) {
  t_values <- seq(t_min, t_max, length.out = n)
  data.frame(
    t = t_values,
    x = point[1] + t_values * direction[1],
    y = point[2] + t_values * direction[2]
  )
}

points_on_line_2d <- function(point, direction, t_values) {
  data.frame(
    t = t_values,
    x = point[1] + t_values * direction[1],
    y = point[2] + t_values * direction[2],
    label = paste0("t = ", t_values)
  )
}

vector_arrows_2d <- function(labels, starts, ends) {
  data.frame(
    label = labels,
    x = starts[, 1],
    y = starts[, 2],
    xend = ends[, 1],
    yend = ends[, 2]
  )
}

arc_between_vectors <- function(a, b, radius = 0.8, n = 100) {
  angle_a <- atan2(a[2], a[1])
  angle_b <- atan2(b[2], b[1])
  angle_values <- seq(angle_a, angle_b, length.out = n)
  data.frame(
    x = radius * cos(angle_values),
    y = radius * sin(angle_values)
  )
}

vector_magnitude <- function(a) {
  sqrt(sum(a^2))
}

scalar_product <- function(a, b) {
  sum(a * b)
}

vector_angle_degrees <- function(a, b, acute = FALSE) {
  numerator <- scalar_product(a, b)
  if (acute) {
    numerator <- abs(numerator)
  }
  denominator <- vector_magnitude(a) * vector_magnitude(b)
  acos(numerator / denominator) * 180 / pi
}

closest_parameter <- function(point, line_point, direction) {
  scalar_product(point - line_point, direction) / scalar_product(direction, direction)
}

closest_point_on_line <- function(point, line_point, direction) {
  t_value <- closest_parameter(point, line_point, direction)
  line_point + t_value * direction
}
```

## Overview

This chapter develops six connected vector ideas:

1. position vectors and displacement vectors;
2. magnitude, unit vectors and parallel vectors;
3. vector equations of straight lines;
4. the scalar product and angles;
5. intersections of straight lines;
6. closest points and distances from points to lines.

The central aim is to understand the meaning of a vector equation of a line and to use scalar products for problem solving. We do not need equations of planes in this chapter.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 9.1 Position vectors and displacement vectors

### Key mathematical ideas

A **position vector** gives the position of a point relative to the origin.

If

$$
\overrightarrow{OA}=\mathbf{a}
$$

and

$$
\overrightarrow{OB}=\mathbf{b},
$$

then

$$
\overrightarrow{AB}=\mathbf{b}-\mathbf{a}.
$$

This is one of the most important ideas in the chapter: to move from $A$ to $B$, subtract the position vector of $A$ from the position vector of $B$.

For a vector

$$
\mathbf{a}=a_1\mathbf{i}+a_2\mathbf{j}+a_3\mathbf{k},
$$

its magnitude is

$$
|\mathbf{a}|=\sqrt{a_1^2+a_2^2+a_3^2}.
$$

A unit vector in the direction of a non-zero vector $\mathbf{a}$ is

$$
\dfrac{\mathbf{a}}{|\mathbf{a}|}.
$$

### Visualisation using R

```{r}
#| label: fig-position-displacement-vectors
#| fig-cap: "Position vectors and the displacement vector from A to B."
#| fig-width: 6
#| fig-height: 5

points_df <- data.frame(
  label = c("O", "A", "B"),
  x = c(0, 2, 5),
  y = c(0, 1, 4)
)

arrows_df <- vector_arrows_2d(
  labels = c("OA", "OB", "AB"),
  starts = matrix(c(0, 0, 0, 0, 2, 1), ncol = 2, byrow = TRUE),
  ends = matrix(c(2, 1, 5, 4, 5, 4), ncol = 2, byrow = TRUE)
)

ggplot() +
  geom_segment(
    data = arrows_df,
    aes(x = x, y = y, xend = xend, yend = yend, linetype = label),
    linewidth = 1.1,
    arrow = arrow_style()
  ) +
  geom_point(data = points_df, aes(x = x, y = y), size = 3) +
  geom_text(data = points_df, aes(x = x, y = y, label = label), nudge_y = 0.25) +
  axis_layers() +
  coord_equal() +
  labs(
    title = "Position and displacement vectors",
    subtitle = "The displacement vector AB is obtained by b - a.",
    x = "x",
    y = "y",
    linetype = "Vector"
  ) +
  book_theme()
```

### Worked example

Let

$$
A(2,-1,3)
$$

and

$$
B(-1,5,4).
$$

Find $\overrightarrow{AB}$ and $|\overrightarrow{AB}|$.

The position vectors are

$$
\overrightarrow{OA}=\begin{pmatrix}2\\-1\\3\end{pmatrix}
$$

and

$$
\overrightarrow{OB}=\begin{pmatrix}-1\\5\\4\end{pmatrix}.
$$

Therefore,

$$
\overrightarrow{AB}
=
\overrightarrow{OB}-\overrightarrow{OA}
=
\begin{pmatrix}-1\\5\\4\end{pmatrix}
-
\begin{pmatrix}2\\-1\\3\end{pmatrix}
=
\begin{pmatrix}-3\\6\\1\end{pmatrix}.
$$

So

$$
|\overrightarrow{AB}|
=
\sqrt{(-3)^2+6^2+1^2}
=
\sqrt{46}.
$$

### Short investigation

Change the coordinates of $A$ and $B$ below and observe how $\overrightarrow{AB}$ changes.

```{r}
#| label: tbl-displacement-vector-check
#| echo: true

A <- c(2, -1, 3)
B <- c(-1, 5, 4)

AB <- B - A

data.frame(
  component = c("x", "y", "z"),
  A = A,
  B = B,
  AB = AB
)

vector_magnitude(AB)
```

### Exercises 9.1

#### Bronze

1. If $A(3,-2)$ and $B(5,4)$, find $\overrightarrow{AB}$.
2. Find the magnitude of $\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
3. If $A(-1,4,0)$ and $B(2,1,5)$, find $\overrightarrow{AB}$.
4. If $\overrightarrow{OA}=\mathbf{a}$ and $\overrightarrow{OB}=\mathbf{b}$, write down $\overrightarrow{BA}$.
5. Find the midpoint of $A(1,2,3)$ and $B(5,0,-1)$.

#### Silver

1. If $A(2,-1,3)$ and $B(-1,5,4)$, find $\overrightarrow{AB}$ and $|\overrightarrow{AB}|$.
2. If $\overrightarrow{AB}=\begin{pmatrix}4\\-2\\6\end{pmatrix}$ and $A(1,3,-2)$, find $B$.
3. Find a unit vector in the direction of $\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
4. A point $P$ divides $AB$ in the ratio $AP:PB=1:2$. If $A(1,0,2)$ and $B(4,6,-1)$, find $P$.
5. Show that $A(1,2,3)$, $B(3,0,5)$ and $C(5,-2,7)$ are collinear.

#### Gold

1. Show that $P(3,-1,5)$ is the midpoint of $A(1,0,2)$ and $B(5,-2,8)$.
2. Find $k$ if $\left|\begin{pmatrix}k\\2\\1\end{pmatrix}\right|=3$.
3. Points $C(1,-1,-3)$, $A(2,1,-1)$ and $B(4,5,3)$ are collinear. Find the ratio $CA:AB$.
4. Find $x$ if $A(1,2,3)$, $B(4,8,12)$ and $C(x,4,6)$ are collinear.
5. Find the distance between $A(-1,4,2)$ and $B(3,-2,5)$.

## 9.2 Vector equations of straight lines

### Key mathematical ideas

A straight line is determined by:

- one point on the line;
- one direction vector.

If a line passes through the point with position vector $\mathbf{a}$ and has direction vector $\mathbf{d}$, then its vector equation is

$$
\mathbf{r}=\mathbf{a}+t\mathbf{d},
$$

where $t$ is a real parameter.

Each value of $t$ gives a point on the line.

If the line passes through points with position vectors $\mathbf{a}$ and $\mathbf{b}$, then a direction vector is

$$
\mathbf{b}-\mathbf{a}.
$$

So the line can be written as

$$
\mathbf{r}=\mathbf{a}+t(\mathbf{b}-\mathbf{a}).
$$

The direction vector is not unique. Any non-zero scalar multiple of a direction vector describes the same direction.

### Visualisation using R

```{r}
#| label: fig-vector-equation-line
#| fig-cap: "The line r = a + td in a two-dimensional projection."
#| fig-width: 6
#| fig-height: 5

a <- c(1, 2)
d <- c(2, -1)

line_df <- make_line_2d(a, d, t_min = -2, t_max = 3)
sample_points <- points_on_line_2d(a, d, t_values = c(-1, 0, 1, 2))

ggplot() +
  geom_line(data = line_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = sample_points, aes(x = x, y = y), size = 3) +
  geom_text(data = sample_points, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  geom_segment(
    aes(x = a[1], y = a[2], xend = a[1] + d[1], yend = a[2] + d[2]),
    linewidth = 1.1,
    arrow = arrow_style()
  ) +
  axis_layers() +
  coord_equal() +
  labs(
    title = "Vector equation of a line",
    subtitle = "The parameter t moves the point along the line.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Find the vector equation of the line through

$$
A(1,2,-1)
$$

and

$$
B(5,-2,3).
$$

A direction vector is

$$
\overrightarrow{AB}
=
\begin{pmatrix}5\\-2\\3\end{pmatrix}
-
\begin{pmatrix}1\\2\\-1\end{pmatrix}
=
\begin{pmatrix}4\\-4\\4\end{pmatrix}.
$$

We may divide this direction vector by $4$ and use

$$
\begin{pmatrix}1\\-1\\1\end{pmatrix}.
$$

Therefore, one vector equation of the line is

$$
\mathbf{r}
=
\begin{pmatrix}1\\2\\-1\end{pmatrix}
+t\begin{pmatrix}1\\-1\\1\end{pmatrix}.
$$

### Short investigation

The table below shows how changing $t$ moves along the line.

```{r}
#| label: tbl-line-parameter-values
#| echo: true

t_values <- -2:4

data.frame(
  t = t_values,
  x = 1 + t_values,
  y = 2 - t_values,
  z = -1 + t_values
)
```

### Exercises 9.2

#### Bronze

1. Write down a vector equation of the line through $(1,2,3)$ with direction vector $\begin{pmatrix}2\\-1\\4\end{pmatrix}$.
2. Find a vector equation of the line through $A(1,0,2)$ and $B(3,4,6)$.
3. Show that $P(5,-2,11)$ lies on the line $\mathbf{r}=\begin{pmatrix}1\\2\\3\end{pmatrix}+t\begin{pmatrix}2\\-2\\4\end{pmatrix}$.
4. Determine whether $\begin{pmatrix}2\\-4\\6\end{pmatrix}$ and $\begin{pmatrix}-1\\2\\-3\end{pmatrix}$ are parallel.
5. Give another vector equation for the same line as $\mathbf{r}=\begin{pmatrix}1\\1\\0\end{pmatrix}+t\begin{pmatrix}2\\-1\\3\end{pmatrix}$.

#### Silver

1. Find a vector equation of the line through $A(-1,2,4)$ and $B(5,-1,1)$.
2. Find $k$ if $P(k,7,3)$ lies on the line $\mathbf{r}=\begin{pmatrix}1\\3\\5\end{pmatrix}+t\begin{pmatrix}2\\4\\-2\end{pmatrix}$.
3. Determine whether the lines $\mathbf{r}=\begin{pmatrix}1\\0\\2\end{pmatrix}+t\begin{pmatrix}2\\-1\\3\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}3\\-1\\5\end{pmatrix}+s\begin{pmatrix}4\\-2\\6\end{pmatrix}$ are the same line.
4. Find the value of $\lambda$ for which $(7,-1,10)$ lies on $\mathbf{r}=\begin{pmatrix}1\\2\\-2\end{pmatrix}+\lambda\begin{pmatrix}2\\-1\\4\end{pmatrix}$.
5. Show that $A(2,-1,0)$, $B(5,5,9)$ and $C(1,-3,-3)$ are collinear.

#### Gold

1. Find $k$ if $A(1,2,3)$, $B(4,k,9)$ and $C(7,8,15)$ are collinear.
2. The line passes through $A(2,1,-1)$ and $B(5,-2,8)$. Show that $P(8,-5,17)$ lies on the line.
3. Prove that if $\mathbf{r}=\mathbf{a}+t\mathbf{d}$ and $\mathbf{r}=\mathbf{a}+s\mathbf{d}$ represent the same point, and $\mathbf{d}\neq\mathbf{0}$, then $t=s$.
4. If $P=A+\lambda(B-A)$ divides $AB$ in the ratio $AP:PB=2:3$, find $\lambda$.
5. Show that $\mathbf{r}=\begin{pmatrix}2\\0\\1\end{pmatrix}+t\begin{pmatrix}1\\-2\\3\end{pmatrix}$ is the same line as the line through $P(-1,6,-8)$ and $Q(2,0,1)$.

## 9.3 The scalar product and angles

### Key mathematical ideas

The scalar product, or dot product, of two vectors is defined by

$$
\mathbf{a}\cdot\mathbf{b}=|\mathbf{a}|\,|\mathbf{b}|\cos\theta,
$$

where $\theta$ is the angle between the two vectors.

If

$$
\mathbf{a}=a_1\mathbf{i}+a_2\mathbf{j}+a_3\mathbf{k}
$$

and

$$
\mathbf{b}=b_1\mathbf{i}+b_2\mathbf{j}+b_3\mathbf{k},
$$

then

$$
\mathbf{a}\cdot\mathbf{b}=a_1b_1+a_2b_2+a_3b_3.
$$

Combining the two formulae gives

$$
\cos\theta=
\dfrac{\mathbf{a}\cdot\mathbf{b}}{|\mathbf{a}|\,|\mathbf{b}|}.
$$

Two non-zero vectors are perpendicular when

$$
\mathbf{a}\cdot\mathbf{b}=0.
$$

### Visualisation using R

```{r}
#| label: fig-scalar-product-angle
#| fig-cap: "The angle between two vectors in a two-dimensional projection."
#| fig-width: 6
#| fig-height: 5

a_vector <- c(3, 1)
b_vector <- c(1, 3)

vectors_df <- vector_arrows_2d(
  labels = c("a", "b"),
  starts = matrix(c(0, 0, 0, 0), ncol = 2, byrow = TRUE),
  ends = rbind(a_vector, b_vector)
)

arc_df <- arc_between_vectors(a_vector, b_vector, radius = 0.8)

ggplot() +
  geom_segment(
    data = vectors_df,
    aes(x = x, y = y, xend = xend, yend = yend, linetype = label),
    linewidth = 1.1,
    arrow = arrow_style()
  ) +
  geom_path(data = arc_df, aes(x = x, y = y), linewidth = 1.1) +
  annotate("text", x = 0.9, y = 0.55, label = "theta") +
  axis_layers() +
  coord_equal(xlim = c(-0.5, 3.5), ylim = c(-0.5, 3.5)) +
  labs(
    title = "Angle between vectors",
    subtitle = "The scalar product connects algebraic components with the angle between vectors.",
    x = "x",
    y = "y",
    linetype = "Vector"
  ) +
  book_theme()
```

### Worked example

Find the angle between

$$
\mathbf{a}=\begin{pmatrix}2\\-1\\2\end{pmatrix}
$$

and

$$
\mathbf{b}=\begin{pmatrix}1\\2\\2\end{pmatrix}.
$$

First compute the scalar product:

$$
\mathbf{a}\cdot\mathbf{b}=2(1)+(-1)(2)+2(2)=4.
$$

Also,

$$
|\mathbf{a}|=\sqrt{2^2+(-1)^2+2^2}=3
$$

and

$$
|\mathbf{b}|=\sqrt{1^2+2^2+2^2}=3.
$$

Therefore,

$$
\cos\theta=\dfrac{4}{3\cdot3}=\dfrac{4}{9}.
$$

So

$$
\theta=\cos^{-1}\left(\dfrac{4}{9}\right)
\approx63.6^\circ.
$$

### Short investigation

Use R to check angles between several vectors.

```{r}
#| label: tbl-vector-angle-checks
#| echo: true

vectors <- data.frame(
  pair = c("Pair 1", "Pair 2", "Pair 3"),
  ax = c(2, 1, 1),
  ay = c(-1, 2, 1),
  az = c(2, 2, 0),
  bx = c(1, 2, 1),
  by = c(2, 1, -1),
  bz = c(2, 2, 0)
)

vectors$dot_product <- vectors$ax * vectors$bx + vectors$ay * vectors$by + vectors$az * vectors$bz
vectors$magnitude_a <- sqrt(vectors$ax^2 + vectors$ay^2 + vectors$az^2)
vectors$magnitude_b <- sqrt(vectors$bx^2 + vectors$by^2 + vectors$bz^2)
vectors$angle_degrees <- acos(vectors$dot_product / (vectors$magnitude_a * vectors$magnitude_b)) * 180 / pi

vectors
```

### Exercises 9.3

#### Bronze

1. Find $\begin{pmatrix}2\\-1\\3\end{pmatrix}\cdot\begin{pmatrix}4\\0\\-2\end{pmatrix}$.
2. Find the magnitude of $\begin{pmatrix}1\\2\\2\end{pmatrix}$.
3. Find the angle between $\begin{pmatrix}1\\0\\0\end{pmatrix}$ and $\begin{pmatrix}0\\1\\0\end{pmatrix}$.
4. Determine whether $\begin{pmatrix}2\\-1\\1\end{pmatrix}$ and $\begin{pmatrix}1\\4\\2\end{pmatrix}$ are perpendicular.
5. Find $\cos\theta$ for the angle between $\begin{pmatrix}1\\2\\2\end{pmatrix}$ and $\begin{pmatrix}2\\1\\2\end{pmatrix}$.

#### Silver

1. Find the angle between $\begin{pmatrix}2\\-1\\2\end{pmatrix}$ and $\begin{pmatrix}1\\2\\2\end{pmatrix}$.
2. Find $k$ if $\begin{pmatrix}1\\k\\2\end{pmatrix}$ is perpendicular to $\begin{pmatrix}3\\-2\\1\end{pmatrix}$.
3. Find $k>0$ if the angle between $\begin{pmatrix}k\\1\\0\end{pmatrix}$ and $\begin{pmatrix}1\\0\\0\end{pmatrix}$ is $60^\circ$.
4. Show that the vectors $\begin{pmatrix}1\\2\\3\end{pmatrix}$ and $\begin{pmatrix}4\\-2\\0\end{pmatrix}$ are perpendicular.
5. Find the scalar projection of $\begin{pmatrix}4\\2\\4\end{pmatrix}$ onto $\begin{pmatrix}1\\0\\1\end{pmatrix}$.

#### Gold

1. Find the acute angle between $\begin{pmatrix}1\\2\\-2\end{pmatrix}$ and $\begin{pmatrix}3\\-1\\2\end{pmatrix}$.
2. Find $k$ if $\begin{pmatrix}1\\k\\2\end{pmatrix}$ and $\begin{pmatrix}2\\-1\\1\end{pmatrix}$ are perpendicular.
3. Find $k$ if $\begin{pmatrix}k\\2\\2\end{pmatrix}$ and $\begin{pmatrix}2\\k\\1\end{pmatrix}$ are perpendicular.
4. Starting from $\mathbf{a}\cdot\mathbf{b}=|\mathbf{a}|\,|\mathbf{b}|\cos\theta$, derive the formula for $\cos\theta$.
5. Find angle $BAC$ for $A(1,0,0)$, $B(3,1,2)$ and $C(0,2,1)$.

## 9.4 Angles between straight lines

### Key mathematical ideas

The angle between two straight lines is the angle between their direction vectors.

If the lines have direction vectors $\mathbf{d}_1$ and $\mathbf{d}_2$, then

$$
\cos\theta=
\dfrac{\mathbf{d}_1\cdot\mathbf{d}_2}{|\mathbf{d}_1|\,|\mathbf{d}_2|}.
$$

If the question asks for the acute angle between two lines, use

$$
\cos\theta=
\dfrac{|\mathbf{d}_1\cdot\mathbf{d}_2|}{|\mathbf{d}_1|\,|\mathbf{d}_2|}.
$$

Two lines are parallel if their direction vectors are scalar multiples.

Two lines are perpendicular if their direction vectors have scalar product zero.

### Visualisation using R

```{r}
#| label: fig-angle-between-lines
#| fig-cap: "The angle between two lines is the angle between their direction vectors."
#| fig-width: 6
#| fig-height: 5

line_one <- make_line_2d(c(0, 0), c(2, 1), t_min = -2, t_max = 2)
line_two <- make_line_2d(c(0, 0), c(2, -3), t_min = -2, t_max = 2)

ggplot() +
  geom_line(data = line_one, aes(x = x, y = y, linetype = "line 1"), linewidth = 1.1) +
  geom_line(data = line_two, aes(x = x, y = y, linetype = "line 2"), linewidth = 1.1) +
  axis_layers() +
  coord_equal() +
  labs(
    title = "Angle between lines",
    subtitle = "In vector work, use the direction vectors rather than trying to read the angle from the graph.",
    x = "x",
    y = "y",
    linetype = "Line"
  ) +
  book_theme()
```

### Worked example

Find the acute angle between the lines with direction vectors

$$
\mathbf{d}_1=\begin{pmatrix}1\\2\\-2\end{pmatrix}
$$

and

$$
\mathbf{d}_2=\begin{pmatrix}3\\-1\\2\end{pmatrix}.
$$

The scalar product is

$$
\mathbf{d}_1\cdot\mathbf{d}_2=1(3)+2(-1)+(-2)(2)=-3.
$$

The magnitudes are

$$
|\mathbf{d}_1|=\sqrt{1^2+2^2+(-2)^2}=3
$$

and

$$
|\mathbf{d}_2|=\sqrt{3^2+(-1)^2+2^2}=\sqrt{14}.
$$

For the acute angle,

$$
\cos\theta=
\dfrac{|-3|}{3\sqrt{14}}=
\dfrac{1}{\sqrt{14}}.
$$

Hence

$$
\theta\approx74.5^\circ.
$$

### Short investigation

Use the direction vectors to compare angles between lines.

```{r}
#| label: tbl-line-angle-checks
#| echo: true

line_pairs <- data.frame(
  pair = c("A", "B", "C"),
  d1x = c(1, 1, 3),
  d1y = c(2, 2, -2),
  d1z = c(-2, 2, 6),
  d2x = c(3, 2, 2),
  d2y = c(-1, -1, 4),
  d2z = c(2, 0, 4)
)

line_pairs$dot_product <- line_pairs$d1x * line_pairs$d2x + line_pairs$d1y * line_pairs$d2y + line_pairs$d1z * line_pairs$d2z
line_pairs$magnitude_1 <- sqrt(line_pairs$d1x^2 + line_pairs$d1y^2 + line_pairs$d1z^2)
line_pairs$magnitude_2 <- sqrt(line_pairs$d2x^2 + line_pairs$d2y^2 + line_pairs$d2z^2)
line_pairs$acute_angle_degrees <- acos(abs(line_pairs$dot_product) / (line_pairs$magnitude_1 * line_pairs$magnitude_2)) * 180 / pi

line_pairs
```

### Exercises 9.4

#### Bronze

1. Find the angle between lines with direction vectors $\begin{pmatrix}1\\2\\2\end{pmatrix}$ and $\begin{pmatrix}2\\1\\2\end{pmatrix}$.
2. Determine whether lines with direction vectors $\begin{pmatrix}1\\2\\-1\end{pmatrix}$ and $\begin{pmatrix}2\\4\\-2\end{pmatrix}$ are parallel.
3. Determine whether lines with direction vectors $\begin{pmatrix}1\\1\\0\end{pmatrix}$ and $\begin{pmatrix}1\\-1\\0\end{pmatrix}$ are perpendicular.
4. Find the angle between lines with direction vectors $\begin{pmatrix}1\\2\\2\end{pmatrix}$ and $\begin{pmatrix}2\\-1\\0\end{pmatrix}$.
5. State the direction vector of $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}3\\-1\\4\end{pmatrix}$.

#### Silver

1. Find the acute angle between $\mathbf{r}=\begin{pmatrix}1\\0\\2\end{pmatrix}+\lambda\begin{pmatrix}1\\2\\-2\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}0\\1\\1\end{pmatrix}+\mu\begin{pmatrix}3\\-1\\2\end{pmatrix}$.
2. Find the angle between the line with direction vector $\begin{pmatrix}2\\1\\2\end{pmatrix}$ and the vector $\begin{pmatrix}1\\0\\0\end{pmatrix}$.
3. Find $k$ if the line with direction vector $\begin{pmatrix}1\\k\\2\end{pmatrix}$ is perpendicular to the line with direction vector $\begin{pmatrix}3\\-2\\1\end{pmatrix}$.
4. Find $k$ if $\begin{pmatrix}2\\k\\1\end{pmatrix}$ is parallel to $\begin{pmatrix}6\\9\\3\end{pmatrix}$.
5. Find the angle between lines with direction vectors $\begin{pmatrix}1\\2\\2\end{pmatrix}$ and $\begin{pmatrix}-1\\2\\2\end{pmatrix}$.

#### Gold

1. Show that lines with direction vectors $\begin{pmatrix}1\\2\\-2\end{pmatrix}$ and $\begin{pmatrix}2\\1\\2\end{pmatrix}$ are perpendicular.
2. Find $k>0$ if the angle between $\begin{pmatrix}1\\2\\k\end{pmatrix}$ and $\begin{pmatrix}2\\-1\\2\end{pmatrix}$ is $60^\circ$.
3. Explain why direction vectors can be used to find the angle between two lines even if the lines do not intersect.
4. Find the angle between $AB$ and $CD$, where $A(1,2,0)$, $B(4,0,6)$, $C(-1,1,2)$ and $D(1,5,6)$.
5. Find the acute angle between the line through $(1,2,3)$ and $(4,2,0)$ and the line through $(0,0,0)$ and $(2,1,2)$.

## 9.5 Intersections of straight lines

### Key mathematical ideas

To find the point of intersection of two vector lines, write out the parametric equations and solve them simultaneously.

For example, if

$$
\mathbf{r}=\mathbf{a}+\lambda\mathbf{d}
$$

and

$$
\mathbf{r}=\mathbf{b}+\mu\mathbf{e},
$$

then an intersection point must satisfy

$$
\mathbf{a}+\lambda\mathbf{d}=\mathbf{b}+\mu\mathbf{e}.
$$

This gives simultaneous equations in $\lambda$ and $\mu$.

In three dimensions, two lines may be:

- intersecting;
- parallel;
- the same line;
- skew.

Skew lines are not parallel but do not intersect.

### Visualisation using R

```{r}
#| label: fig-line-intersection-projection
#| fig-cap: "A projection of two intersecting vector lines."
#| fig-width: 6
#| fig-height: 5

line_1 <- make_line_2d(c(1, 2), c(2, -1), t_min = -1, t_max = 3)
line_2 <- make_line_2d(c(5, 0), c(-1, 1), t_min = -2, t_max = 2)

intersection <- data.frame(
  x = 5,
  y = 0,
  label = "intersection"
)

ggplot() +
  geom_line(data = line_1, aes(x = x, y = y, linetype = "line 1"), linewidth = 1.1) +
  geom_line(data = line_2, aes(x = x, y = y, linetype = "line 2"), linewidth = 1.1) +
  geom_point(data = intersection, aes(x = x, y = y), size = 3) +
  geom_text(data = intersection, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal() +
  labs(
    title = "Intersecting vector lines",
    subtitle = "The actual calculation comes from solving the parametric equations.",
    x = "x",
    y = "y",
    linetype = "Line"
  ) +
  book_theme()
```

### Worked example

Find the intersection point of

$$
\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}
+\lambda\begin{pmatrix}2\\-1\\3\end{pmatrix}
$$

and

$$
\mathbf{r}=\begin{pmatrix}5\\0\\6\end{pmatrix}
+\mu\begin{pmatrix}-1\\1\\0\end{pmatrix}.
$$

Equating components gives

$$
1+2\lambda=5-\mu,
$$

$$
2-\lambda=\mu,
$$

and

$$
3\lambda=6.
$$

From the third equation,

$$
\lambda=2.
$$

Then the second equation gives

$$
\mu=2-2=0.
$$

The first equation is also satisfied:

$$
1+2(2)=5-0.
$$

Therefore, the lines intersect at

$$
\begin{pmatrix}5\\0\\6\end{pmatrix}.
$$

### Short investigation

Use R to check whether the component equations agree.

```{r}
#| label: tbl-line-intersection-check
#| echo: true

lambda <- 2
mu <- 0

line_1_point <- c(1, 2, 0) + lambda * c(2, -1, 3)
line_2_point <- c(5, 0, 6) + mu * c(-1, 1, 0)

data.frame(
  component = c("x", "y", "z"),
  line_1 = line_1_point,
  line_2 = line_2_point
)
```

### Exercises 9.5

#### Bronze

1. Find the point on $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}1\\-1\\2\end{pmatrix}$ when $t=3$.
2. Show that $(4,-1,6)$ lies on $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}1\\-1\\2\end{pmatrix}$.
3. Find the intersection of $\mathbf{r}=\begin{pmatrix}1\\0\end{pmatrix}+t\begin{pmatrix}2\\1\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}5\\2\end{pmatrix}+s\begin{pmatrix}-1\\0\end{pmatrix}$.
4. Determine whether direction vectors $\begin{pmatrix}1\\2\\3\end{pmatrix}$ and $\begin{pmatrix}2\\4\\6\end{pmatrix}$ are parallel.
5. In three dimensions, must two non-parallel lines always intersect?

#### Silver

1. Find the intersection of $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+\lambda\begin{pmatrix}2\\-1\\3\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}5\\0\\6\end{pmatrix}+\mu\begin{pmatrix}-1\\1\\0\end{pmatrix}$.
2. Show that $\mathbf{r}=\begin{pmatrix}0\\1\\2\end{pmatrix}+s\begin{pmatrix}1\\2\\-1\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}2\\-1\\0\end{pmatrix}+t\begin{pmatrix}2\\0\\1\end{pmatrix}$ are skew.
3. Determine whether $\mathbf{r}=\begin{pmatrix}1\\0\\1\end{pmatrix}+\lambda\begin{pmatrix}1\\2\\1\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}3\\4\\3\end{pmatrix}+\mu\begin{pmatrix}2\\4\\2\end{pmatrix}$ are the same line.
4. Find $k$ if $\mathbf{r}=\begin{pmatrix}0\\0\\k\end{pmatrix}+s\begin{pmatrix}1\\2\\1\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}1\\2\\3\end{pmatrix}+t\begin{pmatrix}2\\5\\-1\end{pmatrix}$ intersect.
5. Find the intersection of $\mathbf{r}=\lambda\begin{pmatrix}1\\2\\3\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}3\\1\\0\end{pmatrix}+\mu\begin{pmatrix}-2\\1\\3\end{pmatrix}$.

#### Gold

1. Find $k$ if $P(k,4,7)$ lies on $\mathbf{r}=\begin{pmatrix}1\\2\\3\end{pmatrix}+t\begin{pmatrix}2\\1\\2\end{pmatrix}$.
2. Determine whether the lines $\mathbf{r}=\begin{pmatrix}1\\1\\0\end{pmatrix}+s\begin{pmatrix}2\\1\\1\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}3\\0\\2\end{pmatrix}+t\begin{pmatrix}1\\-1\\1\end{pmatrix}$ intersect.
3. Explain why inconsistent simultaneous equations imply that two non-parallel lines in three dimensions are skew.
4. Find the vector equation of the line through the intersection point of $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+\lambda\begin{pmatrix}2\\-1\\3\end{pmatrix}$ and $\mathbf{r}=\begin{pmatrix}5\\0\\6\end{pmatrix}+\mu\begin{pmatrix}-1\\1\\0\end{pmatrix}$, and through the point $(6,2,7)$.
5. For the two original lines used in Gold Question 4, find the intersection point and the acute angle between them.

## 9.6 Closest points and distances from points to lines

### Key mathematical ideas

The scalar product is very useful for finding the closest point from a point to a line.

Let the line be

$$
\mathbf{r}=\mathbf{a}+t\mathbf{d},
$$

and let $P$ have position vector $\mathbf{p}$.

A general point on the line is

$$
\mathbf{x}=\mathbf{a}+t\mathbf{d}.
$$

At the closest point, the vector from the point on the line to $P$ is perpendicular to the direction vector of the line. Therefore,

$$
(\mathbf{p}-\mathbf{x})\cdot\mathbf{d}=0.
$$

Substituting $\mathbf{x}=\mathbf{a}+t\mathbf{d}$ gives

$$
(\mathbf{p}-\mathbf{a}-t\mathbf{d})\cdot\mathbf{d}=0.
$$

So

$$
t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}.
$$

### Visualisation using R

```{r}
#| label: fig-point-to-line-distance
#| fig-cap: "Closest point from a point to a line in a two-dimensional projection."
#| fig-width: 6
#| fig-height: 5

line_point <- c(1, 2)
direction <- c(2, -1)
point_p <- c(4, 0)

line_df <- make_line_2d(line_point, direction, t_min = -1, t_max = 3)
foot <- closest_point_on_line(point_p, line_point, direction)

points_df <- data.frame(
  label = c("P", "foot"),
  x = c(point_p[1], foot[1]),
  y = c(point_p[2], foot[2])
)

ggplot() +
  geom_line(data = line_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_segment(
    aes(x = point_p[1], y = point_p[2], xend = foot[1], yend = foot[2]),
    linetype = "dashed",
    linewidth = 1.1
  ) +
  geom_point(data = points_df, aes(x = x, y = y), size = 3) +
  geom_text(data = points_df, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal() +
  labs(
    title = "Closest point to a line",
    subtitle = "The shortest segment is perpendicular to the line.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Find the closest point from

$$
P(4,0,5)
$$

to the line

$$
\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}2\\-1\\2\end{pmatrix}.
$$

Here,

$$
\mathbf{a}=\begin{pmatrix}1\\2\\0\end{pmatrix},
\quad
\mathbf{d}=\begin{pmatrix}2\\-1\\2\end{pmatrix},
\quad
\mathbf{p}=\begin{pmatrix}4\\0\\5\end{pmatrix}.
$$

Using

$$
t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}},
$$

we get

$$
\mathbf{p}-\mathbf{a}=\begin{pmatrix}3\\-2\\5\end{pmatrix}.
$$

So

$$
(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}=3(2)+(-2)(-1)+5(2)=18,
$$

and

$$
\mathbf{d}\cdot\mathbf{d}=2^2+(-1)^2+2^2=9.
$$

Hence

$$
t=\dfrac{18}{9}=2.
$$

The closest point is

$$
\mathbf{x}=\begin{pmatrix}1\\2\\0\end{pmatrix}+2\begin{pmatrix}2\\-1\\2\end{pmatrix}
=
\begin{pmatrix}5\\0\\4\end{pmatrix}.
$$

The distance from $P$ to the line is

$$
\left|\begin{pmatrix}4\\0\\5\end{pmatrix}-\begin{pmatrix}5\\0\\4\end{pmatrix}\right|
=
\left|\begin{pmatrix}-1\\0\\1\end{pmatrix}\right|
=
\sqrt2.
$$

### Short investigation

Use R to check the perpendicular condition.

```{r}
#| label: tbl-perpendicular-foot-check
#| echo: true

line_point <- c(1, 2, 0)
direction <- c(2, -1, 2)
point_p <- c(4, 0, 5)

foot <- closest_point_on_line(point_p, line_point, direction)
perpendicular_vector <- point_p - foot

data.frame(
  component = c("x", "y", "z"),
  foot = foot,
  perpendicular_vector = perpendicular_vector,
  direction_vector = direction
)

scalar_product(perpendicular_vector, direction)
```

The final value is $0$, confirming that the shortest segment is perpendicular to the line.

### Exercises 9.6

#### Bronze

1. Find the closest point from $P(3,4,0)$ to the line $\mathbf{r}=t\begin{pmatrix}1\\0\\0\end{pmatrix}$.
2. Find the closest point from $P(4,5,2)$ to the line $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}0\\1\\0\end{pmatrix}$.
3. If $\mathbf{p}-\mathbf{a}=\begin{pmatrix}3\\-2\\5\end{pmatrix}$ and $\mathbf{d}=\begin{pmatrix}2\\-1\\2\end{pmatrix}$, find $\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}$.
4. Show that $\begin{pmatrix}-1\\0\\1\end{pmatrix}$ is perpendicular to $\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
5. Find the distance from $P(4,5,2)$ to the line in Question 2.

#### Silver

1. Find the closest point from $P(4,0,5)$ to $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
2. Find the closest point from $P(2,3,4)$ to $\mathbf{r}=\begin{pmatrix}1\\1\\1\end{pmatrix}+t\begin{pmatrix}1\\1\\0\end{pmatrix}$.
3. Find the shortest distance from the origin to $\mathbf{r}=\begin{pmatrix}1\\2\\2\end{pmatrix}+t\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
4. Find the distance from $P(1,2,3)$ to the line through $A(1,0,0)$ and $B(1,4,0)$.
5. For the line $\mathbf{r}=\mathbf{a}+t\mathbf{d}$ and point $P$ with position vector $\mathbf{p}$, explain why the closest point satisfies $(\mathbf{p}-\mathbf{a}-t\mathbf{d})\cdot\mathbf{d}=0$.

#### Gold

1. Prove that the closest point parameter is
   $$
   t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}.
   $$
2. Find the closest point and distance from $P(4,0,5)$ to $\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
3. Find the closest point from $P(0,3,5)$ to $\mathbf{r}=\begin{pmatrix}2\\-1\\0\end{pmatrix}+t\begin{pmatrix}1\\2\\2\end{pmatrix}$ and find the distance.
4. Find the point on $\mathbf{r}=\begin{pmatrix}1\\0\\2\end{pmatrix}+t\begin{pmatrix}2\\1\\-2\end{pmatrix}$ whose position vector is perpendicular to the direction vector of the line.
5. Show that the minimum value of $|\mathbf{p}-\mathbf{a}-t\mathbf{d}|^2$ occurs when $t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}$.

## STEP-style extension questions

These questions are designed to require proof, structure or exact reasoning.

### STEP-style Question 1

Let $A$ and $B$ have position vectors $\mathbf{a}$ and $\mathbf{b}$, where $\mathbf{a}\neq\mathbf{b}$.

Prove that every point on the line through $A$ and $B$ can be written in the form

$$
\mathbf{r}=\mathbf{a}+t(\mathbf{b}-\mathbf{a}).
$$

Explain the meaning of the cases $t=0$, $t=1$ and $0<t<1$.

### STEP-style Question 2

Let

$$
\mathbf{a}=\begin{pmatrix}1\\2\\-2\end{pmatrix}
$$

and

$$
\mathbf{b}=\begin{pmatrix}3\\-1\\2\end{pmatrix}.
$$

Find the acute angle between $\mathbf{a}$ and $\mathbf{b}$ exactly in terms of an inverse cosine, then give the angle to one decimal place.

### STEP-style Question 3

The lines

$$
\ell_1:\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+\lambda\begin{pmatrix}2\\-1\\3\end{pmatrix}
$$

and

$$
\ell_2:\mathbf{r}=\begin{pmatrix}5\\0\\6\end{pmatrix}+\mu\begin{pmatrix}-1\\1\\0\end{pmatrix}
$$

intersect. Find their point of intersection and the acute angle between them.

### STEP-style Question 4

Let

$$
\ell:\mathbf{r}=\mathbf{a}+t\mathbf{d}
$$

and let $P$ have position vector $\mathbf{p}$.

Prove that the point on $\ell$ closest to $P$ has parameter

$$
t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}.
$$

Then apply the formula to find the closest point from $P(4,0,5)$ to

$$
\mathbf{r}=\begin{pmatrix}1\\2\\0\end{pmatrix}+t\begin{pmatrix}2\\-1\\2\end{pmatrix}.
$$

### STEP-style Question 5

Find $k$ such that the two lines

$$
\mathbf{r}=\begin{pmatrix}0\\0\\k\end{pmatrix}+s\begin{pmatrix}1\\2\\1\end{pmatrix}
$$

and

$$
\mathbf{r}=\begin{pmatrix}1\\2\\3\end{pmatrix}+t\begin{pmatrix}2\\5\\-1\end{pmatrix}
$$

intersect. Then find the point of intersection.

## Reflection questions

1. Why is $\overrightarrow{AB}=\mathbf{b}-\mathbf{a}$ rather than $\mathbf{a}-\mathbf{b}$?
2. What does the parameter $t$ mean in $\mathbf{r}=\mathbf{a}+t\mathbf{d}$?
3. Why is the direction vector of a line not unique?
4. How does the scalar product connect algebra and geometry?
5. Why does $\mathbf{a}\cdot\mathbf{b}=0$ mean perpendicular vectors?
6. Why can two non-parallel lines in three dimensions fail to intersect?
7. Why does solving line intersections require all components to agree?
8. Why is the shortest distance from a point to a line perpendicular to the line?
9. What can R show clearly about vector lines that algebra alone may hide?
10. What can exact vector reasoning prove that a diagram can only suggest?

## Answers

### Answers to Exercises 9.1

#### Bronze

1. $\overrightarrow{AB}=\begin{pmatrix}2\\6\end{pmatrix}$.
2. $3$.
3. $\overrightarrow{AB}=\begin{pmatrix}3\\-3\\5\end{pmatrix}$.
4. $\overrightarrow{BA}=\mathbf{a}-\mathbf{b}$.
5. $(3,1,1)$.

#### Silver

1. $\overrightarrow{AB}=\begin{pmatrix}-3\\6\\1\end{pmatrix}$ and $|\overrightarrow{AB}|=\sqrt{46}$.
2. $B(5,1,4)$.
3. $\dfrac{1}{3}\begin{pmatrix}2\\-1\\2\end{pmatrix}$.
4. $P(2,2,1)$.
5. $\overrightarrow{AB}=\begin{pmatrix}2\\-2\\2\end{pmatrix}$ and $\overrightarrow{BC}=\begin{pmatrix}2\\-2\\2\end{pmatrix}$, so the points are collinear.

#### Gold

1. The midpoint of $A$ and $B$ is $\left(\dfrac{1+5}{2},\dfrac{0-2}{2},\dfrac{2+8}{2}\right)=(3,-1,5)$.
2. $k=\pm2$.
3. $CA:AB=1:2$.
4. $x=2$.
5. $\sqrt{61}$.

### Answers to Exercises 9.2

#### Bronze

1. $\mathbf{r}=\begin{pmatrix}1\\2\\3\end{pmatrix}+t\begin{pmatrix}2\\-1\\4\end{pmatrix}$.
2. $\mathbf{r}=\begin{pmatrix}1\\0\\2\end{pmatrix}+t\begin{pmatrix}1\\2\\2\end{pmatrix}$.
3. When $t=2$, the line gives $\begin{pmatrix}5\\-2\\11\end{pmatrix}$, so $P$ lies on the line.
4. Yes. The second vector is $-\dfrac{1}{2}$ times the first.
5. For example, $\mathbf{r}=\begin{pmatrix}5\\-1\\6\end{pmatrix}+s\begin{pmatrix}2\\-1\\3\end{pmatrix}$.

#### Silver

1. $\mathbf{r}=\begin{pmatrix}-1\\2\\4\end{pmatrix}+t\begin{pmatrix}2\\-1\\-1\end{pmatrix}$.
2. $k=3$.
3. They are the same line.
4. $\lambda=3$.
5. $\overrightarrow{AB}=\begin{pmatrix}3\\6\\9\end{pmatrix}$ and $\overrightarrow{AC}=\begin{pmatrix}-1\\-2\\-3\end{pmatrix}$, so the points are collinear.

#### Gold

1. $k=5$.
2. $\overrightarrow{AB}=\begin{pmatrix}3\\-3\\9\end{pmatrix}$ and $\overrightarrow{AP}=\begin{pmatrix}6\\-6\\18\end{pmatrix}=2\overrightarrow{AB}$, so $P$ lies on the line.
3. If the two expressions represent the same point, then $(t-s)\mathbf{d}=\mathbf{0}$. Since $\mathbf{d}\neq\mathbf{0}$, $t-s=0$, so $t=s$.
4. $\lambda=\dfrac{2}{5}$.
5. $\overrightarrow{PQ}=\begin{pmatrix}3\\-6\\9\end{pmatrix}=3\begin{pmatrix}1\\-2\\3\end{pmatrix}$ and $Q$ lies on the given line, so the two descriptions give the same line.

### Answers to Exercises 9.3

#### Bronze

1. $2$.
2. $3$.
3. $90^\circ$.
4. Yes, since the scalar product is $0$.
5. $\dfrac{8}{9}$.

#### Silver

1. $\theta=\cos^{-1}\left(\dfrac{4}{9}\right)\approx63.6^\circ$.
2. $k=\dfrac{5}{2}$.
3. $k=\dfrac{1}{\sqrt3}$.
4. Their scalar product is $4-4+0=0$, so they are perpendicular.
5. $4\sqrt2$.

#### Gold

1. $\theta=\cos^{-1}\left(\dfrac{1}{\sqrt{14}}\right)\approx74.5^\circ$.
2. $k=4$.
3. $k=-\dfrac{1}{2}$.
4. Dividing both sides by $|\mathbf{a}|\,|\mathbf{b}|$ gives $\cos\theta=\dfrac{\mathbf{a}\cdot\mathbf{b}}{|\mathbf{a}|\,|\mathbf{b}|}$.
5. $\cos\angle BAC=\dfrac{2}{3\sqrt6}$, so $\angle BAC\approx74.2^\circ$.

### Answers to Exercises 9.4

#### Bronze

1. $\theta=\cos^{-1}\left(\dfrac{8}{9}\right)\approx27.3^\circ$.
2. Yes, they are parallel.
3. Yes, they are perpendicular.
4. $90^\circ$.
5. $\begin{pmatrix}3\\-1\\4\end{pmatrix}$.

#### Silver

1. $\theta=\cos^{-1}\left(\dfrac{1}{\sqrt{14}}\right)\approx74.5^\circ$.
2. $\theta=\cos^{-1}\left(\dfrac{2}{3}\right)\approx48.2^\circ$.
3. $k=\dfrac{5}{2}$.
4. $k=3$.
5. $\theta=\cos^{-1}\left(\dfrac{7}{9}\right)\approx38.9^\circ$.

#### Gold

1. Their scalar product is $2+2-4=0$, so the lines are perpendicular.
2. $k=\dfrac{3\sqrt{35}}{7}$.
3. Direction vectors describe orientation. The angle between two lines is defined as the angle between their directions, even if the lines are skew and do not meet.
4. $\cos\theta=\dfrac{11}{21}$, so $\theta\approx58.4^\circ$.
5. The direction vectors are $\begin{pmatrix}3\\0\\-3\end{pmatrix}$ and $\begin{pmatrix}2\\1\\2\end{pmatrix}$. Their scalar product is $0$, so the angle is $90^\circ$.

### Answers to Exercises 9.5

#### Bronze

1. $(4,-1,6)$.
2. When $t=3$, the line gives $(4,-1,6)$, so the point lies on the line.
3. The intersection is $(5,2)$.
4. Yes, they are parallel.
5. No. In three dimensions, two non-parallel lines may be skew.

#### Silver

1. The intersection is $(5,0,6)$.
2. Solving the $y$-component gives $s=-1$. The $x$-component then gives $t=-\dfrac{3}{2}$, but the $z$-component is inconsistent. Therefore the lines are skew.
3. They are the same line.
4. $k=2$.
5. The intersection is $(1,2,3)$.

#### Gold

1. $k=5$.
2. The equations are inconsistent, so the lines do not intersect. Since their directions are not scalar multiples, they are skew.
3. If the direction vectors are not parallel, the lines cannot be parallel. If the simultaneous equations are inconsistent, there is no common point. Therefore the lines are skew.
4. The intersection point is $(5,0,6)$. The direction vector to $(6,2,7)$ is $\begin{pmatrix}1\\2\\1\end{pmatrix}$, so the line is $\mathbf{r}=\begin{pmatrix}5\\0\\6\end{pmatrix}+t\begin{pmatrix}1\\2\\1\end{pmatrix}$.
5. The intersection point is $(5,0,6)$. The direction vectors are $\begin{pmatrix}2\\-1\\3\end{pmatrix}$ and $\begin{pmatrix}-1\\1\\0\end{pmatrix}$, so $\cos\theta=\dfrac{3}{\sqrt{28}}$. Hence $\theta\approx55.5^\circ$.

### Answers to Exercises 9.6

#### Bronze

1. $(3,0,0)$.
2. $(1,5,0)$.
3. $2$.
4. The scalar product is $-2+0+2=0$, so the vectors are perpendicular.
5. $\sqrt{13}$.

#### Silver

1. The closest point is $(5,0,4)$.
2. The closest point is $\left(\dfrac{5}{2},\dfrac{5}{2},1\right)$.
3. The shortest distance is $\dfrac{\sqrt{65}}{3}$.
4. The distance is $3$.
5. At the closest point, the vector from the point on the line to $P$ is perpendicular to the line direction. Hence the scalar product with $\mathbf{d}$ is $0$.

#### Gold

1. At the closest point, $(\mathbf{p}-\mathbf{a}-t\mathbf{d})\cdot\mathbf{d}=0$. Expanding gives $(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}-t(\mathbf{d}\cdot\mathbf{d})=0$, so $t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}$.
2. The closest point is $(5,0,4)$ and the distance is $\sqrt2$.
3. The closest point is $\left(\dfrac{34}{9},\dfrac{23}{9},\dfrac{32}{9}\right)$ and the distance is $\dfrac{\sqrt{149}}{3}$.
4. The required point is $\left(\dfrac{13}{9},\dfrac{2}{9},\dfrac{14}{9}\right)$.
5. Expand $|\mathbf{p}-\mathbf{a}-t\mathbf{d}|^2$ as a quadratic in $t$. Its derivative is $-2(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}+2t(\mathbf{d}\cdot\mathbf{d})$. Setting this equal to zero gives the required value of $t$.

### Answers to STEP-style extension questions

1. A point on the line through $A$ and $B$ is obtained by starting at $A$ and moving some scalar multiple of $\overrightarrow{AB}$. Since $\overrightarrow{AB}=\mathbf{b}-\mathbf{a}$, every point has the form $\mathbf{r}=\mathbf{a}+t(\mathbf{b}-\mathbf{a})$. When $t=0$, the point is $A$. When $t=1$, the point is $B$. When $0<t<1$, the point lies between $A$ and $B$.

2. The scalar product is $-3$, and the magnitudes are $3$ and $\sqrt{14}$. The acute angle satisfies
   $$
   \cos\theta=\dfrac{3}{3\sqrt{14}}=\dfrac{1}{\sqrt{14}}.
   $$
   Therefore
   $$
   \theta=\cos^{-1}\left(\dfrac{1}{\sqrt{14}}\right)\approx74.5^\circ.
   $$

3. Equating components gives $\lambda=2$ and $\mu=0$, so the intersection point is $(5,0,6)$. The direction vectors are $\begin{pmatrix}2\\-1\\3\end{pmatrix}$ and $\begin{pmatrix}-1\\1\\0\end{pmatrix}$. Hence
   $$
   \cos\theta=\dfrac{|2(-1)+(-1)(1)+3(0)|}{\sqrt{14}\sqrt2}=\dfrac{3}{\sqrt{28}}.
   $$
   Therefore $\theta\approx55.5^\circ$.

4. At the closest point, the vector from the line to the point is perpendicular to the direction vector. Therefore
   $$
   (\mathbf{p}-\mathbf{a}-t\mathbf{d})\cdot\mathbf{d}=0.
   $$
   Solving gives
   $$
   t=\dfrac{(\mathbf{p}-\mathbf{a})\cdot\mathbf{d}}{\mathbf{d}\cdot\mathbf{d}}.
   $$
   For the given point and line, $t=2$, so the closest point is $(5,0,4)$.

5. Equating components gives
   $$
   s=1+2t,
   $$
   $$
   2s=2+5t,
   $$
   and
   $$
   k+s=3-t.
   $$
   From the first two equations, $t=0$ and $s=1$. Substituting into the third equation gives $k+1=3$, so $k=2$. The point of intersection is $(1,2,3)$.