# Differential equations

```{r}
#| label: setup-differential-equations
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

growth_solution <- function(t, initial_value, rate) {
  initial_value * exp(rate * t)
}

cooling_solution <- function(t, ambient, initial_value, cooling_constant) {
  ambient + (initial_value - ambient) * exp(-cooling_constant * t)
}

logistic_solution <- function(t, capacity, initial_value, rate) {
  capacity / (1 + ((capacity - initial_value) / initial_value) * exp(-rate * t))
}

slope_field <- function(x_min, x_max, y_min, y_max, derivative_function, n_x = 16, n_y = 12, segment_length = 0.18) {
  grid_df <- expand.grid(
    x = seq(x_min, x_max, length.out = n_x),
    y = seq(y_min, y_max, length.out = n_y)
  )
  
  grid_df$slope <- derivative_function(grid_df$x, grid_df$y)
  grid_df$angle <- atan(grid_df$slope)
  grid_df$xend <- grid_df$x + segment_length * cos(grid_df$angle)
  grid_df$yend <- grid_df$y + segment_length * sin(grid_df$angle)
  grid_df$xstart <- grid_df$x - segment_length * cos(grid_df$angle)
  grid_df$ystart <- grid_df$y - segment_length * sin(grid_df$angle)
  
  grid_df
}
```

## Overview

This chapter develops six connected differential-equation ideas:

1. forming differential equations from statements about rates of change;
2. solving separable differential equations;
3. using initial conditions to find particular solutions;
4. modelling growth, decay and saturation;
5. modelling cooling and weather-type relaxation;
6. interpreting deterministic models in contexts such as finance and environmental data.

The central technique in this chapter is **separating the variables**. We will not use integrating factors, second-order differential equations, or stochastic differential equations as solution techniques.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops simple application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

::: {.callout-warning}
Real-world data changes. The R examples in this chapter use small classroom datasets so that the book renders offline. Before using the notes for a live lesson, update the contextual values from official sources such as NIWA, Vietnam's national hydrometeorological service, the Reserve Bank of New Zealand, or a bank exchange-rate feed.
:::

## Inspiration: differential equations behind real-world data

Differential equations are not just symbolic exercises. They are one of the main languages used to describe change in the real world.

In this chapter, our technique is deliberately limited to **separating the variables**. However, the same basic idea sits underneath much more sophisticated models used in weather forecasting, finance, population modelling, epidemics, cooling, chemical reactions and environmental science.

The guiding question is:

> If we know how fast something changes, can we predict what happens next?

Real applications usually begin with data. For example:

- weather observations from New Zealand and Vietnam;
- temperature, rainfall, wind and sunshine summaries;
- exchange-rate data such as NZD/VND or USD/VND;
- population or environmental indices;
- school-based data such as attendance, enrolment, revision completion or assessment readiness.

The full real-world models may be complicated. But in this chapter we study simplified models that still capture the same central idea: **a rate of change determines a future state**.

### From real-world systems to differential equations

| Context | Real-world data students can recognise | More advanced differential-equation model | Separable model used in this chapter |
|---|---|---|---|
| Weather forecasting | Temperature, rainfall, wind speed, pressure | Coupled nonlinear partial differential equations from fluid dynamics and thermodynamics | $\dfrac{dT}{dt}=-k(T-A)$ |
| Cooling and heating | Tea cooling in a room, classroom temperature, outdoor/indoor comparison | Heat equation or energy-balance models | $\dfrac{dT}{dt}=-k(T-A)$ |
| Population growth | Population index, school roll, bacteria growth | Logistic or multi-compartment systems | $\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right)$ |
| Finance | Exchange rates, investment indices, price series | Stochastic differential equations with random noise | $\dfrac{dS}{dt}=rS$ |
| Epidemics | Number infected, recovered or susceptible | Coupled SIR-type differential equations | $\dfrac{dI}{dt}=kI$ or $\dfrac{dI}{dt}=kI(M-I)$ |
| Environmental recovery | Pollution level, rainfall deficit, concentration decay | Transport-diffusion-reaction equations | $\dfrac{dC}{dt}=-kC$ |

The important point is that our simple models are not the whole story. They are **entry points** into the same mathematical way of thinking.

### Example 1: weather and climate

Modern weather forecasting does not use one simple equation. A full numerical weather model may involve velocity, pressure, temperature, moisture, radiation, surface effects and the rotation of the Earth.

A simplified momentum equation may contain terms such as

$$
\dfrac{D\mathbf{u}}{Dt}
+
f\mathbf{k}\times\mathbf{u}
=
-\dfrac{1}{\rho}\nabla p
+
\mathbf{g}
+
\nu\nabla^2\mathbf{u}.
$$

This is far beyond our course. It is a partial differential equation involving vectors, pressure gradients, rotation and diffusion.

But a simple classroom version might ask:

> A hot surface cools towards the surrounding air temperature. How does its temperature change?

That leads to the separable model

$$
\dfrac{dT}{dt}=-k(T-A),
$$

where $A$ is the ambient temperature.

This simple equation already captures an important real-world behaviour: the further $T$ is from $A$, the faster it changes.

### Example 2: finance and exchange rates

Financial data such as NZD/VND or USD/VND exchange rates do not move smoothly. They respond to interest rates, trade, policy, news, expectations and market shocks.

A more advanced stochastic model may be written as

$$
dS_t=\mu S_t\,dt+\sigma S_t\,dW_t.
$$

Here, the term

$$
\mu S_t\,dt
$$

represents the deterministic trend, while

$$
\sigma S_t\,dW_t
$$

represents random fluctuation.

Stochastic differential equations are not part of this chapter. However, if we ignore the random term temporarily, we get the separable deterministic model

$$
\dfrac{dS}{dt}=\mu S.
$$

Solving this gives

$$
S=S_0e^{\mu t}.
$$

So the model in this chapter is a simplified backbone for a much more complex financial idea.

### Example 3: epidemics and saturation

A simple early-stage epidemic model may assume that the number of infected individuals grows at a rate proportional to the current number infected:

$$
\dfrac{dI}{dt}=kI.
$$

This gives exponential growth:

$$
I=I_0e^{kt}.
$$

But unlimited exponential growth is not realistic. In a fixed population, growth slows down as the number of susceptible people decreases.

A more realistic model may involve several coupled equations, such as a susceptible-infected-recovered system. A simplified version in this chapter is logistic growth:

$$
\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right).
$$

This is still separable, but it captures a better idea: growth is fast at first, then slows as the quantity approaches a maximum capacity $M$.

### Visualisation using R

The graph below compares three simplified model shapes. The values are scaled so that the shapes can be compared on one plot.

```{r}
#| label: fig-real-world-de-inspiration
#| fig-cap: "Three simplified differential-equation model shapes."
#| fig-width: 6
#| fig-height: 4

time_values <- seq(0, 12, length.out = 600)

inspiration_df <- data.frame(
  time = rep(time_values, 3),
  context = rep(
    c(
      "cooling towards ambient",
      "deterministic financial growth",
      "logistic saturation"
    ),
    each = length(time_values)
  )
)

cooling_values <- 26 + (80 - 26) * exp(-0.25 * time_values)
finance_values <- 15000 * exp(0.04 * time_values)
logistic_values <- 1000 / (1 + 9 * exp(-0.7 * time_values))

inspiration_df$model_value <- c(
  cooling_values,
  finance_values,
  logistic_values
)

inspiration_df$scaled_value <- ave(
  inspiration_df$model_value,
  inspiration_df$context,
  FUN = function(values) {
    (values - min(values)) / (max(values) - min(values))
  }
)

ggplot(inspiration_df, aes(x = time, y = scaled_value, linetype = context)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Differential equations model different types of change",
    subtitle = "The values are scaled so the shapes can be compared.",
    x = "time",
    y = "scaled model value",
    linetype = "Context"
  ) +
  book_theme()
```
### Real-world example: using a differential equation to model an NZ bank-related share price

Heartland Group Holdings Limited is listed on the NZX under the ticker HGH. Heartland describes its story as one involving specialist finance and banking products, so this gives us a useful New Zealand financial context for a differential-equations model.

The NZX page for HGH gives the following delayed market data:

```{r}
#| label: tbl-hgh-real-nzx-data
#| echo: true

hgh_data <- data.frame(
  source = "NZX delayed market data",
  ticker = "HGH",
  company = "Heartland Group Holdings Limited",
  data_timestamp = "27/06/2026, 16:31 NZST",
  current_price_nzd = 1.2050,
  change_52_week_nzd = 0.4350,
  change_52_week_percent = 56.49
)

hgh_data
```

This gives us two real data points for a simple classroom model.

The current price is approximately

$$
S(1)=1.205.
$$

The 52-week change is approximately

$$
0.435.
$$

So the approximate price one year earlier was

$$
S(0)=1.205-0.435=0.770.
$$

We now model the share price $S$ using the separable differential equation

$$
\dfrac{dS}{dt}=rS,
$$

where $t$ is measured in years from approximately 52 weeks ago.

Separating the variables gives

$$
\dfrac{1}{S}\dfrac{dS}{dt}=r.
$$

So

$$
\int \dfrac{1}{S}\,dS=\int r\,dt.
$$

Hence

$$
\ln S=rt+C.
$$

Therefore,

$$
S=S_0e^{rt}.
$$

Using the real NZX data,

$$
S(0)=0.770
$$

and

$$
S(1)=1.205.
$$

So

$$
1.205=0.770e^r.
$$

Therefore,

$$
e^r=\dfrac{1.205}{0.770}.
$$

Taking logarithms gives

$$
r=\ln\left(\dfrac{1.205}{0.770}\right).
$$

Thus

$$
r\approx0.448.
$$

The fitted differential-equation model is

$$
S(t)=0.770e^{0.448t}.
$$

::: {.callout-warning}
This is not investment advice and not a genuine financial forecast. It is a mathematical modelling example using two real NZX data points. Real share prices are affected by earnings, interest rates, dividends, credit conditions, market sentiment, news and random shocks.
:::

### Visualisation using R: fitting the model to real NZX data

```{r}
#| label: fig-hgh-real-nzx-fit
#| fig-cap: "A separable differential-equation model fitted to real NZX HGH data."
#| fig-width: 6
#| fig-height: 4

price_current <- 1.2050
change_52_week <- 0.4350
price_one_year_ago <- price_current - change_52_week

growth_rate <- log(price_current / price_one_year_ago)

observed_prices <- data.frame(
  t = c(0, 1),
  price = c(price_one_year_ago, price_current),
  label = c("approx. 52 weeks ago", "current NZX price")
)

fitted_curve <- data.frame(
  t = seq(0, 1, length.out = 400)
)

fitted_curve$price <- price_one_year_ago * exp(growth_rate * fitted_curve$t)

ggplot() +
  geom_line(data = fitted_curve, aes(x = t, y = price), linewidth = 1.1) +
  geom_point(data = observed_prices, aes(x = t, y = price), size = 3) +
  geom_text(
    data = observed_prices,
    aes(x = t, y = price, label = label),
    nudge_y = 0.05
  ) +
  labs(
    title = "Fitting dS/dt = rS to real NZX HGH data",
    subtitle = paste0("Estimated continuous annual rate r = ", round(growth_rate, 3)),
    x = "time in years from approximately 52 weeks ago",
    y = "share price in NZD"
  ) +
  book_theme()
```

### A light prediction

A full one-year extrapolation would be too aggressive for a classroom model based on only two data points. Instead, we make a light 13-week projection.

Since

$$
13\text{ weeks}=\dfrac{13}{52}\text{ years}=0.25\text{ years},
$$

we estimate

$$
S(1.25)=1.205e^{0.448(0.25)}.
$$

This gives

$$
S(1.25)\approx1.348.
$$

So the simple model gives a light 13-week projection of approximately

$$
\$1.35.
$$

This should be interpreted as a mathematical projection, not a financial prediction.

```{r}
#| label: tbl-hgh-light-prediction
#| echo: true

prediction_weeks <- c(4, 8, 13)
prediction_years <- prediction_weeks / 52

prediction_table <- data.frame(
  weeks_after_current_price = prediction_weeks,
  years_after_current_price = round(prediction_years, 3),
  predicted_price_nzd = round(price_current * exp(growth_rate * prediction_years), 3)
)

prediction_table
```

### Visualisation using R: light prediction region

```{r}
#| label: fig-hgh-light-prediction
#| fig-cap: "A light 13-week projection from the fitted differential-equation model."
#| fig-width: 6
#| fig-height: 4

prediction_curve <- data.frame(
  t = seq(0, 1.25, length.out = 600)
)

prediction_curve$price <- price_one_year_ago * exp(growth_rate * prediction_curve$t)
prediction_curve$interval <- ifelse(
  prediction_curve$t <= 1,
  "observed 52-week interval",
  "13-week projection"
)

future_point <- data.frame(
  t = 1.25,
  price = price_current * exp(growth_rate * 0.25),
  label = "13-week projection"
)

ggplot() +
  annotate(
    "rect",
    xmin = 1,
    xmax = 1.25,
    ymin = -Inf,
    ymax = Inf,
    alpha = 0.12
  ) +
  geom_line(
    data = prediction_curve,
    aes(x = t, y = price, linetype = interval),
    linewidth = 1.1
  ) +
  geom_point(data = observed_prices, aes(x = t, y = price), size = 3) +
  geom_point(data = future_point, aes(x = t, y = price), size = 3) +
  geom_text(
    data = future_point,
    aes(x = t, y = price, label = label),
    nudge_y = 0.05
  ) +
  labs(
    title = "Light projection from dS/dt = rS",
    subtitle = "The shaded region shows the 13-week extrapolation beyond the real NZX data.",
    x = "time in years from approximately 52 weeks ago",
    y = "share price in NZD",
    linetype = "Interval"
  ) +
  book_theme()
```

### Scenario check

Because the fitted value

$$
r\approx0.448
$$

comes from a strong 52-week movement, it is risky to assume that this rate continues. A more honest modelling approach is to compare scenarios.

```{r}
#| label: fig-hgh-real-nzx-scenario-check
#| fig-cap: "Scenario comparison after the current NZX price."
#| fig-width: 6
#| fig-height: 4

future_time <- seq(0, 0.25, length.out = 300)

scenario_df <- expand.grid(
  years_after_current_price = future_time,
  scenario = c(
    "same fitted rate",
    "half fitted rate",
    "zero growth",
    "negative half fitted rate"
  )
)

scenario_df$rate <- ifelse(
  scenario_df$scenario == "same fitted rate",
  growth_rate,
  ifelse(
    scenario_df$scenario == "half fitted rate",
    growth_rate / 2,
    ifelse(
      scenario_df$scenario == "zero growth",
      0,
      -growth_rate / 2
    )
  )
)

scenario_df$predicted_price <- price_current * exp(
  scenario_df$rate * scenario_df$years_after_current_price
)

ggplot(scenario_df, aes(x = years_after_current_price, y = predicted_price, linetype = scenario)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = price_current, linetype = "dashed") +
  labs(
    title = "Scenario check for a 13-week projection",
    subtitle = "The model is sensitive to the assumed continuous rate r.",
    x = "years after current NZX price",
    y = "projected share price in NZD",
    linetype = "Scenario"
  ) +
  book_theme()
```

### Interpretation

This example follows the modelling cycle:

$$
\text{real NZX data}
\quad\longrightarrow\quad
\dfrac{dS}{dt}=rS
\quad\longrightarrow\quad
S=S_0e^{rt}
\quad\longrightarrow\quad
\text{fit }r
\quad\longrightarrow\quad
\text{make a cautious projection}.
$$

The strength of the differential equation is that the assumption is explicit:

> The price changes at a rate proportional to the current price.

Once that assumption is made, the mathematics forces an exponential model. The final step is not just calculation, but critique: we ask whether the assumption is reasonable.

### Why we still study simple separable models

A simple separable differential equation is useful because it teaches the core modelling cycle:

1. identify the changing quantity;
2. describe its rate of change;
3. write a differential equation;
4. separate the variables;
5. integrate;
6. use an initial condition;
7. interpret the solution in context.

The models in this chapter are simplified, but they are not artificial. They are the first layer of the same mathematical thinking used in larger models.

For example,

$$
\dfrac{dT}{dt}=-k(T-A)
$$

is not a full weather model, but it captures relaxation towards an ambient value.

Similarly,

$$
\dfrac{dS}{dt}=rS
$$

is not a full financial model, but it captures the deterministic backbone of continuous growth or decay.

And

$$
\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right)
$$

is not a full population or epidemic model, but it captures the idea that growth slows as a limiting capacity is approached.

This chapter therefore acts as a bridge:

$$
\text{data}
\quad\longrightarrow\quad
\text{rate of change}
\quad\longrightarrow\quad
\text{differential equation}
\quad\longrightarrow\quad
\text{solution}
\quad\longrightarrow\quad
\text{interpretation}.
$$

## 10.1 Forming differential equations from rates of change

### Key mathematical ideas

A differential equation describes how a quantity changes.

If a quantity $y$ changes with respect to $x$, then its rate of change is

$$
\dfrac{dy}{dx}.
$$

Many modelling statements translate directly into differential equations.

If $y$ increases at a rate proportional to $y$, then

$$
\dfrac{dy}{dx}=ky,
$$

where $k>0$.

If $y$ decreases at a rate proportional to $y$, then

$$
\dfrac{dy}{dx}=-ky,
$$

where $k>0$.

If a temperature $T$ approaches an ambient temperature $A$, then a common model is

$$
\dfrac{dT}{dt}=-k(T-A),
$$

where $k>0$.

This means the rate of temperature change depends on the difference between the current temperature and the ambient temperature.

### Visualisation using R

```{r}
#| label: fig-growth-slope-field
#| fig-cap: "Slope field for dy/dx = 0.5y."
#| fig-width: 6
#| fig-height: 4

field_df <- slope_field(
  x_min = 0,
  x_max = 4,
  y_min = 0,
  y_max = 6,
  derivative_function = function(x_value, y_value) 0.5 * y_value
)

curve_df <- data.frame(
  x = seq(0, 4, length.out = 400)
)

curve_df$y <- 2 * exp(0.5 * curve_df$x)

ggplot() +
  geom_segment(
    data = field_df,
    aes(x = xstart, y = ystart, xend = xend, yend = yend),
    alpha = 0.55
  ) +
  geom_line(data = curve_df, aes(x = x, y = y), linewidth = 1.1) +
  labs(
    title = "A rate proportional to the current amount",
    subtitle = "When dy/dx = 0.5y, larger y-values have steeper gradients.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

A cup of tea has temperature $T$ degrees Celsius after $t$ minutes. The surrounding air temperature is $22^\circ\text{C}$. The tea cools at a rate proportional to how much hotter it is than the surrounding air.

Form a differential equation.

The temperature difference is

$$
T-22.
$$

Since the tea is cooling, the rate of change is negative when $T>22$.

Therefore,

$$
\dfrac{dT}{dt}=-k(T-22),
\quad k>0.
$$

This equation also has the sensible equilibrium

$$
T=22,
$$

because when $T=22$,

$$
\dfrac{dT}{dt}=0.
$$

### Short investigation

The following table translates modelling language into differential equations.

```{r}
#| label: tbl-rate-language
#| echo: true

data.frame(
  statement = c(
    "growth proportional to amount",
    "decay proportional to amount",
    "cooling towards ambient A",
    "saturation towards capacity M",
    "logistic growth with capacity M"
  ),
  differential_equation = c(
    "dy/dt = ky",
    "dy/dt = -ky",
    "dT/dt = -k(T - A)",
    "dW/dt = k(M - W)",
    "dP/dt = kP(1 - P/M)"
  )
)
```

### Exercises 10.1

#### Bronze

1. A quantity $y$ increases at a rate proportional to $y$. Write a differential equation.
2. A quantity $y$ decreases at a rate proportional to $y$. Write a differential equation.
3. A population $P$ decreases at $5\%$ per hour. Write a differential equation.
4. A speed $v$ decreases at a rate proportional to $v^2$. Write a differential equation.
5. A water level $W$ rises towards a maximum level $M$ at a rate proportional to $M-W$. Write a differential equation.

#### Silver

1. A drink of temperature $T$ cools in a room of temperature $22^\circ\text{C}$. Its rate of cooling is proportional to $T-22$. Write a differential equation.
2. A number of infected individuals $I$ grows at a rate proportional to $I(1000-I)$. Write a differential equation.
3. A deterministic exchange-rate model assumes a value $S$ changes at a rate proportional to $S$. Write a differential equation.
4. A rainfall deficit $D$ decreases at a rate proportional to $D$. Write a differential equation.
5. A population satisfies $\dfrac{dN}{dt}=kN$ and has initial value $200$. Write the initial condition.

#### Gold

1. A cold object warms towards ambient temperature $30^\circ\text{C}$ at a rate proportional to the temperature difference. Write a differential equation.
2. The rate of increase of $y$ with respect to $x$ is proportional to $xy$. Write a differential equation.
3. The rate of decrease of $y$ with respect to $x$ is proportional to $\dfrac{y}{1+x}$. Write a differential equation.
4. A population $P$ grows logistically with carrying capacity $5000$. Write a suitable differential equation.
5. Explain why $T=A$ is an equilibrium solution of $\dfrac{dT}{dt}=-k(T-A)$.

## 10.2 Solving separable differential equations

### Key mathematical ideas

A first-order differential equation is **separable** if it can be rearranged so that all the $y$ terms are on one side and all the $x$ terms are on the other.

For example,

$$
\dfrac{dy}{dx}=xy
$$

can be rearranged as

$$
\dfrac{1}{y}\dfrac{dy}{dx}=x.
$$

This leads to

$$
\int \dfrac{1}{y}\,dy=\int x\,dx.
$$

So

$$
\ln|y|=\dfrac{x^2}{2}+C.
$$

A common final form is

$$
y=Ae^{\dfrac{x^2}{2}},
$$

where $A$ is a non-zero constant.

### Visualisation using R

```{r}
#| label: fig-solution-family-2xy
#| fig-cap: "A family of solutions to dy/dx = 2xy."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-2, 2, length.out = 600)

solution_df <- expand.grid(
  x = x_values,
  initial_value = c(1, 2, 4)
)

solution_df$y <- solution_df$initial_value * exp(solution_df$x^2)
solution_df$curve_label <- paste0("y(0) = ", solution_df$initial_value)

ggplot(solution_df, aes(x = x, y = y, linetype = curve_label)) +
  geom_line(linewidth = 1.1) +
  coord_cartesian(ylim = c(0, 25)) +
  labs(
    title = "Solution family for dy/dx = 2xy",
    subtitle = "Different initial conditions give different curves.",
    x = "x",
    y = "y",
    linetype = "Initial condition"
  ) +
  book_theme()
```

### Worked example

Solve

$$
\dfrac{dy}{dx}=2xy.
$$

Separate the variables:

$$
\dfrac{1}{y}\dfrac{dy}{dx}=2x.
$$

Integrate both sides:

$$
\int \dfrac{1}{y}\,dy
=
\int 2x\,dx.
$$

So

$$
\ln|y|=x^2+C.
$$

Exponentiating gives

$$
|y|=e^{x^2+C}=e^C e^{x^2}.
$$

So the general solution may be written as

$$
y=Ae^{x^2},
$$

where $A$ is a constant.

### Short investigation

Check the solution numerically by differentiating the expression in R.

```{r}
#| label: tbl-solution-check-2xy
#| echo: true

x_values <- c(-1, 0, 1, 2)
y_values <- 3 * exp(x_values^2)
dy_dx_values <- 6 * x_values * exp(x_values^2)
right_side_values <- 2 * x_values * y_values

data.frame(
  x = x_values,
  y = round(y_values, 5),
  derivative = round(dy_dx_values, 5),
  right_side = round(right_side_values, 5)
)
```

### Exercises 10.2

#### Bronze

1. Solve $\dfrac{dy}{dx}=3x^2$.
2. Solve $\dfrac{dy}{dx}=2xy$.
3. Solve $\dfrac{dy}{dx}=y$.
4. Solve $\dfrac{dy}{dx}=\dfrac{x}{y}$.
5. Solve $\dfrac{dy}{dx}=4xy^2$.

#### Silver

1. Solve $\dfrac{dy}{dx}=(x+1)y$, given that $y=2$ when $x=0$.
2. Solve $\dfrac{dy}{dx}=\dfrac{3x}{y}$, given that $y=4$ when $x=0$.
3. Solve $\dfrac{dy}{dx}=\dfrac{y}{1+x}$, given that $y=5$ when $x=0$.
4. Solve $\dfrac{dy}{dx}=2x(1+y)$, given that $y=0$ when $x=0$.
5. Solve $\dfrac{dy}{dx}=\dfrac{6x}{y+1}$, given that $y=2$ when $x=0$.

#### Gold

1. Solve $\dfrac{dy}{dx}=xy^2$, given that $y=1$ when $x=0$.
2. Solve $\dfrac{dy}{dx}=y\cos x$, given that $y=3$ when $x=0$.
3. Solve $\dfrac{dy}{dx}=\dfrac{1+x}{y}$, given that $y=1$ when $x=0$.
4. Solve $\dfrac{dy}{dx}=e^xy^2$, given that $y=2$ when $x=0$.
5. Solve $\dfrac{dy}{dx}=(2x+1)(y-4)$, given that $y=5$ when $x=0$.

## 10.3 Particular solutions and interpretation

### Key mathematical ideas

A **general solution** contains an arbitrary constant.

A **particular solution** uses extra information, usually an initial condition, to find that constant.

For example, the general solution

$$
y=Ae^{x^2}
$$

becomes a particular solution if we know the value of $y$ at one point.

If

$$
y(0)=7,
$$

then

$$
7=Ae^0=A.
$$

So the particular solution is

$$
y=7e^{x^2}.
$$

### Visualisation using R

```{r}
#| label: fig-particular-solution-through-point
#| fig-cap: "The solution y = 7e^(x^2) passes through (0, 7)."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-1.5, 1.5, length.out = 600)

curve_df <- data.frame(
  x = x_values,
  y = 7 * exp(x_values^2)
)

point_df <- data.frame(
  x = 0,
  y = 7,
  label = "(0, 7)"
)

ggplot() +
  geom_line(data = curve_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = point_df, aes(x = x, y = y), size = 3) +
  geom_text(data = point_df, aes(x = x, y = y, label = label), nudge_y = 3) +
  labs(
    title = "Particular solution",
    subtitle = "The initial condition selects one curve from the family.",
    x = "x",
    y = "y"
  ) +
  book_theme()
```

### Worked example

Solve

$$
\dfrac{dy}{dx}=\dfrac{3y}{1+x},
\quad y(0)=2.
$$

Separate the variables:

$$
\dfrac{1}{y}\dfrac{dy}{dx}=\dfrac{3}{1+x}.
$$

Integrate:

$$
\int \dfrac{1}{y}\,dy
=
\int \dfrac{3}{1+x}\,dx.
$$

So

$$
\ln|y|=3\ln|1+x|+C.
$$

For the solution through $x=0$, we work on the interval

$$
x>-1.
$$

Thus

$$
\ln|y|=\ln(1+x)^3+C.
$$

Hence

$$
y=A(1+x)^3.
$$

Using $y(0)=2$ gives

$$
2=A.
$$

Therefore,

$$
y=2(1+x)^3,
\quad x>-1.
$$

### Short investigation

Use R to compare several particular solutions.

```{r}
#| label: fig-particular-solution-family
#| fig-cap: "Particular solutions of dy/dx = 3y/(1+x)."
#| fig-width: 6
#| fig-height: 4

x_values <- seq(-0.9, 2, length.out = 600)

solution_df <- expand.grid(
  x = x_values,
  initial_value = c(1, 2, 4)
)

solution_df$y <- solution_df$initial_value * (1 + solution_df$x)^3
solution_df$curve_label <- paste0("y(0) = ", solution_df$initial_value)

ggplot(solution_df, aes(x = x, y = y, linetype = curve_label)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Initial conditions change the constant",
    subtitle = "The shape is controlled by the differential equation; the starting value selects the curve.",
    x = "x",
    y = "y",
    linetype = "Initial condition"
  ) +
  book_theme()
```

### Exercises 10.3

#### Bronze

1. The general solution is $y=Ae^{2x}$. Find $A$ if $y(0)=5$.
2. The general solution is $y=A(1+x)^3$. Find $A$ if $y(0)=4$.
3. The general solution is $y^2=x^2+C$. Find $C$ if $y(0)=3$.
4. The general solution is $\ln y=x+C$. Find $y$ if $y(0)=2$.
5. The general solution is $y=Ae^{-3t}$. Find $A$ if $y(0)=80$.

#### Silver

1. Solve $\dfrac{dy}{dx}=4y$, given $y(0)=3$.
2. Solve $\dfrac{dy}{dx}=xy$, given $y(0)=2$.
3. Solve $\dfrac{dy}{dx}=\dfrac{2y}{1+x}$, given $y(0)=5$.
4. Solve $\dfrac{dy}{dx}=\dfrac{4x}{y}$, given $y(0)=1$.
5. Solve $\dfrac{dy}{dx}=x(y+2)$, given $y(0)=0$.

#### Gold

1. Solve $\dfrac{dy}{dx}=3x^2(y-1)$, given $y(0)=2$.
2. Solve $\dfrac{dy}{dx}=\dfrac{2x+1}{y}$, given $y(0)=3$.
3. Solve $\dfrac{dy}{dx}=y\tan x$, given $y(0)=4$.
4. Solve $\dfrac{dy}{dx}=x(y-3)^2$, given $y(0)=4$.
5. Find the solution of $\dfrac{dy}{dx}=ky$, $y(0)=100$, if $y(5)=200$.

## 10.4 Growth, decay and saturation models

### Key mathematical ideas

The simplest growth model is

$$
\dfrac{dP}{dt}=kP,
\quad k>0.
$$

Separating variables gives

$$
\dfrac{1}{P}\dfrac{dP}{dt}=k.
$$

So

$$
\ln P=kt+C,
$$

and therefore

$$
P=P_0e^{kt}.
$$

The simplest decay model is

$$
\dfrac{dP}{dt}=-kP,
\quad k>0,
$$

with solution

$$
P=P_0e^{-kt}.
$$

Growth proportional to current amount is a useful first model, but it cannot grow forever in a restricted environment. A saturation model has the form

$$
\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right),
$$

where $M$ is the carrying capacity.

### Visualisation using R

```{r}
#| label: fig-growth-decay-saturation
#| fig-cap: "Exponential growth, exponential decay and logistic saturation."
#| fig-width: 6
#| fig-height: 4

t_values <- seq(0, 12, length.out = 600)

model_df <- data.frame(
  t = rep(t_values, 3),
  model = rep(c("growth", "decay", "saturation"), each = length(t_values))
)

model_df$value <- c(
  growth_solution(t_values, initial_value = 100, rate = 0.18),
  growth_solution(t_values, initial_value = 100, rate = -0.18),
  logistic_solution(t_values, capacity = 1000, initial_value = 100, rate = 0.45)
)

ggplot(model_df, aes(x = t, y = value, linetype = model)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Three model shapes",
    subtitle = "Growth, decay and saturation have distinct long-term behaviour.",
    x = "time",
    y = "quantity",
    linetype = "Model"
  ) +
  book_theme()
```

### Worked example

A quantity $P$ grows at a rate proportional to $P$. Initially,

$$
P(0)=500.
$$

After $10$ years,

$$
P(10)=700.
$$

Find a model for $P$.

Since

$$
\dfrac{dP}{dt}=kP,
$$

we have

$$
P=500e^{kt}.
$$

Using $P(10)=700$ gives

$$
700=500e^{10k}.
$$

Thus

$$
e^{10k}=\dfrac{7}{5}.
$$

So

$$
10k=\ln\left(\dfrac{7}{5}\right),
$$

and

$$
k=\dfrac{1}{10}\ln\left(\dfrac{7}{5}\right).
$$

Therefore,

$$
P=500e^{\dfrac{t}{10}\ln(\dfrac{7}{5})}.
$$

Equivalently,

$$
P=500\left(\dfrac{7}{5}\right)^{t/10}.
$$

### Short investigation: NZ and Vietnam context

The following small dataset is a classroom-style structure for comparing a deterministic model with updated observations. Replace the values with current official values before using the activity live.

```{r}
#| label: fig-context-growth-data
#| fig-cap: "A deterministic curve compared with a small classroom dataset."
#| fig-width: 6
#| fig-height: 4

context_df <- data.frame(
  year = 0:5,
  index_value = c(100, 106, 113, 121, 128, 137)
)

rate_estimate <- log(context_df$index_value[6] / context_df$index_value[1]) / 5

curve_df <- data.frame(
  year = seq(0, 5, length.out = 400)
)

curve_df$model <- context_df$index_value[1] * exp(rate_estimate * curve_df$year)

ggplot() +
  geom_point(data = context_df, aes(x = year, y = index_value), size = 3) +
  geom_line(data = curve_df, aes(x = year, y = model), linewidth = 1.1) +
  labs(
    title = "Fitting a simple exponential model",
    subtitle = "This structure can be reused with updated NZ, Vietnam or school data.",
    x = "years from start",
    y = "index"
  ) +
  book_theme()
```

### Exercises 10.4

#### Bronze

1. Solve $\dfrac{dP}{dt}=0.2P$, given $P(0)=50$.
2. Solve $\dfrac{dP}{dt}=-0.3P$, given $P(0)=80$.
3. A quantity satisfies $P=100e^{0.05t}$. Find $\dfrac{dP}{dt}$ in terms of $P$.
4. A quantity satisfies $P=200e^{-0.1t}$. Find its half-life.
5. For $\dfrac{dP}{dt}=kP$, explain the meaning of $k>0$ and $k<0$.

#### Silver

1. A population grows from $500$ to $700$ in $10$ years according to $\dfrac{dP}{dt}=kP$. Find $k$.
2. A value decays from $1200$ to $900$ in $4$ years according to $\dfrac{dV}{dt}=-kV$. Find $k$.
3. A bacteria culture satisfies $\dfrac{dN}{dt}=0.4N$, $N(0)=200$. Find $N(5)$.
4. A pollutant level satisfies $\dfrac{dC}{dt}=-0.25C$, $C(0)=60$. Find the time when $C=10$.
5. A quantity doubles every $8$ years. Find the exponential growth constant.

#### Gold

1. A quantity satisfies $\dfrac{dP}{dt}=kP$ and triples in $12$ years. Find $k$.
2. A quantity satisfies $\dfrac{dP}{dt}=-kP$ and halves in $6$ years. Find $k$.
3. Solve $\dfrac{dP}{dt}=0.4P\left(1-\dfrac{P}{1000}\right)$, given $P(0)=100$.
4. For the logistic model in Question 3, find the time when $P=800$.
5. Show that the logistic model $\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right)$ has equilibrium solutions $P=0$ and $P=M$.

## 10.5 Cooling and weather-type relaxation models

### Key mathematical ideas

Newton's law of cooling has the form

$$
\dfrac{dT}{dt}=-k(T-A),
$$

where:

- $T$ is the object's temperature;
- $A$ is the ambient temperature;
- $k>0$ is the cooling constant.

This model is separable:

$$
\dfrac{1}{T-A}\dfrac{dT}{dt}=-k.
$$

So

$$
\ln|T-A|=-kt+C.
$$

Therefore,

$$
T=A+Ce^{-kt}.
$$

If $T(0)=T_0$, then

$$
T=A+(T_0-A)e^{-kt}.
$$

This form also models many **relaxation** processes: a quantity moves gradually towards a long-term level.

### Visualisation using R

```{r}
#| label: fig-cooling-towards-ambient
#| fig-cap: "Cooling towards ambient temperature."
#| fig-width: 6
#| fig-height: 4

time_values <- seq(0, 40, length.out = 600)

cooling_df <- data.frame(
  time = time_values,
  temperature = cooling_solution(
    time_values,
    ambient = 26,
    initial_value = 80,
    cooling_constant = log(9 / 4) / 10
  )
)

ggplot(cooling_df, aes(x = time, y = temperature)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 26, linetype = "dashed") +
  labs(
    title = "Newton cooling model",
    subtitle = "The temperature approaches the ambient level without crossing it.",
    x = "time in minutes",
    y = "temperature in degrees Celsius"
  ) +
  book_theme()
```

### Worked example

A hot drink is placed in a room at temperature

$$
26^\circ\text{C}.
$$

Initially, the drink is at

$$
80^\circ\text{C}.
$$

After $10$ minutes, its temperature is

$$
50^\circ\text{C}.
$$

Assume

$$
\dfrac{dT}{dt}=-k(T-26).
$$

Find $T$ in terms of $t$.

The general solution is

$$
T=26+Ce^{-kt}.
$$

Using $T(0)=80$ gives

$$
80=26+C,
$$

so

$$
C=54.
$$

Thus

$$
T=26+54e^{-kt}.
$$

Using $T(10)=50$ gives

$$
50=26+54e^{-10k}.
$$

So

$$
24=54e^{-10k},
$$

and

$$
e^{-10k}=\dfrac{4}{9}.
$$

Taking logarithms,

$$
-10k=\ln\left(\dfrac{4}{9}\right).
$$

Hence

$$
k=\dfrac{1}{10}\ln\left(\dfrac{9}{4}\right).
$$

Therefore,

$$
T=26+54e^{-\dfrac{t}{10}\ln(\dfrac{9}{4})}.
$$

### Short investigation: comparing ambient settings

The same drink cools differently in different ambient conditions.

```{r}
#| label: fig-nz-vietnam-ambient-comparison
#| fig-cap: "Cooling curves under two illustrative ambient temperatures."
#| fig-width: 6
#| fig-height: 4

time_values <- seq(0, 45, length.out = 600)

ambient_df <- expand.grid(
  time = time_values,
  place = c("cooler ambient", "warmer ambient")
)

ambient_df$ambient <- ifelse(ambient_df$place == "cooler ambient", 16, 30)
ambient_df$temperature <- cooling_solution(
  ambient_df$time,
  ambient = ambient_df$ambient,
  initial_value = 80,
  cooling_constant = 0.08
)

ggplot(ambient_df, aes(x = time, y = temperature, linetype = place)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Same cooling constant, different ambient temperatures",
    subtitle = "This is a useful classroom link to weather data from NZ and Vietnam.",
    x = "time in minutes",
    y = "temperature in degrees Celsius",
    linetype = "Context"
  ) +
  book_theme()
```

### Exercises 10.5

#### Bronze

1. Solve $\dfrac{dT}{dt}=-0.2(T-20)$, given $T(0)=80$.
2. Solve $\dfrac{dT}{dt}=-k(T-25)$, given $T(0)=70$.
3. A cooling model is $T=22+58e^{-0.1t}$. Find $T(0)$.
4. A cooling model is $T=20+60e^{-0.05t}$. What is the long-term temperature?
5. Explain why $T=A$ is a constant solution of $\dfrac{dT}{dt}=-k(T-A)$.

#### Silver

1. A drink cools in a room at $20^\circ\text{C}$. Initially $T=90$. After $5$ minutes, $T=60$. Find $k$.
2. For Question 1, find the time when $T=30$.
3. Solve $\dfrac{dT}{dt}=-0.1(T-18)$, given $T(0)=28$.
4. A temperature warms towards $30^\circ\text{C}$ according to $\dfrac{dT}{dt}=0.2(30-T)$, with $T(0)=10$. Find $T(t)$.
5. A weather-index model satisfies $\dfrac{dW}{dt}=-0.3(W-100)$, $W(0)=70$. Find $W(t)$.

#### Gold

1. A hot drink cools in ambient temperature $26^\circ\text{C}$. Initially $T=80$, and after $10$ minutes $T=50$. Find $T(t)$.
2. For Question 1, find the time when $T=30$.
3. A cold object warms towards $32^\circ\text{C}$. Initially it is $12^\circ\text{C}$ and after $5$ minutes it is $22^\circ\text{C}$. Find the model.
4. Show that if $T=A+(T_0-A)e^{-kt}$, then it satisfies $\dfrac{dT}{dt}=-k(T-A)$.
5. In the model $T=A+Ce^{-kt}$ with $k>0$, explain why the curve approaches $A$ but never reaches it in finite time unless $C=0$.

## 10.6 Deterministic finance models and stochastic context

### Key mathematical ideas

A simple deterministic finance model assumes a value $S$ changes at a rate proportional to itself:

$$
\dfrac{dS}{dt}=rS.
$$

Separating variables gives

$$
\dfrac{1}{S}\dfrac{dS}{dt}=r.
$$

Thus

$$
\ln S=rt+C,
$$

and

$$
S=S_0e^{rt}.
$$

This is the deterministic backbone behind many compound-growth models.

Real financial data are not smooth. Exchange rates, asset prices and interest-rate-sensitive values include random shocks. A stochastic model may add a random component, but stochastic differential equations are beyond this chapter.

In this chapter, we use only the deterministic separable model.

### Visualisation using R

```{r}
#| label: fig-deterministic-finance-vs-noisy-paths
#| fig-cap: "A deterministic exponential curve with simulated noisy paths."
#| fig-width: 6
#| fig-height: 4

set.seed(9709)

time_values <- seq(0, 1, length.out = 252)
initial_rate <- 15000
annual_drift <- 0.02
annual_volatility <- 0.08

deterministic_df <- data.frame(
  time = time_values,
  rate_value = growth_solution(time_values, initial_rate, annual_drift),
  path = "deterministic"
)

simulated_paths <- do.call(
  rbind,
  lapply(1:6, function(path_id) {
    random_steps <- rnorm(length(time_values) - 1, mean = 0, sd = sqrt(1 / 252))
    log_values <- numeric(length(time_values))
    log_values[1] <- log(initial_rate)
    
    for (i in 2:length(time_values)) {
      log_values[i] <- log_values[i - 1] +
        (annual_drift - 0.5 * annual_volatility^2) / 252 +
        annual_volatility * random_steps[i - 1]
    }
    
    data.frame(
      time = time_values,
      rate_value = exp(log_values),
      path = paste0("simulated path ", path_id)
    )
  })
)

ggplot() +
  geom_line(
    data = simulated_paths,
    aes(x = time, y = rate_value, group = path),
    alpha = 0.35
  ) +
  geom_line(
    data = deterministic_df,
    aes(x = time, y = rate_value),
    linewidth = 1.2
  ) +
  labs(
    title = "Deterministic model and stochastic-looking variation",
    subtitle = "Only the smooth deterministic model is solved in this chapter.",
    x = "time in years",
    y = "index value"
  ) +
  book_theme()
```

### Worked example

A deterministic exchange-rate index $S$ satisfies

$$
\dfrac{dS}{dt}=0.02S,
$$

where $t$ is measured in years.

Suppose

$$
S(0)=15000.
$$

Find $S(t)$ and estimate $S(0.5)$.

Since

$$
\dfrac{dS}{dt}=0.02S,
$$

we have

$$
S=15000e^{0.02t}.
$$

At $t=0.5$,

$$
S(0.5)=15000e^{0.01}.
$$

Thus

$$
S(0.5)\approx15150.75.
$$

This is not a forecast. It is a simple deterministic model for practising separable differential equations.

### Short investigation

Use the deterministic formula to compare different annual rates.

```{r}
#| label: tbl-deterministic-finance-values
#| echo: true

rates <- c(-0.03, 0, 0.02, 0.05)

data.frame(
  annual_rate = rates,
  value_after_half_year = round(15000 * exp(rates * 0.5), 2),
  value_after_one_year = round(15000 * exp(rates), 2)
)
```

### Exercises 10.6

#### Bronze

1. Solve $\dfrac{dS}{dt}=0.04S$, given $S(0)=1000$.
2. Solve $\dfrac{dS}{dt}=-0.02S$, given $S(0)=5000$.
3. If $S=15000e^{0.02t}$, find $S(1)$.
4. If $S=100e^{-0.1t}$, find the percentage rate of continuous decrease.
5. Explain why $S=S_0e^{rt}$ is deterministic.

#### Silver

1. A value grows continuously at $3\%$ per year from $2000$. Find its value after $5$ years.
2. A value decreases continuously at $4\%$ per year from $8000$. Find its value after $2$ years.
3. A value changes from $15000$ to $15150.75$ in half a year under $\dfrac{dS}{dt}=rS$. Find $r$.
4. Find the time for $S=1000e^{0.06t}$ to reach $1500$.
5. Find the time for $S=2000e^{-0.08t}$ to fall to $1200$.

#### Gold

1. A deterministic index satisfies $\dfrac{dS}{dt}=rS$, $S(0)=S_0$. Prove that $S=S_0e^{rt}$.
2. A value doubles in $3$ years under continuous growth. Find $r$.
3. A value loses $20\%$ in $4$ years under continuous decay. Find the decay constant $k$.
4. Explain why a deterministic exponential model may be a poor forecast for daily exchange rates.
5. A model satisfies $\dfrac{dS}{dt}=rS$ and $S(1)=18000$, $S(3)=22000$. Find $r$ and $S(0)$.

## STEP-style extension questions

These questions are designed to require proof, structure or exact reasoning.

### STEP-style Question 1

Solve

$$
\dfrac{dy}{dx}=xy,
\quad y(0)=2.
$$

Then show directly by differentiation that your answer satisfies the differential equation.

### STEP-style Question 2

Solve

$$
\dfrac{dy}{dx}=x(y-1)^2,
\quad y(0)=2.
$$

State the largest interval containing $0$ on which your solution is defined.

### STEP-style Question 3

A quantity $P$ satisfies the logistic differential equation

$$
\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{M}\right),
$$

where $k>0$ and $M>0$.

Show that the general solution may be written as

$$
P=\dfrac{M}{1+Ae^{-kt}}.
$$

### STEP-style Question 4

A hot object cools according to

$$
\dfrac{dT}{dt}=-k(T-A).
$$

Given $T(0)=T_0$, prove that

$$
T=A+(T_0-A)e^{-kt}.
$$

Then explain why the model predicts approach to ambient temperature rather than reaching it in finite time.

### STEP-style Question 5

A deterministic financial index satisfies

$$
\dfrac{dS}{dt}=rS.
$$

Suppose $S(1)=18000$ and $S(3)=22000$.

Find $r$ and $S(0)$ exactly in logarithmic form.

## Reflection questions

1. What does a differential equation describe?
2. What does it mean to separate the variables?
3. Why do we usually need an initial condition?
4. Why can the same general solution represent infinitely many curves?
5. Why does $\dfrac{dP}{dt}=kP$ lead to exponential growth?
6. Why does $\dfrac{dT}{dt}=-k(T-A)$ approach an ambient value?
7. Why is a logistic model more realistic than unlimited exponential growth in some contexts?
8. Why is stochastic finance beyond the technique of this chapter?
9. What can R show clearly about solution families that algebra alone may hide?
10. What can exact separation of variables prove that a graph can only suggest?

## Answers

### Answers to Exercises 10.1

#### Bronze

1. $\dfrac{dy}{dt}=ky$, where $k>0$.
2. $\dfrac{dy}{dt}=-ky$, where $k>0$.
3. $\dfrac{dP}{dt}=-0.05P$.
4. $\dfrac{dv}{dt}=-kv^2$, where $k>0$.
5. $\dfrac{dW}{dt}=k(M-W)$, where $k>0$.

#### Silver

1. $\dfrac{dT}{dt}=-k(T-22)$, where $k>0$.
2. $\dfrac{dI}{dt}=kI(1000-I)$, where $k>0$.
3. $\dfrac{dS}{dt}=rS$.
4. $\dfrac{dD}{dt}=-kD$, where $k>0$.
5. $N(0)=200$.

#### Gold

1. $\dfrac{dT}{dt}=k(30-T)$, where $k>0$.
2. $\dfrac{dy}{dx}=kxy$, where $k>0$.
3. $\dfrac{dy}{dx}=-\dfrac{ky}{1+x}$, where $k>0$.
4. For example, $\dfrac{dP}{dt}=kP\left(1-\dfrac{P}{5000}\right)$.
5. If $T=A$, then $T-A=0$, so $\dfrac{dT}{dt}=0$. Therefore $T=A$ is an equilibrium solution.

### Answers to Exercises 10.2

#### Bronze

1. $y=x^3+C$.
2. $y=Ae^{x^2}$.
3. $y=Ae^x$.
4. $y^2=x^2+C$.
5. $y=\dfrac{1}{C-2x^2}$.

#### Silver

1. $y=2e^{\dfrac{x^2}{2}+x}$.
2. $y^2=3x^2+16$.
3. $y=5(1+x)$, on the interval $x>-1$.
4. $y=e^{x^2}-1$.
5. $y=-1+\sqrt{6x^2+9}$.

#### Gold

1. $y=\dfrac{1}{1-\dfrac{x^2}{2}}$.
2. $y=3e^{\sin x}$.
3. $y=x+1$ on the interval $x>-1$.
4. $y=\dfrac{2}{3-2e^x}$.
5. $y=4+e^{x^2+x}$.

### Answers to Exercises 10.3

#### Bronze

1. $A=5$.
2. $A=4$.
3. $C=9$.
4. $y=2e^x$.
5. $A=80$.

#### Silver

1. $y=3e^{4x}$.
2. $y=2e^{\dfrac{x^2}{2}}$.
3. $y=5(1+x)^2$, on the interval $x>-1$.
4. $y=\sqrt{4x^2+1}$ near $x=0$.
5. $y=2e^{\dfrac{x^2}{2}}-2$.

#### Gold

1. $y=1+e^{x^3}$.
2. $y=\sqrt{2x^2+2x+9}$ near $x=0$.
3. $y=4\sec x$ on an interval containing $0$.
4. $y=3+\dfrac{1}{1-\dfrac{x^2}{2}}$.
5. $k=\dfrac{1}{5}\ln2$, so $y=100e^{\dfrac{x}{5}\ln2}$.

### Answers to Exercises 10.4

#### Bronze

1. $P=50e^{0.2t}$.
2. $P=80e^{-0.3t}$.
3. $\dfrac{dP}{dt}=0.05P$.
4. The half-life is $\dfrac{\ln2}{0.1}=10\ln2$.
5. $k>0$ means growth; $k<0$ means decay.

#### Silver

1. $k=\dfrac{1}{10}\ln\left(\dfrac{7}{5}\right)$.
2. $k=\dfrac{1}{4}\ln\left(\dfrac{4}{3}\right)$.
3. $N(5)=200e^2$.
4. $t=4\ln6$.
5. $k=\dfrac{\ln2}{8}$.

#### Gold

1. $k=\dfrac{\ln3}{12}$.
2. $k=\dfrac{\ln2}{6}$.
3. $P=\dfrac{1000}{1+9e^{-0.4t}}$.
4. $t=\dfrac{\ln36}{0.4}=2.5\ln36\approx8.96$.
5. If $P=0$ or $P=M$, then $\dfrac{dP}{dt}=0$, so both are equilibrium solutions.

### Answers to Exercises 10.5

#### Bronze

1. $T=20+60e^{-0.2t}$.
2. $T=25+45e^{-kt}$.
3. $T(0)=80$.
4. The long-term temperature is $20^\circ\text{C}$.
5. If $T=A$, then $T-A=0$, so the derivative is $0$.

#### Silver

1. $k=\dfrac{1}{5}\ln\left(\dfrac{7}{4}\right)$.
2. $t=\dfrac{\ln7}{k}=\dfrac{5\ln7}{\ln(7/4)}$.
3. $T=18+10e^{-0.1t}$.
4. $T=30-20e^{-0.2t}$.
5. $W=100-30e^{-0.3t}$.

#### Gold

1. $T=26+54e^{-\dfrac{t}{10}\ln(\dfrac{9}{4})}$.
2. $t=\dfrac{10\ln(27/2)}{\ln(9/4)}\approx32.1$ minutes.
3. $T=32-20e^{-kt}$, where $k=\dfrac{1}{5}\ln2$.
4. Differentiating gives $\dfrac{dT}{dt}=-k(T_0-A)e^{-kt}=-k(T-A)$.
5. If $C\neq0$, then $e^{-kt}$ is positive for every finite $t$, so $T-A=Ce^{-kt}$ is never $0$.

### Answers to Exercises 10.6

#### Bronze

1. $S=1000e^{0.04t}$.
2. $S=5000e^{-0.02t}$.
3. $S(1)=15000e^{0.02}\approx15303.02$.
4. It decreases continuously at $10\%$ per unit time.
5. It has no random term, so each value of $t$ gives exactly one predicted value.

#### Silver

1. The value is $2000e^{0.15}\approx2323.67$.
2. The value is $8000e^{-0.08}\approx7384.93$.
3. $r=0.02$.
4. $t=\dfrac{\ln(3/2)}{0.06}\approx6.76$.
5. $t=\dfrac{\ln(5/3)}{0.08}\approx6.39$.

#### Gold

1. Separating variables gives $\ln S=rt+C$, so $S=Ae^{rt}$. Since $S(0)=S_0$, $A=S_0$.
2. $r=\dfrac{\ln2}{3}$.
3. $k=\dfrac{1}{4}\ln\left(\dfrac{5}{4}\right)$.
4. Daily exchange rates have shocks, interventions, liquidity effects and news effects, so a smooth deterministic exponential curve is only a simplified backbone.
5. $r=\dfrac{1}{2}\ln\left(\dfrac{11}{9}\right)$ and $S(0)=18000e^{-r}$.

### Answers to STEP-style extension questions

1. Separating variables gives
   $$
   \int \dfrac{1}{y}\,dy=\int x\,dx.
   $$
   Hence
   $$
   \ln|y|=\dfrac{x^2}{2}+C.
   $$
   Since $y(0)=2$,
   $$
   y=2e^{x^2/2}.
   $$
   Differentiating gives
   $$
   \dfrac{dy}{dx}=2e^{x^2/2}\cdot x=xy.
   $$

2. Separate variables:
   $$
   \int \dfrac{1}{(y-1)^2}\,dy=\int x\,dx.
   $$
   Thus
   $$
   -\dfrac{1}{y-1}=\dfrac{x^2}{2}+C.
   $$
   Since $y(0)=2$, we get $C=-1$. Hence
   $$
   \dfrac{1}{y-1}=1-\dfrac{x^2}{2}.
   $$
   Therefore,
   $$
   y=1+\dfrac{1}{1-\dfrac{x^2}{2}}.
   $$
   The largest interval containing $0$ is
   $$
   -\sqrt2<x<\sqrt2.
   $$

3. Separate variables:
   $$
   \dfrac{dP}{P(1-P/M)}=k\,dt.
   $$
   Since
   $$
   \dfrac{1}{P(1-P/M)}=\dfrac{M}{P(M-P)}
   =
   \dfrac{1}{P}+\dfrac{1}{M-P},
   $$
   integration gives
   $$
   \ln|P|-\ln|M-P|=kt+C.
   $$
   Hence
   $$
   \ln\left|\dfrac{P}{M-P}\right|=kt+C.
   $$
   Therefore
   $$
   \dfrac{P}{M-P}=Be^{kt}.
   $$
   Rearranging gives
   $$
   P=\dfrac{M}{1+Ae^{-kt}}.
   $$

4. Separating variables gives
   $$
   \int \dfrac{1}{T-A}\,dT=\int -k\,dt.
   $$
   Hence
   $$
   \ln|T-A|=-kt+C.
   $$
   Therefore
   $$
   T=A+Ce^{-kt}.
   $$
   Using $T(0)=T_0$ gives $C=T_0-A$, so
   $$
   T=A+(T_0-A)e^{-kt}.
   $$
   Since $e^{-kt}>0$ for finite $t$, the model approaches $A$ but does not reach it in finite time unless $T_0=A$.

5. Since
   $$
   S=S_0e^{rt},
   $$
   the two conditions give
   $$
   18000=S_0e^r
   $$
   and
   $$
   22000=S_0e^{3r}.
   $$
   Dividing gives
   $$
   \dfrac{11}{9}=e^{2r}.
   $$
   Hence
   $$
   r=\dfrac{1}{2}\ln\left(\dfrac{11}{9}\right).
   $$
   Also
   $$
   S_0=18000e^{-r}
   =
   18000\sqrt{\dfrac{9}{11}}.
   $$