# Complex numbers

```{r}
#| label: setup-complex-numbers
#| include: false

library(ggplot2)

book_theme <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = base_size - 1),
      panel.grid.minor = element_blank()
    )
}

axis_layers <- function() {
  list(
    geom_hline(yintercept = 0),
    geom_vline(xintercept = 0)
  )
}

argand_points <- function(z_values, labels) {
  data.frame(
    x = Re(z_values),
    y = Im(z_values),
    label = labels
  )
}

circle_locus <- function(centre, radius, n = 500) {
  theta <- seq(0, 2 * pi, length.out = n)
  data.frame(
    x = Re(centre) + radius * cos(theta),
    y = Im(centre) + radius * sin(theta)
  )
}

ray_locus <- function(start, angle, length_value = 7, n = 300) {
  t_values <- seq(0, length_value, length.out = n)
  data.frame(
    x = Re(start) + t_values * cos(angle),
    y = Im(start) + t_values * sin(angle),
    t = t_values
  )
}

region_grid <- function(x_min, x_max, y_min, y_max, n = 300) {
  expand.grid(
    x = seq(x_min, x_max, length.out = n),
    y = seq(y_min, y_max, length.out = n)
  )
}
```

## Overview

This chapter develops six connected ideas:

1. imaginary numbers and Cartesian form;
2. arithmetic, conjugates, modulus and argument;
3. polar form and geometric transformations;
4. equations with complex roots;
5. loci and shaded regions in the complex plane;
6. mixed problem solving with exact reasoning.

The central aim is to make complex numbers both algebraic and geometric. A complex number

$$
z=x+iy
$$

is also the point

$$
(x,y)
$$

on an Argand diagram. This link between algebra and geometry is especially important when solving loci questions.

::: {.callout-note}
The exercises are organised into tiers.

**Bronze** develops fluency.  
**Silver** develops application.  
**Gold** develops exam-style reasoning.  
The **STEP-style** section extends the chapter beyond routine A Level fluency.
:::

## 12.1 Imaginary numbers and Cartesian form

### Key mathematical ideas

The imaginary unit $i$ is defined by

$$
i^2=-1.
$$

A complex number has Cartesian form

$$
z=x+iy,
$$

where $x$ and $y$ are real numbers.

The real part is

$$
\operatorname{Re}(z)=x,
$$

and the imaginary part is

$$
\operatorname{Im}(z)=y.
$$

Two complex numbers are equal if and only if their real and imaginary parts are equal. Therefore, if

$$
a+bi=c+di,
$$

then

$$
a=c
\quad\text{and}\quad
b=d.
$$

### Visualisation using R

```{r}
#| label: fig-12-basic-argand-points
#| fig-cap: "Complex numbers represented as points on an Argand diagram."
#| fig-width: 6
#| fig-height: 5

points_df <- argand_points(
  z_values = c(3 + 2i, -2 + 4i, -3 - 1i, 2 - 3i),
  labels = c("3 + 2i", "-2 + 4i", "-3 - i", "2 - 3i")
)

ggplot(points_df, aes(x = x, y = y)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal(xlim = c(-5, 5), ylim = c(-5, 5)) +
  labs(
    title = "The complex plane",
    subtitle = "The real part is horizontal; the imaginary part is vertical.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Worked example

Given

$$
3+2i+a+bi=7-i,
$$

find $a$ and $b$.

The left-hand side is

$$
(3+a)+(2+b)i.
$$

Comparing real and imaginary parts gives

$$
3+a=7
$$

and

$$
2+b=-1.
$$

Therefore,

$$
a=4
\quad\text{and}\quad
b=-3.
$$

### Short investigation

Use R to separate real and imaginary parts.

```{r}
#| label: tbl-12-real-imaginary-parts
#| echo: true

z_values <- c(3 + 2i, -2 + 4i, -3 - 1i, 2 - 3i)

data.frame(
  z = as.character(z_values),
  real_part = Re(z_values),
  imaginary_part = Im(z_values)
)
```

### Exercises 12.1

#### Bronze

1. Simplify $i^2+i^4$.
2. Write down the real and imaginary parts of $5-3i$.
3. If $a+2i=7+bi$, find $a$ and $b$.
4. Simplify $(2+3i)+(4-i)$.
5. Simplify $(7-2i)-(3+5i)$.

#### Silver

1. If $(a+1)+(b-2)i=6+3i$, find $a$ and $b$.
2. Simplify $i^{17}$.
3. Simplify $(3+2i)(1-4i)$.
4. Find real numbers $a$ and $b$ such that $(2+i)+(a+bi)=5-4i$.
5. Show that $(1+i)^2=2i$.

#### Gold

1. Find positive real numbers $a$ and $b$ if $(a+bi)^2=7+24i$.
2. Find all real $x$ such that $(x+2i)^2$ is real.
3. Find all real $y$ such that $(3+yi)^2$ is purely imaginary.
4. If $z=x+iy$ and $z+\bar z=10$, find $x$.
5. If $z=x+iy$ and $z-\bar z=8i$, find $y$.

## 12.2 Conjugates, modulus, argument and division

### Key mathematical ideas

If

$$
z=x+iy,
$$

then the conjugate of $z$ is

$$
\bar z=x-iy.
$$

The modulus of $z$ is

$$
|z|=\sqrt{x^2+y^2}.
$$

The argument of $z$ is the angle measured from the positive real axis to the line from the origin to the point representing $z$.

The key identity is

$$
z\bar z=(x+iy)(x-iy)=x^2+y^2=|z|^2.
$$

This is why conjugates are useful for division.

### Visualisation using R

```{r}
#| label: fig-12-conjugate-reflection
#| fig-cap: "Conjugation reflects a complex number in the real axis."
#| fig-width: 6
#| fig-height: 5

points_df <- argand_points(
  z_values = c(3 + 4i, 3 - 4i),
  labels = c("z = 3 + 4i", "conjugate = 3 - 4i")
)

segments_df <- data.frame(
  x = c(0, 0),
  y = c(0, 0),
  xend = points_df$x,
  yend = points_df$y,
  label = points_df$label
)

ggplot() +
  geom_segment(
    data = segments_df,
    aes(x = x, y = y, xend = xend, yend = yend, linetype = label),
    linewidth = 1.1
  ) +
  geom_point(data = points_df, aes(x = x, y = y), size = 3) +
  geom_text(data = points_df, aes(x = x, y = y, label = label), nudge_x = 0.25) +
  axis_layers() +
  coord_equal(xlim = c(-1, 5), ylim = c(-5, 5)) +
  labs(
    title = "Complex conjugates",
    subtitle = "The real part is unchanged; the imaginary part changes sign.",
    x = "Re(z)",
    y = "Im(z)",
    linetype = "Point"
  ) +
  book_theme()
```

### Worked example

Calculate

$$
\dfrac{5+2i}{1-3i}.
$$

Multiply numerator and denominator by the conjugate of the denominator:

$$
\dfrac{5+2i}{1-3i}
=
\dfrac{(5+2i)(1+3i)}{(1-3i)(1+3i)}.
$$

The numerator is

$$
(5+2i)(1+3i)=5+15i+2i+6i^2=-1+17i.
$$

The denominator is

$$
1^2+3^2=10.
$$

Therefore,

$$
\dfrac{5+2i}{1-3i}
=
\dfrac{-1+17i}{10}
=
-\dfrac{1}{10}+\dfrac{17}{10}i.
$$

### Short investigation

R can check moduli and principal arguments.

```{r}
#| label: tbl-12-modulus-argument
#| echo: true

z_values <- c(3 + 4i, -3 + 4i, -3 - 4i, 3 - 4i)

data.frame(
  z = as.character(z_values),
  modulus = Mod(z_values),
  argument_radians = round(Arg(z_values), 4),
  argument_degrees = round(Arg(z_values) * 180 / pi, 1)
)
```

### Exercises 12.2

#### Bronze

1. Find the conjugate of $4-7i$.
2. Find $|3+4i|$.
3. Find $|1-i|$.
4. Find $(2+3i)(2-3i)$.
5. Find the argument of $1+i$.

#### Silver

1. Find $\dfrac{3+4i}{1+i}$.
2. Find $\dfrac{5-2i}{2+i}$.
3. Find $|2-5i|$.
4. Find the principal argument of $-1+\sqrt3 i$.
5. Show that $\dfrac{1}{3-4i}=\dfrac{3}{25}+\dfrac{4}{25}i$.

#### Gold

1. Find $z$ if $z+\bar z=6$ and $z\bar z=25$, with $\operatorname{Im}(z)>0$.
2. Find $z$ if $z-\bar z=10i$ and $|z|=13$, with $\operatorname{Re}(z)>0$.
3. Find all complex numbers $z=x+iy$ such that $z^2$ is real.
4. Find all complex numbers $z=x+iy$ such that $z^2$ is purely imaginary.
5. Find $\arg\left(\dfrac{1+i}{\sqrt3+i}\right)$ in exact form.

## 12.3 Polar form, multiplication and transformations

### Key mathematical ideas

A complex number can be written in Cartesian form as

$$
z=x+iy.
$$

It can also be written in polar form as

$$
z=r(\cos\theta+i\sin\theta),
$$

where

$$
r=|z|
$$

and

$$
\theta=\arg z.
$$

If

$$
z_1=r_1(\cos\theta_1+i\sin\theta_1)
$$

and

$$
z_2=r_2(\cos\theta_2+i\sin\theta_2),
$$

then

$$
z_1z_2
=
r_1r_2\left(\cos(\theta_1+\theta_2)+i\sin(\theta_1+\theta_2)\right).
$$

Multiplication multiplies moduli and adds arguments. Division divides moduli and subtracts arguments:

$$
\dfrac{z_1}{z_2}
=
\dfrac{r_1}{r_2}\left(\cos(\theta_1-\theta_2)+i\sin(\theta_1-\theta_2)\right).
$$

### Visualisation using R

```{r}
#| label: fig-12-complex-multiplication
#| fig-cap: "Multiplication by 1 + i rotates by pi/4 and scales by sqrt(2)."
#| fig-width: 6
#| fig-height: 5

original <- c(1 + 0i, 0 + 1i, -1 + 0i, 0 - 1i)
multiplier <- 1 + 1i
transformed <- multiplier * original

points_df <- rbind(
  data.frame(x = Re(original), y = Im(original), set = "original"),
  data.frame(x = Re(transformed), y = Im(transformed), set = "after multiplying by 1 + i")
)

segments_df <- data.frame(
  x = Re(original),
  y = Im(original),
  xend = Re(transformed),
  yend = Im(transformed)
)

ggplot() +
  geom_segment(data = segments_df, aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.55) +
  geom_point(data = points_df, aes(x = x, y = y, shape = set), size = 3) +
  axis_layers() +
  coord_equal(xlim = c(-2, 2), ylim = c(-2, 2)) +
  labs(
    title = "Complex multiplication as a transformation",
    subtitle = "Multiplication by 1 + i scales by sqrt(2) and rotates by pi/4.",
    x = "Re(z)",
    y = "Im(z)",
    shape = "Point set"
  ) +
  book_theme()
```

### Worked example

Write

$$
z=-1+\sqrt3 i
$$

in polar form.

First,

$$
r=\sqrt{(-1)^2+(\sqrt3)^2}=2.
$$

The point lies in Quadrant II, with reference angle

$$
\dfrac{\pi}{3}.
$$

Therefore,

$$
\arg z=\dfrac{2\pi}{3}.
$$

Hence

$$
z=2\left(\cos\dfrac{2\pi}{3}+i\sin\dfrac{2\pi}{3}\right).
$$

### Worked example

Let

$$
z=\sqrt3+i
$$

and

$$
w=1+i.
$$

Find $zw$ in polar form.

We have

$$
z=2\left(\cos\dfrac{\pi}{6}+i\sin\dfrac{\pi}{6}\right)
$$

and

$$
w=\sqrt2\left(\cos\dfrac{\pi}{4}+i\sin\dfrac{\pi}{4}\right).
$$

Therefore,

$$
zw=2\sqrt2\left(\cos\dfrac{5\pi}{12}+i\sin\dfrac{5\pi}{12}\right).
$$

### Exercises 12.3

#### Bronze

1. Write $1+i$ in polar form.
2. Write $\sqrt3+i$ in polar form.
3. Write $-1+i$ in polar form.
4. Convert $2\left(\cos\dfrac{\pi}{3}+i\sin\dfrac{\pi}{3}\right)$ into Cartesian form.
5. Describe the effect of multiplying by $i$.

#### Silver

1. Write $-1+\sqrt3i$ in polar form.
2. Write $-\sqrt3-i$ in polar form using the principal argument.
3. Find the polar form of $(1+i)(\sqrt3+i)$.
4. Find the polar form of $\dfrac{\sqrt3+i}{1+i}$.
5. Describe the transformation caused by multiplying by $2i$.

#### Gold

1. Find the polar form of $-3+3\sqrt3i$.
2. Find the polar form of $2-2\sqrt3i$ using the principal argument.
3. If $z=4(\cos\theta+i\sin\theta)$ and $\operatorname{Re}(z)=2$, find possible values of $\theta$ in $(-\pi,\pi]$.
4. If multiplying by $a+bi$ rotates every point by $\dfrac{\pi}{3}$ and scales every distance by $2$, find $a+bi$.
5. Show that $|zw|=|z||w|$ when $z$ and $w$ are written in polar form.

## 12.4 Equations with complex roots

### Key mathematical ideas

Complex numbers allow equations such as

$$
x^2+1=0
$$

to have solutions:

$$
x=\pm i.
$$

For polynomials with real coefficients, non-real roots occur in conjugate pairs. Therefore, if

$$
2+3i
$$

is a root of a polynomial with real coefficients, then

$$
2-3i
$$

is also a root.

If a quadratic has roots

$$
2+3i
\quad\text{and}\quad
2-3i,
$$

then the quadratic is

$$
(x-(2+3i))(x-(2-3i)).
$$

This simplifies to

$$
(x-2)^2+9=x^2-4x+13.
$$

### Visualisation using R

```{r}
#| label: fig-12-complex-conjugate-roots
#| fig-cap: "Complex conjugate roots are reflected in the real axis."
#| fig-width: 6
#| fig-height: 5

root_points <- argand_points(
  z_values = c(2 + 3i, 2 - 3i),
  labels = c("2 + 3i", "2 - 3i")
)

ggplot(root_points, aes(x = x, y = y)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), nudge_x = 0.25) +
  axis_layers() +
  coord_equal(xlim = c(-1, 5), ylim = c(-4, 4)) +
  labs(
    title = "Conjugate pair of roots",
    subtitle = "For real coefficients, non-real roots occur in conjugate pairs.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Worked example

Find the two square roots of

$$
3+4i.
$$

Let

$$
(a+bi)^2=3+4i.
$$

Expanding gives

$$
a^2-b^2+2abi=3+4i.
$$

Therefore,

$$
a^2-b^2=3
$$

and

$$
2ab=4.
$$

Also, taking moduli gives

$$
a^2+b^2=|3+4i|=5.
$$

Adding

$$
a^2-b^2=3
$$

and

$$
a^2+b^2=5
$$

gives

$$
2a^2=8,
$$

so

$$
a=\pm2.
$$

Since $ab=2$, the roots are

$$
2+i
\quad\text{and}\quad
-2-i.
$$

Therefore,

$$
z=\pm(2+i).
$$

### Exercises 12.4

#### Bronze

1. Solve $x^2+9=0$.
2. Solve $x^2-4x+13=0$.
3. If $3+2i$ is a root of a real polynomial, write down another root.
4. Form the quadratic with roots $1\pm2i$.
5. Find the sum and product of the roots $4\pm i$.

#### Silver

1. Solve $x^2+2x+10=0$.
2. Form the quadratic with roots $-3\pm4i$.
3. Find the square roots of $3+4i$.
4. Find the square roots of $5-12i$.
5. A cubic with real coefficients has roots $1$ and $2+i$. Find the cubic with leading coefficient $1$.

#### Gold

1. Find the square roots of $-7+24i$.
2. A quartic with real coefficients has roots $1+i$ and $2+3i$. Form a quartic with leading coefficient $1$.
3. Find $a$ and $b$ if $x^2+ax+b=0$ has root $3-2i$.
4. If $z^2=8i$, find the roots in polar and Cartesian form.
5. Show that if a real quadratic has non-real roots, then its discriminant is negative.

## 12.5 Loci and shaded regions in the complex plane

### Key mathematical ideas

Let

$$
z=x+iy.
$$

Then complex equations and inequalities can describe loci in the plane.

A circle locus has the form

$$
|z-(a+bi)|=r.
$$

This means the point $z$ is distance $r$ from the fixed point $a+bi$. In Cartesian form,

$$
(x-a)^2+(y-b)^2=r^2.
$$

A perpendicular-bisector locus has the form

$$
|z-a|=|z-b|.
$$

This means the point $z$ is equidistant from two fixed points.

An argument locus has the form

$$
\arg(z-a)=\theta.
$$

This is a ray starting from the point $a$, at angle $\theta$ to the positive real axis. The starting point is excluded because the argument is undefined there.

### Visualisation using R: shaded circle region

```{r}
#| label: fig-12-locus-disk
#| fig-cap: "The shaded region |z - (2 + i)| <= 3."
#| fig-width: 6
#| fig-height: 5

region_df <- region_grid(-2, 6, -3, 5)
region_df$inside <- (region_df$x - 2)^2 + (region_df$y - 1)^2 <= 9

circle_df <- circle_locus(2 + 1i, 3)
centre_df <- argand_points(2 + 1i, "2 + i")

ggplot() +
  geom_raster(data = subset(region_df, inside), aes(x = x, y = y), alpha = 0.25) +
  geom_path(data = circle_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = centre_df, aes(x = x, y = y), size = 3) +
  geom_text(data = centre_df, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal(xlim = c(-2, 6), ylim = c(-3, 5)) +
  labs(
    title = "A modulus inequality gives a disk",
    subtitle = "|z - (2 + i)| <= 3 means within 3 units of 2 + i.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Visualisation using R: shaded perpendicular-bisector region

```{r}
#| label: fig-12-locus-half-plane
#| fig-cap: "The shaded region |z - 1| <= |z - (5 + 2i)|."
#| fig-width: 6
#| fig-height: 5

region_df <- region_grid(-1, 7, -3, 6)
region_df$inside <- (region_df$x - 1)^2 + region_df$y^2 <=
  (region_df$x - 5)^2 + (region_df$y - 2)^2

line_df <- data.frame(x = seq(-1, 7, length.out = 400))
line_df$y <- 7 - 2 * line_df$x

fixed_points <- argand_points(
  z_values = c(1 + 0i, 5 + 2i),
  labels = c("1", "5 + 2i")
)

ggplot() +
  geom_raster(data = subset(region_df, inside), aes(x = x, y = y), alpha = 0.25) +
  geom_line(data = line_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = fixed_points, aes(x = x, y = y), size = 3) +
  geom_text(data = fixed_points, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal(xlim = c(-1, 7), ylim = c(-3, 6)) +
  labs(
    title = "A distance comparison gives a half-plane",
    subtitle = "The boundary is the perpendicular bisector of 1 and 5 + 2i.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Visualisation using R: shaded argument region

```{r}
#| label: fig-12-locus-argument-sector
#| fig-cap: "The shaded region 0 <= arg(z - (1 + i)) <= pi/4."
#| fig-width: 6
#| fig-height: 5

region_df <- region_grid(-1, 7, -1, 6)
region_df$dx <- region_df$x - 1
region_df$dy <- region_df$y - 1
region_df$inside <- region_df$dx > 0 & region_df$dy >= 0 & region_df$dy <= region_df$dx

ray_1 <- ray_locus(1 + 1i, 0, length_value = 7)
ray_2 <- ray_locus(1 + 1i, pi / 4, length_value = 7)
start_df <- argand_points(1 + 1i, "1 + i")

ggplot() +
  geom_raster(data = subset(region_df, inside), aes(x = x, y = y), alpha = 0.25) +
  geom_path(data = ray_1, aes(x = x, y = y), linewidth = 1.1) +
  geom_path(data = ray_2, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = start_df, aes(x = x, y = y), size = 3, shape = 1) +
  geom_text(data = start_df, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal(xlim = c(-1, 7), ylim = c(-1, 6)) +
  labs(
    title = "An argument inequality gives an angular region",
    subtitle = "The starting point is excluded because the argument is undefined.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Worked example

Find the Cartesian equation of

$$
|z-(2+i)|=3.
$$

Let

$$
z=x+iy.
$$

Then

$$
z-(2+i)=(x-2)+i(y-1).
$$

So

$$
|z-(2+i)|=\sqrt{(x-2)^2+(y-1)^2}.
$$

Therefore,

$$
\sqrt{(x-2)^2+(y-1)^2}=3.
$$

Squaring gives

$$
(x-2)^2+(y-1)^2=9.
$$

### Exercises 12.5

#### Bronze

1. Write the Cartesian equation of $|z-(2+i)|=3$.
2. Sketch and describe $|z|=4$.
3. Sketch and describe $\operatorname{Im}(z)=2$.
4. Sketch and describe $\operatorname{Re}(z)=-1$.
5. Describe the locus $\arg z=\dfrac{\pi}{4}$.

#### Silver

1. Write the Cartesian equation of $|z-(3-2i)|=5$.
2. Find the Cartesian equation of $|z-1|=|z-(5+2i)|$.
3. Shade the region $|z-(2+i)|\leq3$.
4. Shade the region $|z-1|\leq|z-(5+2i)|$.
5. Describe the region $0\leq\arg(z-(1+i))\leq\dfrac{\pi}{4}$.

#### Gold

1. Find the intersection points of $|z-(3+2i)|=2$ and $\arg(z-1)=\dfrac{\pi}{4}$.
2. Find the least and greatest possible values of $|z|$ if $|z-(8+6i)|=5$.
3. Find the greatest and least possible values of $\arg z$ if $|z-(5+5i)|=5$.
4. Explain geometrically why $|z-a|=|z-b|$ is a perpendicular bisector.
5. Find the Cartesian inequality represented by $|z-1|<|z-(5+2i)|$.

## 12.6 Mixed problem solving

### Worked example: intersection of loci

Find all complex numbers satisfying

$$
|z-(3+2i)|=2
$$

and

$$
\arg(z-1)=\dfrac{\pi}{4}.
$$

Let

$$
z=x+iy.
$$

The circle is

$$
(x-3)^2+(y-2)^2=4.
$$

The argument condition means the ray from $1$ at angle $\dfrac{\pi}{4}$, so

$$
y=x-1,
\quad x>1.
$$

Substitute into the circle:

$$
(x-3)^2+(x-3)^2=4.
$$

Thus

$$
2(x-3)^2=4,
$$

so

$$
x=3\pm\sqrt2.
$$

Since both values satisfy $x>1$,

$$
y=2\pm\sqrt2.
$$

Therefore,

$$
z=(3+\sqrt2)+(2+\sqrt2)i
$$

or

$$
z=(3-\sqrt2)+(2-\sqrt2)i.
$$

### Visualisation using R

```{r}
#| label: fig-12-locus-intersection
#| fig-cap: "Intersections of |z - (3 + 2i)| = 2 and arg(z - 1) = pi/4."
#| fig-width: 6
#| fig-height: 5

circle_df <- circle_locus(3 + 2i, 2)
ray_df <- ray_locus(1 + 0i, pi / 4, length_value = 7)
intersection_df <- data.frame(
  x = c(3 - sqrt(2), 3 + sqrt(2)),
  y = c(2 - sqrt(2), 2 + sqrt(2)),
  label = c("lower point", "upper point")
)

ggplot() +
  geom_path(data = circle_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_path(data = ray_df, aes(x = x, y = y), linewidth = 1.1) +
  geom_point(data = intersection_df, aes(x = x, y = y), size = 3) +
  geom_text(data = intersection_df, aes(x = x, y = y, label = label), nudge_y = 0.35) +
  axis_layers() +
  coord_equal(xlim = c(-1, 7), ylim = c(-1, 6)) +
  labs(
    title = "Intersecting complex loci",
    subtitle = "The algebra confirms the points suggested by the Argand diagram.",
    x = "Re(z)",
    y = "Im(z)"
  ) +
  book_theme()
```

### Mixed exercises

1. Find all roots of $z^2=3+4i$.
2. Find the quadratic with real coefficients whose roots are $4\pm3i$.
3. Write $-2+2\sqrt3i$ in polar form.
4. Describe the transformation caused by multiplication by $-2i$.
5. Find the Cartesian equation of $|z+2-i|=4$.
6. Find the Cartesian inequality represented by $|z-(1+i)|\leq2$.
7. Find the Cartesian equation of $|z-2|=|z-(4+4i)|$.
8. Find the exact values of $z$ satisfying $|z-(3+2i)|=2$ and $\arg(z-1)=\dfrac{\pi}{4}$.
9. Find the least and greatest possible values of $|z|$ if $|z-(8+6i)|=5$.
10. Find the greatest and least possible values of $\arg z$ if $|z-(5+5i)|=5$.

## STEP-style extension questions

### STEP-style Question 1

Let

$$
z=x+iy.
$$

Prove that

$$
z\bar z=|z|^2.
$$

Use this result to derive the division formula for

$$
\dfrac{a+bi}{c+di}.
$$

### STEP-style Question 2

Find all complex numbers satisfying

$$
z^2=3+4i.
$$

Represent the roots on an Argand diagram and explain why the roots are opposite points.

### STEP-style Question 3

Let

$$
z=r(\cos\theta+i\sin\theta).
$$

Show that multiplication by $z$ maps any complex number $w$ to a point obtained by rotating $w$ through angle $\theta$ and scaling its distance from the origin by $r$.

### STEP-style Question 4

The complex number $z$ satisfies

$$
|z-(3+2i)|=2
$$

and

$$
\arg(z-1)=\dfrac{\pi}{4}.
$$

Find all possible values of $z$ exactly.

### STEP-style Question 5

Let $z=x+iy$. The locus of $z$ is

$$
|z-(5+5i)|=5.
$$

Find the greatest and least possible values of $\arg z$, explaining your reasoning geometrically.

## Reflection questions

1. Why do we need the number $i$?
2. What does it mean to compare real and imaginary parts?
3. Why is the conjugate useful for division?
4. Why is $|z|$ interpreted as a distance?
5. Why must arguments be handled carefully in different quadrants?
6. Why does multiplication by $i$ rotate a point by $\dfrac{\pi}{2}$?
7. Why do non-real roots of real polynomials occur in conjugate pairs?
8. Why does $|z-a|=r$ represent a circle?
9. Why does $|z-a|=|z-b|$ represent a perpendicular bisector?
10. What can an Argand diagram show that algebra alone may hide?

## Answers

### Answers to Exercises 12.1

#### Bronze

1. $i^2+i^4=-1+1=0$.
2. Real part $5$, imaginary part $-3$.
3. $a=7$ and $b=2$.
4. $6+2i$.
5. $4-7i$.

#### Silver

1. $a=5$ and $b=5$.
2. $i^{17}=i$.
3. $(3+2i)(1-4i)=11-10i$.
4. $a=3$ and $b=-5$.
5. $(1+i)^2=1+2i+i^2=2i$.

#### Gold

1. $(4+3i)^2=7+24i$, so $a=4$ and $b=3$.
2. $(x+2i)^2=x^2-4+4xi$, so $x=0$.
3. $(3+yi)^2=9-y^2+6yi$. For this to be purely imaginary, $9-y^2=0$, so $y=\pm3$.
4. $z+\bar z=2x=10$, so $x=5$.
5. $z-\bar z=2iy=8i$, so $y=4$.

### Answers to Exercises 12.2

#### Bronze

1. $4+7i$.
2. $5$.
3. $\sqrt2$.
4. $13$.
5. $\dfrac{\pi}{4}$.

#### Silver

1. $\dfrac{3+4i}{1+i}=\dfrac{7+i}{2}$.
2. $\dfrac{5-2i}{2+i}=\dfrac{8-9i}{5}$.
3. $\sqrt{29}$.
4. $\dfrac{2\pi}{3}$.
5. Multiplying by the conjugate gives $\dfrac{1}{3-4i}=\dfrac{3+4i}{25}$.

#### Gold

1. $z+\bar z=6$ gives $x=3$. Since $z\bar z=|z|^2=25$, we have $9+y^2=25$. Since $\operatorname{Im}(z)>0$, $y=4$, so $z=3+4i$.
2. $z-\bar z=10i$ gives $y=5$. Since $|z|=13$, $x^2+25=169$. Since $\operatorname{Re}(z)>0$, $x=12$, so $z=12+5i$.
3. $z^2=x^2-y^2+2xyi$ is real when $2xy=0$, so $x=0$ or $y=0$.
4. $z^2=x^2-y^2+2xyi$ is purely imaginary when $x^2-y^2=0$, so $y=\pm x$.
5. $\arg(1+i)-\arg(\sqrt3+i)=\dfrac{\pi}{4}-\dfrac{\pi}{6}=\dfrac{\pi}{12}$.

### Answers to Exercises 12.3

#### Bronze

1. $\sqrt2\left(\cos\dfrac{\pi}{4}+i\sin\dfrac{\pi}{4}\right)$.
2. $2\left(\cos\dfrac{\pi}{6}+i\sin\dfrac{\pi}{6}\right)$.
3. $\sqrt2\left(\cos\dfrac{3\pi}{4}+i\sin\dfrac{3\pi}{4}\right)$.
4. $1+\sqrt3 i$.
5. Multiplication by $i$ is rotation by $\dfrac{\pi}{2}$ anticlockwise about the origin.

#### Silver

1. $2\left(\cos\dfrac{2\pi}{3}+i\sin\dfrac{2\pi}{3}\right)$.
2. $2\left(\cos\left(-\dfrac{5\pi}{6}\right)+i\sin\left(-\dfrac{5\pi}{6}\right)\right)$.
3. $2\sqrt2\left(\cos\dfrac{5\pi}{12}+i\sin\dfrac{5\pi}{12}\right)$.
4. $\sqrt2\left(\cos\left(-\dfrac{\pi}{12}\right)+i\sin\left(-\dfrac{\pi}{12}\right)\right)$.
5. Multiplication by $2i$ is scale factor $2$ and rotation by $\dfrac{\pi}{2}$ anticlockwise.

#### Gold

1. $6\left(\cos\dfrac{2\pi}{3}+i\sin\dfrac{2\pi}{3}\right)$.
2. $4\left(\cos\left(-\dfrac{\pi}{3}\right)+i\sin\left(-\dfrac{\pi}{3}\right)\right)$.
3. $4\cos\theta=2$, so $\cos\theta=\dfrac{1}{2}$. Hence $\theta=\pm\dfrac{\pi}{3}$.
4. $a+bi=2\left(\cos\dfrac{\pi}{3}+i\sin\dfrac{\pi}{3}\right)=1+\sqrt3 i$.
5. If $z=r(\cos\alpha+i\sin\alpha)$ and $w=s(\cos\beta+i\sin\beta)$, then $zw=rs(\cos(\alpha+\beta)+i\sin(\alpha+\beta))$, so $|zw|=rs=|z||w|$.

### Answers to Exercises 12.4

#### Bronze

1. $x=\pm3i$.
2. $x=2\pm3i$.
3. $3-2i$.
4. $(x-(1+2i))(x-(1-2i))=(x-1)^2+4=x^2-2x+5$.
5. Sum $8$, product $17$.

#### Silver

1. $x=-1\pm3i$.
2. $(x+3)^2+16=x^2+6x+25$.
3. $\pm(2+i)$.
4. $\pm(3-2i)$.
5. The cubic is $(x-1)((x-2)^2+1)=x^3-5x^2+9x-5$.

#### Gold

1. The square roots of $-7+24i$ are $\pm(3+4i)$.
2. The roots are $1\pm i$ and $2\pm3i$. The quartic is $((x-1)^2+1)((x-2)^2+9)=x^4-6x^3+23x^2-34x+26$.
3. The other root is $3+2i$. The sum is $6$ and the product is $13$. Since the quadratic is $x^2+ax+b=0$, $a=-6$ and $b=13$.
4. $8i=8\left(\cos\dfrac{\pi}{2}+i\sin\dfrac{\pi}{2}\right)$. The square roots have modulus $2\sqrt2$ and arguments $\dfrac{\pi}{4}$ and $\dfrac{5\pi}{4}$, giving $2+2i$ and $-2-2i$.
5. A real quadratic with non-real roots has no real roots, so its graph does not cross the real axis and its discriminant is negative.

### Answers to Exercises 12.5

#### Bronze

1. $(x-2)^2+(y-1)^2=9$.
2. A circle centred at the origin with radius $4$.
3. The horizontal line $y=2$.
4. The vertical line $x=-1$.
5. A ray from the origin at angle $\dfrac{\pi}{4}$, excluding the origin.

#### Silver

1. $(x-3)^2+(y+2)^2=25$.
2. $(x-1)^2+y^2=(x-5)^2+(y-2)^2$, which simplifies to $2x+y=7$.
3. The closed disk centred at $(2,1)$ with radius $3$.
4. The half-plane $2x+y\leq7$.
5. The sector starting at $(1,1)$ between the positive real direction and the line of gradient $1$, excluding the starting point.

#### Gold

1. $z=(3+\sqrt2)+(2+\sqrt2)i$ or $z=(3-\sqrt2)+(2-\sqrt2)i$.
2. The centre $8+6i$ has distance $10$ from the origin and the circle has radius $5$, so the least value is $5$ and the greatest value is $15$.
3. The centre has argument $\dfrac{\pi}{4}$ and distance $5\sqrt2$ from the origin. The tangent angle satisfies $\sin\alpha=\dfrac{5}{5\sqrt2}=\dfrac{1}{\sqrt2}$, so $\alpha=\dfrac{\pi}{4}$. Hence the least argument is $0$ and the greatest argument is $\dfrac{\pi}{2}$.
4. The locus is the set of points equidistant from $a$ and $b$, so it is the perpendicular bisector of the segment joining $a$ and $b$.
5. Expanding gives $(x-1)^2+y^2<(x-5)^2+(y-2)^2$, so $2x+y<7$.

### Answers to Mixed exercises

1. $z=2+i$ or $z=-2-i$.
2. $(x-4)^2+9=x^2-8x+25$.
3. $4\left(\cos\dfrac{2\pi}{3}+i\sin\dfrac{2\pi}{3}\right)$.
4. Multiplication by $-2i$ is scale factor $2$ and rotation by $-\dfrac{\pi}{2}$, or equivalently $\dfrac{3\pi}{2}$.
5. Since $z+2-i=z-(-2+i)$, the equation is $(x+2)^2+(y-1)^2=16$.
6. $(x-1)^2+(y-1)^2\leq4$.
7. $(x-2)^2+y^2=(x-4)^2+(y-4)^2$, which simplifies to $x+2y=7$.
8. $z=(3+\sqrt2)+(2+\sqrt2)i$ or $z=(3-\sqrt2)+(2-\sqrt2)i$.
9. Least value $5$, greatest value $15$.
10. Least argument $0$, greatest argument $\dfrac{\pi}{2}$.

### Answers to STEP-style extension questions

1. If $z=x+iy$, then $\bar z=x-iy$. Hence
   $$
   z\bar z=(x+iy)(x-iy)=x^2+y^2=|z|^2.
   $$
   Therefore,
   $$
   \dfrac{a+bi}{c+di}
   =
   \dfrac{(a+bi)(c-di)}{c^2+d^2}
   =
   \dfrac{ac+bd}{c^2+d^2}+\dfrac{bc-ad}{c^2+d^2}i.
   $$

2. Let $z=a+bi$. Then $a^2-b^2=3$ and $2ab=4$. Also $a^2+b^2=5$. Hence $a^2=4$ and $ab=2$, giving $z=2+i$ or $z=-2-i$. These are opposite points on the Argand diagram because one is the negative of the other.

3. Write $w=s(\cos\phi+i\sin\phi)$. Then
   $$
   zw=rs(\cos(\theta+\phi)+i\sin(\theta+\phi)).
   $$
   So the modulus is multiplied by $r$ and the argument is increased by $\theta$.

4. The ray gives $y=x-1$ with $x>1$. Substitution into $(x-3)^2+(y-2)^2=4$ gives $2(x-3)^2=4$, so $x=3\pm\sqrt2$. Therefore
   $$
   z=(3+\sqrt2)+(2+\sqrt2)i
   $$
   or
   $$
   z=(3-\sqrt2)+(2-\sqrt2)i.
   $$

5. The circle has centre $(5,5)$ and radius $5$. The line from the origin to the centre has argument $\dfrac{\pi}{4}$ and length $5\sqrt2$. The tangent angle satisfies
   $$
   \sin\alpha=\dfrac{5}{5\sqrt2}=\dfrac{1}{\sqrt2},
   $$
   so $\alpha=\dfrac{\pi}{4}$. Hence the least argument is $0$ and the greatest argument is $\dfrac{\pi}{2}$.
